---
title: "Load MSK-CHORD data"
author: "Puck Quarles van Ufford"
date: today
format: html
---

# Set up

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
```

# Extract and list files in MSK-CHORD

```{r}
#| label: list_files
#| warning: false
#| echo: true

untar(tarfile = "../data/_raw/msk_chord_2024.tar.gz",
      exdir = "../data/_raw/")

list.files("../data/_raw/msk_chord_2024/")

readLines("../data/_raw/msk_chord_2024/Readme.txt")
```

Every data set in this folder has a data file and an accompanying meta file.

# Load data

## Clinical patient

```{r}
#| label: load_clinical_patient
#| message: false
#| warning: false
#| echo: true

data_clinical_patient <- read_delim(file = "../data/_raw/msk_chord_2024/data_clinical_patient.txt", 
                                    delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_clinical_patient.txt")

head(data_clinical_patient) 

# Remove 4 rows of column descriptions
data_clinical_patient <- read_delim(file = "../data/_raw/msk_chord_2024/data_clinical_patient.txt", 
                                    delim = "\t", 
                                    skip = 4)

head(data_clinical_patient)

write_tsv(x = data_clinical_patient,
          file = "../data/01_data_clinical_patient.tsv")
```

## Clinical sample

```{r}
#| label: load_clinical_sample
#| message: false
#| warning: false
#| echo: true

data_clinical_sample <- read_delim(file = "../data/_raw/msk_chord_2024/data_clinical_sample.txt", 
                                   delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_clinical_sample.txt")

head(data_clinical_sample) 

# Remove 4 rows of column descriptions
data_clinical_sample <- read_delim(file = "../data/_raw/msk_chord_2024/data_clinical_sample.txt",
                                   delim = "\t", 
                                   skip = 4)

head(data_clinical_sample)

write_tsv(x = data_clinical_sample,
          file = "../data/01_data_clinical_sample.tsv")
```

## CNA hg19

```{r}
#| label: load_cna_hg19
#| message: false
#| warning: false
#| echo: true

data_cna_hg19 <- read_delim(file = "../data/_raw/msk_chord_2024/data_cna_hg19.seg",
                            delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_cna_hg19_seg.txt")

head(data_cna_hg19) 

write_tsv(x = data_cna_hg19,
          file = "../data/01_data_cna_hg19.tsv")
```

## CNA

```{r}
#| label: load_cna
#| message: false
#| warning: false
#| echo: true

data_cna <- read_delim(file = "../data/_raw/msk_chord_2024/data_cna.txt",
                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_cna.txt")

head(data_cna) 

write_tsv(x = data_cna,
          file = "../data/01_data_cna.tsv")
```

## Gene panel matrix

```{r}
#| label: load_gene_panel_matrix
#| message: false
#| warning: false
#| echo: true

data_gene_panel_matrix <- read_delim(file = "../data/_raw/msk_chord_2024/data_gene_panel_matrix.txt",
                                     delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_gene_panel_matrix.txt")

head(data_gene_panel_matrix) 

write_tsv(x = data_gene_panel_matrix,
          file = "../data/01_data_gene_panel_matrix.tsv")
```

## Mutations

```{r}
#| label: load_mutations
#| message: false
#| warning: false
#| echo: true

data_mutations <- read_delim(file = "../data/_raw/msk_chord_2024/data_mutations.txt",
                             delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_mutations.txt")

head(data_mutations) 

write_tsv(x = data_mutations,
          file = "../data/01_data_mutations.tsv")
```

## SV

```{r}
#| label: load_sv
#| message: false
#| warning: false
#| echo: true

data_sv <- read_delim(file = "../data/_raw/msk_chord_2024/data_sv.txt",
                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_sv.txt")

head(data_sv) 

write_tsv(x = data_sv,
          file = "../data/01_data_sv.tsv")
```

## Timeline data

### CA 15-3 labs

```{r}
#| label: load_timeline_ca_15_3_labs
#| message: false
#| warning: false
#| echo: true

data_timeline_ca_15_3_labs <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_ca_15-3_labs.txt",
                                         delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_ca_15-3_labs.txt")

head(data_timeline_ca_15_3_labs) 

write_tsv(x = data_timeline_ca_15_3_labs,
          file = "../data/01_data_timeline_ca_15_3_labs.tsv")
```

### CA 19-9 labs

```{r}
#| label: load_timeline_ca_19_9_labs
#| message: false
#| warning: false
#| echo: true

data_timeline_ca_19_9_labs <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_ca_19-9_labs.txt",
                                         delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_ca_19-9_labs.txt")

head(data_timeline_ca_19_9_labs) 

write_tsv(x = data_timeline_ca_19_9_labs,
          file = "../data/01_data_timeline_ca_19_9_labs.tsv")
```

### Cancer presence

```{r}
#| label: load_timeline_cancer_presence
#| message: false
#| warning: false
#| echo: true

data_timeline_cancer_presence <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_cancer_presence.txt",
                                            delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_cancer_presence.txt")

head(data_timeline_cancer_presence) 

write_tsv(x = data_timeline_cancer_presence,
          file = "../data/01_data_timeline_cancer_presence.tsv")
```

### CEA labs

```{r}
#| label: load_timeline_cea_labs
#| message: false
#| warning: false
#| echo: true

data_timeline_cea_labs <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_cea_labs.txt",
                                     delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_cea_labs.txt")

head(data_timeline_cea_labs) 

write_tsv(x = data_timeline_cea_labs,
          file = "../data/01_data_timeline_cea_labs.tsv")
```

### Diagnosis

```{r}
#| label: load_timeline_diagnosis
#| message: false
#| warning: false
#| echo: true

data_timeline_diagnosis <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_diagnosis.txt",
                                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_diagnosis.txt")

head(data_timeline_diagnosis) 

write_tsv(x = data_timeline_diagnosis,
          file = "../data/01_data_timeline_diagnosis.tsv")
```

### Gleason

```{r}
#| label: load_timeline_gleason
#| message: false
#| warning: false
#| echo: true

data_timeline_gleason <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_gleason.txt",
                                    delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_gleason.txt")

head(data_timeline_gleason) 

write_tsv(x = data_timeline_gleason,
          file = "../data/01_data_timeline_gleason.tsv")
```

### MMR

```{r}
#| label: load_timeline_mmr
#| message: false
#| warning: false
#| echo: true

data_timeline_mmr <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_mmr.txt",
                                delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_mmr.txt")

head(data_timeline_mmr) 

write_tsv(x = data_timeline_mmr,
          file = "../data/01_data_timeline_mmr.tsv")
```

### PD-L1

```{r}
#| label: load_timeline_pdl1
#| message: false
#| warning: false
#| echo: true

data_timeline_pdl1 <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_pdl1.txt",
                                 delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_pdl1.txt")

head(data_timeline_pdl1) 

write_tsv(x = data_timeline_pdl1,
          file = "../data/01_data_timeline_pdl1.tsv")
```

### Performance status

```{r}
#| label: load_timeline_performance_status
#| message: false
#| warning: false
#| echo: true

data_timeline_performance_status <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_performance_status.txt",
                                               delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_performance_status.txt")

head(data_timeline_performance_status) 

write_tsv(x = data_timeline_performance_status,
          file = "../data/01_data_timeline_performance_status.tsv")
```

### Prior meds

```{r}
#| label: load_timeline_prior_meds
#| message: false
#| warning: false
#| echo: true

data_timeline_prior_meds <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_prior_meds.txt",
                                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_prior_meds.txt")

head(data_timeline_prior_meds) 

write_tsv(x = data_timeline_prior_meds,
          file = "../data/01_data_timeline_prior_meds.tsv")
```

### Progression

```{r}
#| label: load_timeline_progression
#| message: false
#| warning: false
#| echo: true

data_timeline_progression <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_progression.txt",
                                        delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_progression.txt")

head(data_timeline_progression) 

write_tsv(x = data_timeline_progression,
          file = "../data/01_data_timeline_progression.tsv")
```

### PSA labs

```{r}
#| label: load_timeline_psa_labs
#| message: false
#| warning: false
#| echo: true

data_timeline_psa_labs <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_psa_labs.txt",
                                     delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_psa_labs.txt")

head(data_timeline_psa_labs) 

write_tsv(x = data_timeline_psa_labs,
          file = "../data/01_data_timeline_psa_labs.tsv")
```

### Radiation

```{r}
#| label: load_timeline_radiation
#| message: false
#| warning: false
#| echo: true

data_timeline_radiation <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_radiation.txt",
                                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_radiation.txt")

head(data_timeline_radiation) 

write_tsv(x = data_timeline_radiation,
          file = "../data/01_data_timeline_radiation.tsv")
```

### Specimen surgery

```{r}
#| label: load_timeline_specimen_surgery
#| message: false
#| warning: false
#| echo: true

data_timeline_specimen_surgery <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_specimen_surgery.txt",
                                             delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_specimen_surgery.txt")

head(data_timeline_specimen_surgery) 

write_tsv(x = data_timeline_specimen_surgery,
          file = "../data/01_data_timeline_specimen_surgery.tsv")
```

### Specimen

```{r}
#| label: load_timeline_specimen
#| message: false
#| warning: false
#| echo: true

data_timeline_specimen <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_specimen.txt",
                                     delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_specimen.txt")

head(data_timeline_specimen) 

write_tsv(x = data_timeline_specimen,
          file = "../data/01_data_timeline_specimen.tsv")
```

### Surgery

```{r}
#| label: load_timeline_surgery
#| message: false
#| warning: false
#| echo: true

data_timeline_surgery <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_surgery.txt",
                                    delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_surgery.txt")

head(data_timeline_surgery) 

write_tsv(x = data_timeline_surgery,
          file = "../data/01_data_timeline_surgery.tsv")
```

### Treatment

```{r}
#| label: load_timeline_treatment
#| message: false
#| warning: false
#| echo: true

data_timeline_treatment <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_treatment.txt",
                                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_treatment.txt")

head(data_timeline_treatment) 

write_tsv(x = data_timeline_treatment,
          file = "../data/01_data_timeline_treatment.tsv")
```

### Tumor sites

```{r}
#| label: load_timeline_tumor_sites
#| message: false
#| warning: false
#| echo: true

data_timeline_tumor_sites <- read_delim(file = "../data/_raw/msk_chord_2024/data_timeline_tumor_sites.txt",
                                        delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_tumor_sites.txt")

head(data_timeline_tumor_sites) 

write_tsv(x = data_timeline_tumor_sites,
          file = "../data/01_data_timeline_tumor_sites.tsv")
```
