---
title: "Prepare data"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
library("tidyverse")
library("patchwork")
```

# Extract data

Extract and list the files in the dataset.

```{r}
untar(tarfile = "../data/_raw/msk_chord_2024.tar.gz",
      exdir = "../data/_raw/")

list.files("../data/_raw/msk_chord_2024/")

readLines("../data/_raw/msk_chord_2024/Readme.txt")
```

Every set of data in this folder has a data file and an accompanying meta file.

# Clean data

## clinical_patient

```{r}
data_clinical_patient <- read_delim("../data/_raw/msk_chord_2024/data_clinical_patient.txt", 
                                    delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_clinical_patient.txt")

head(data_clinical_patient) # Need to remove 4 lines of column descriptions

data_clinical_patient <- read_delim("../data/_raw/msk_chord_2024/data_clinical_patient.txt", 
                                    delim = "\t", 
                                    skip = 4)

head(data_clinical_patient)

write_tsv(x = data_clinical_patient,
          file = "../data/01_data_clinical_patient.tsv")
```

## clinical_sample

```{r}
data_clinical_sample <- read_delim("../data/_raw/msk_chord_2024/data_clinical_sample.txt", 
                                   delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_clinical_sample.txt")

head(data_clinical_sample) # Need to remove 4 lines of column descriptions

data_clinical_sample <- read_delim("../data/_raw/msk_chord_2024/data_clinical_sample.txt",
                                   delim = "\t", 
                                   skip = 4)

head(data_clinical_sample)

write_tsv(x = data_clinical_sample,
          file = "../data/01_data_clinical_sample.tsv")
```

## cna_hg19

```{r}
data_cna_hg19 <- read_delim("../data/_raw/msk_chord_2024/data_cna_hg19.seg",
                            delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_cna_hg19_seg.txt")

head(data_cna_hg19) 

write_tsv(x = data_cna_hg19,
          file = "../data/01_data_cna_hg19.tsv")
```

## cna

```{r}
data_cna <- read_delim("../data/_raw/msk_chord_2024/data_cna.txt",
                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_cna.txt")

head(data_cna) 

write_tsv(x = data_cna,
          file = "../data/01_data_cna.tsv")
```

## gene_panel_matrix

```{r}
data_gene_panel_matrix <- read_delim("../data/_raw/msk_chord_2024/data_gene_panel_matrix.txt",
                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_gene_panel_matrix.txt")

head(data_gene_panel_matrix) 

write_tsv(x = data_gene_panel_matrix,
          file = "../data/01_data_gene_panel_matrix.tsv")
```

## mutations

```{r}
data_mutations <- read_delim("../data/_raw/msk_chord_2024/data_mutations.txt",
                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_mutations.txt")

head(data_mutations) 

write_tsv(x = data_mutations,
          file = "../data/01_data_mutations.tsv")
```

## sv

```{r}
data_sv <- read_delim("../data/_raw/msk_chord_2024/data_sv.txt",
                       delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_sv.txt")

head(data_sv) 

write_tsv(x = data_sv,
          file = "../data/01_data_sv.tsv")
```

## Timeline data

Exploring selected

-   data_timeline_ca_15-3_labs
-   data_timeline_ca_19-9_labs
-   **data_timeline_cancer_presence**
-   data_timeline_cea_labs
-   **data_timeline_diagnosis**
-   data_timeline_gleason
-   data_timeline_mmr
-   data_timeline_pdl1
-   **data_timeline_performance_status**
-   data_timeline_prior_meds
-   **data_timeline_progression**
-   data_timeline_psa_labs
-   data_timeline_radiation
-   data_timeline_specimen_surgery
-   data_timeline_specimen
-   **data_timeline_surgery**
-   **data_timeline_treatment**
-   data_timeline_tumor_sites

### cancer_presence

```{r}
data_timeline_cancer_presence <- read_delim("../data/_raw/msk_chord_2024/data_timeline_cancer_presence.txt",
                                            delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_cancer_presence.txt")

head(data_timeline_cancer_presence) 

write_tsv(x = data_timeline_cancer_presence,
          file = "../data/01_data_timeline_cancer_presence.tsv")
```

### diagnosis

```{r}
data_timeline_diagnosis <- read_delim("../data/_raw/msk_chord_2024/data_timeline_diagnosis.txt",
                                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_diagnosis.txt")

head(data_timeline_diagnosis) 

write_tsv(x = data_timeline_diagnosis,
          file = "../data/01_data_timeline_diagnosis.tsv")
```

### performance_status

```{r}
data_timeline_performance_status <- read_delim("../data/_raw/msk_chord_2024/data_timeline_performance_status.txt",
                                              delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_performance_status.txt")

head(data_timeline_performance_status) 

write_tsv(x = data_timeline_performance_status,
          file = "../data/01_data_timeline_performance_status.tsv")
```

### progression

```{r}
data_timeline_progression <- read_delim("../data/_raw/msk_chord_2024/data_timeline_progression.txt",
                                        delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_progression.txt")

head(data_timeline_progression) 

write_tsv(x = data_timeline_progression,
          file = "../data/01_data_timeline_progression.tsv")
```

### surgery

```{r}
data_timeline_surgery <- read_delim("../data/_raw/msk_chord_2024/data_timeline_surgery.txt",
                                    delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_surgery.txt")

head(data_timeline_surgery) 

write_tsv(x = data_timeline_surgery,
          file = "../data/01_data_timeline_surgery.tsv")
```

### treatment

```{r}
data_timeline_treatment <- read_delim("../data/_raw/msk_chord_2024/data_timeline_treatment.txt",
                                      delim = "\t")

readLines("../data/_raw/msk_chord_2024/meta_timeline_treatment.txt")

head(data_timeline_treatment) 

write_tsv(x = data_timeline_treatment,
          file = "../data/01_data_timeline_treatment.tsv")
```

# Explore data

## clinical_patient

```{r}
clinical_patient_p1 <- data_clinical_patient |> 
    group_by(GENDER) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = GENDER)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void() 

clinical_patient_p2 <- data_clinical_patient |> 
  group_by(RACE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = RACE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

clinical_patient_p3 <- ggplot(data = data_clinical_patient,
                              mapping = aes(x = CURRENT_AGE_DEID)) + 
  geom_histogram() + 
  theme_minimal() 

clinical_patient_p4 <- data_clinical_patient |> 
  group_by(STAGE_HIGHEST_RECORDED) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = STAGE_HIGHEST_RECORDED)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

clinical_patient_p5 <- data_clinical_patient |> 
  group_by(OS_STATUS) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = OS_STATUS)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()


(clinical_patient_p5  + clinical_patient_p1) / (clinical_patient_p3 + plot_spacer()) / (clinical_patient_p4 + clinical_patient_p2) + 
  plot_annotation(
    title = "Overview of patient attributes in MSK-CHORD"
  )

ggsave(filename = "../results/01_clinical_patient_overview.png")
```

## clinical_sample

```{r}
clinical_sample_p1 <- data_clinical_sample |> 
  group_by(PATIENT_ID) |> 
  summarise(n_samples = n()) |> 
  group_by(n_samples) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_samples,
                       y = n)) +
  geom_col()

clinical_sample_p2 <- data_clinical_sample |> 
  group_by(CANCER_TYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = CANCER_TYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

clinical_sample_p3 <- data_clinical_sample |> 
  group_by(SAMPLE_TYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = SAMPLE_TYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

clinical_sample_p1  + clinical_sample_p2 + clinical_sample_p3 + 
  plot_annotation(
    title = "Overview of sample attributes in MSK-CHORD"
  )

ggsave(filename = "../results/01_clinical_sample_overview.png")
```

## cna_hg19

```{r}
cna_hg19_p1 <- ggplot(data = data_cna_hg19,
                      mapping = aes(x = num.mark)) + 
  geom_histogram() + 
  theme_minimal() +
  scale_x_log10() + 
  labs(
    title = "Number of markers used to assess segment mean"
  )


cna_hg19_p2 <- ggplot(data = data_cna_hg19,
                      mapping = aes(x = seg.mean)) + 
  geom_density() + 
  theme_minimal() + 
  labs(
    title = "Distribution of segment means",
    subtitle = "Postitive = gain. Negative = loss."
  )

cna_hg19_p1 / cna_hg19_p2 + 
  plot_annotation(
    title = "Overview of copy nummer alteration results compared the to HG19 reference genome"
  )

ggsave(filename = "../results/01_cna_hg19_overview.png")
```

## cna

```{r}
nrow(data_cna) # number of genes
ncol(data_cna) - 1 # number of samples
```

## gene_panel_matrix

```{r}
gene_panel_matrix_p1 <- ggplot(data = data_gene_panel_matrix,
                               mapping = aes(x = mutations,
                                             fill = mutations)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

gene_panel_matrix_p2 <- ggplot(data = data_gene_panel_matrix,
                               mapping = aes(x = cna,
                                             fill = cna)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

gene_panel_matrix_p3 <- ggplot(data = data_gene_panel_matrix,
                               mapping = aes(x = structural_variants,
                                             fill = structural_variants)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

gene_panel_matrix_p1 / gene_panel_matrix_p2 / gene_panel_matrix_p3 + 
  plot_annotation(
    title = "Distribution of gene panels in MSK-CHORD"
  )

ggsave(filename = "../results/01_gene_panel_matrix_distribution.png")
```

## mutations

```{r}
mutations_p1 <- ggplot(data = data_mutations,
                       mapping = aes(x = Variant_Type,
                                     fill = Variant_Type)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

mutations_p2 <- ggplot(data = data_mutations,
                       mapping = aes(x = Variant_Classification,
                                     fill = Variant_Classification)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

mutations_p3 <- ggplot(data = data_mutations,
                       mapping = aes(x = Mutation_Status,
                                     fill = Mutation_Status)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")


mutations_p1 / mutations_p2 / mutations_p3 + 
  plot_annotation(
    title = "Overview of mutations data in MSK-CHORD"
  )

ggsave(filename = "../results/01_mutations_overview.png")
```

## sv

```{r}
sv_p1 <- ggplot(data = data_sv,
                mapping = aes(x = Class,
                              fill = Class)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

sv_p2 <- ggplot(data = data_sv,
                mapping = aes(x = Connection_Type,
                              fill = Connection_Type)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

sv_p3 <- ggplot(data = data_sv,
                mapping = aes(x = Breakpoint_Type,
                              fill = Breakpoint_Type)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

sv_p1 / sv_p2 / sv_p3 + 
  plot_annotation(
    title = "Overview of structural variant data in MSK-CHORD"
  )

ggsave(filename = "../results/01_sv_overview.png")
```

## Timeline data

### cancer_presence

```{r}
timeline_cancer_presence_p1 <- data_timeline_cancer_presence |> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_cancer_presence_p2 <- data_timeline_cancer_presence |> 
  group_by(HAS_CANCER) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = HAS_CANCER)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_cancer_presence_p3 <- data_timeline_cancer_presence |> 
  group_by(PROCEDURE_TYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = PROCEDURE_TYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_cancer_presence_p1 / timeline_cancer_presence_p2 / timeline_cancer_presence_p3 + 
  plot_annotation(
    title = "Timeline data for cancer presence"
  )

ggsave(filename = "../results/01_timeline_cancer_presence_overview.png")
```

### diagnosis

```{r}
timeline_diagnosis_p1 <- data_timeline_diagnosis|> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_diagnosis_p2 <- data_timeline_diagnosis |> 
  group_by(STAGE_CDM_DERIVED) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = STAGE_CDM_DERIVED)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_diagnosis_p3 <- data_timeline_diagnosis |> 
  group_by(SUMMARY) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = SUMMARY)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()



timeline_diagnosis_p1 / timeline_diagnosis_p2 / timeline_diagnosis_p3 + 
  plot_annotation(
    title = "Timeline data for diagnosis"
  )

ggsave(filename = "../results/01_timeline_diagnosis_overview.png")
```

### performance_status

```{r}
timeline_performance_status_p1 <- data_timeline_performance_status |> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_performance_status_p2 <- ggplot(data = data_timeline_performance_status,
                mapping = aes(x = ECOG)) + 
  geom_bar() + 
  theme_minimal() + 
  theme(legend.position = "None")

timeline_performance_status_p1+ timeline_performance_status_p2 + 
  plot_annotation(
    title = "Timeline data for performance status or ECOG score"
  )

ggsave(filename = "../results/01_timeline_performance_status_overview.png")
```

### progression

```{r}
timeline_progression_p1 <- data_timeline_progression |> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_progression_p2 <- data_timeline_progression |> 
  group_by(PROGRESSION) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = PROGRESSION)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_progression_p3 <- data_timeline_progression |> 
  group_by(PROCEDURE_TYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = PROCEDURE_TYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_progression_p1 / timeline_progression_p2 / timeline_progression_p3 + 
  plot_annotation(
    title = "Timeline data for progression"
  )

ggsave(filename = "../results/01_timeline_progression_overview.png")
```

### surgery

```{r}
timeline_surgery_p1 <- data_timeline_surgery |> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_surgery_p2 <- data_timeline_surgery |> 
  group_by(SUBTYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = SUBTYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_surgery_p3 <- data_timeline_surgery |> 
  group_by(PATIENT_ID, SUBTYPE) |> 
  summarise(n_records = n()) |> 
  group_by(n_records, SUBTYPE) |>  
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n,
                       fill = SUBTYPE)) +
  geom_col() + 
  facet_wrap(~ SUBTYPE) +
  theme(legend.position = "None")

(timeline_surgery_p1 + timeline_surgery_p2) / timeline_surgery_p3 + 
  plot_annotation(
    title = "Timeline data for surgery"
  )

ggsave(filename = "../results/01_timeline_surgery_overview.png")
```

### treatment

```{r}
timeline_treatment_p1 <- data_timeline_treatment |> 
  group_by(PATIENT_ID) |> 
  summarise(n_records = n()) |> 
  group_by(n_records) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = n_records,
                       y = n)) +
  geom_col()

timeline_treatment_p2 <- data_timeline_treatment |> 
  group_by(SUBTYPE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = SUBTYPE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_treatment_p3 <- data_timeline_treatment |> 
  group_by(RX_INVESTIGATIVE) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = RX_INVESTIGATIVE)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

timeline_treatment_p4 <- data_timeline_treatment |> 
  group_by(FLAG_OROTOPICAL) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = "", 
                       y = n, 
                       fill = FLAG_OROTOPICAL)) +
  geom_bar(stat = "identity", 
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void()

(timeline_treatment_p1 + timeline_treatment_p2) / (timeline_treatment_p3 + timeline_treatment_p4) +
  plot_annotation(
    title = "Timeline data for treatment"
  )

ggsave(filename = "../results/01_timeline_treatment_overview.png")
```


