---
title: "Define Progression-Free Survival"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
library("tidyverse")
library("survival")
library("survminer")
library("patchwork")

# Fix mistake in survminer
is_ggplot <- function(x)is.ggplot(x)
```

# Load data

```{r}
data_timeline_progression <- read_tsv("../data/01_data_timeline_progression.tsv")
data_timeline_cancer_presence <- read_tsv("../data/01_data_timeline_cancer_presence.tsv")
data_clinical_patient <- read_tsv("../data/01_data_clinical_patient.tsv")
data_timeline_diagnosis <- read_tsv("../data/01_data_timeline_diagnosis.tsv")
data_timeline_surgery <- read_tsv("../data/01_data_timeline_surgery.tsv")
data_timeline_treatment <- read_tsv("../data/01_data_timeline_treatment.tsv")
data_clinical_sample <- read_tsv("../data/01_data_clinical_sample.tsv")
```

# Define PFS

The [NCI Dictionary of Cancer Terms](https://www.cancer.gov/publications/dictionaries/cancer-terms/def/progression-free-survival "NCI Dictionary of Cancer Terms: Progression-free survival") defined PFS as follows:

> The length of time during and after the treatment of a disease, such as cancer, that a patient lives with the disease but it does not get worse. In a clinical trial, measuring the progression-free survival is one way to see how well a new treatment works. Also called PFS.

I define PFS based on a chosen timepoint 0 and a relevant PFS end-point.
In the timeline data, timepoint 0 is defined as the time of sequencing. This can be far after sample acquisition, diagnosis or surgery, so it is not necessarily a good timepoint 0 for the purposes of this project. I define three options for timepoint 0 and will discuss with my supervisors which to use. Relevant timepoint 0's include:
1. Date of first diagnosis, defined by data_timeline_diagnosis;
2. Date of first treatment, defined by data_timeline_treatment and data_timeline_surgery;
3. Date of first surgery, defined by data_timeline_surgery.

Relevant PFS end-points include:
1. Progression, defined by data_timeline_progression;
2. Recurrence, defined by data_timeline_cancer_presence;
3. Death, defined by data_clinical_patient;
4. Censored or last follow-up, defined by data_clinical patient.

## Create startpoint data

For each patient, I first define the diagnosis date, date of first surgery, and date of first treatment. These form the foundations of three potential PFS datasets with different start dates.

```{r}
PFS_from_diagnosis <- data_timeline_diagnosis |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  select(PATIENT_ID,
         DATE_DIAGNOSIS = START_DATE)

PFS_from_first_surgery <- data_timeline_surgery |> 
  filter(SUBTYPE == "PROCEDURE") |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  select(PATIENT_ID,
         DATE_FIRST_SURGERY = START_DATE)

PFS_from_first_treatment <- data_timeline_treatment |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  select(PATIENT_ID,
         DATE_FIRST_NON_SURGICAL_TREATMENT = START_DATE) |> 
  full_join(PFS_from_first_surgery, join_by(PATIENT_ID)) |> 
  mutate(DATE_FIRST_TREATMENT = case_when(is.na(DATE_FIRST_NON_SURGICAL_TREATMENT) ~ DATE_FIRST_SURGERY,
                                          is.na(DATE_FIRST_SURGERY) ~ DATE_FIRST_NON_SURGICAL_TREATMENT,
                                          .default = min(DATE_FIRST_NON_SURGICAL_TREATMENT, DATE_FIRST_SURGERY))) |> 
  select(PATIENT_ID,
         DATE_FIRST_TREATMENT)
```

Looking at how far the start date is from the date the samples were sequenced:

```{r}
p1 <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample) |> 
  ggplot(mapping = aes(x = -DATE_DIAGNOSIS,
                       y = CANCER_TYPE,
                       color = CANCER_TYPE,
                       fill = CANCER_TYPE))+
  geom_boxplot(alpha = 0.5) +
  theme_bw() + 
  theme(
    axis.title.y = element_blank(),
    legend.position = "None"
  ) + 
  labs(
    x = "Days",
    title = "Time between date of diagnosis and inclusion into the cohort"
  )

p2 <- PFS_from_first_surgery |> 
  left_join(data_clinical_sample) |> 
  ggplot(mapping = aes(x = -DATE_FIRST_SURGERY,
                       y = CANCER_TYPE,
                       color = CANCER_TYPE,
                       fill = CANCER_TYPE))+
  geom_boxplot(alpha = 0.5) +
  theme_bw() + 
  theme(
    axis.title.y = element_blank(),
    legend.position = "None"
  ) + 
  labs(
    x = "Days",
    title = "Time between date of first surgery and inclusion into the cohort"
  )

p3 <- PFS_from_first_treatment |> 
  left_join(data_clinical_sample) |> 
  ggplot(mapping = aes(x = -DATE_FIRST_TREATMENT,
                       y = CANCER_TYPE,
                       color = CANCER_TYPE,
                       fill = CANCER_TYPE))+
  geom_boxplot(alpha = 0.5) +
  theme_bw() + 
  theme(
    axis.title.y = element_blank(),
    legend.position = "None"
  ) + 
  labs(
    x = "Days",
    title = "Time between date of first treatment and inclusion into the cohort"
  )

p1 / p2 / p3 + 
  plot_layout(
    axis_titles = "collect"
  )

ggsave(filename = "../results/02_start_time_vs_inclusion.png", width = 3200, height = 3200, units = "px")
```


## Create PFS data

Now I will create a datasets for the PFS relevant events, based on using diagnosis as timepoint 0.

```{r}
# Progression
PFS_from_diagnosis <- PFS_from_diagnosis |> 
  left_join(data_timeline_progression, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_DIAGNOSIS,
         PROGRESSION == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_PROGRESSION = min(START_DATE)) |> 
  right_join(PFS_from_diagnosis, join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = !is.na(DATE_FIRST_PROGRESSION))

# Recurrence
PFS_from_diagnosis <- PFS_from_diagnosis |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_DIAGNOSIS) |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  filter(HAS_CANCER == "N") |> 
  select(PATIENT_ID, DATE_DIAGNOSIS, PROGRESSION, DATE_FIRST_PROGRESSION) |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_DIAGNOSIS,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_RECURRENCE = min(START_DATE)) |> 
  right_join(PFS_from_diagnosis, join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_FIRST_RECURRENCE)) 

# Death
PFS_from_diagnosis <- PFS_from_diagnosis |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, OS_MONTHS) |> 
  right_join(PFS_from_diagnosis, join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censor
PFS_from_diagnosis <- PFS_from_diagnosis |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  group_by(PATIENT_ID) |> 
  mutate(PFS_STATUS = ifelse(PROGRESSION == FALSE & RECURRENCE == FALSE & DEATH == FALSE, "0:No progression", "1:Progression event"),
         PFS_DATE = ifelse(PFS_STATUS == "1:Progression event", pmin(DATE_FIRST_PROGRESSION, DATE_FIRST_RECURRENCE, DATE_DEATH, na.rm = TRUE), round(OS_MONTHS*30.44)),
         DAYS_TO_EVENT = PFS_DATE - DATE_DIAGNOSIS,
         YEARS_TO_EVENT = DAYS_TO_EVENT/365.25) |> 
  select(PATIENT_ID, PFS_STATUS, DAYS_TO_EVENT, YEARS_TO_EVENT, PFS_DATE, DATE_DIAGNOSIS, PROGRESSION, DATE_FIRST_PROGRESSION, RECURRENCE, DATE_FIRST_RECURRENCE, DEATH, DATE_DEATH)

write_tsv(x = PFS_from_diagnosis,
          file = "../data/02_PFS_from_diagnosis.tsv")

PFS_from_diagnosis |> 
  group_by(PFS_STATUS) |> 
  summarise(n = n(), pct = n()/nrow(PFS_from_diagnosis))

PFS_from_diagnosis |> 
  ggplot(mapping = aes(x = YEARS_TO_EVENT,
                       fill = PFS_STATUS)) + 
  geom_histogram(color = "white") + 
  facet_wrap(~ PFS_STATUS, dir = "v") + 
  theme_minimal() + 
  theme(legend.position = "None")

ggsave(filename = "../results/02_years_to_progression_from_diagnosis.png")
```

Based on first surgery as timepoint 0.

```{r}
# Progression
  PFS_from_first_surgery <- PFS_from_first_surgery |> 
  left_join(data_timeline_progression, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         PROGRESSION == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_PROGRESSION = min(START_DATE)) |> 
  right_join(PFS_from_first_surgery, join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = !is.na(DATE_FIRST_PROGRESSION))

# Recurrence
PFS_from_first_surgery <- PFS_from_first_surgery |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  filter(HAS_CANCER == "N") |> 
  select(PATIENT_ID, DATE_FIRST_SURGERY, PROGRESSION, DATE_FIRST_PROGRESSION) |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_RECURRENCE = min(START_DATE)) |> 
  right_join(PFS_from_first_surgery, join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_FIRST_RECURRENCE)) 

# Death
PFS_from_first_surgery <- PFS_from_first_surgery |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, OS_MONTHS) |> 
  right_join(PFS_from_first_surgery, join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censor
PFS_from_first_surgery <- PFS_from_first_surgery |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  group_by(PATIENT_ID) |> 
  mutate(PFS_STATUS = ifelse(PROGRESSION == FALSE & RECURRENCE == FALSE & DEATH == FALSE, "0:No progression", "1:Progression event"),
         PFS_DATE = ifelse(PFS_STATUS == "1:Progression event", pmin(DATE_FIRST_PROGRESSION, DATE_FIRST_RECURRENCE, DATE_DEATH, na.rm = TRUE), round(OS_MONTHS*30.44)),
         DAYS_TO_EVENT = PFS_DATE - DATE_FIRST_SURGERY,
         YEARS_TO_EVENT = DAYS_TO_EVENT/365.25) |> 
  select(PATIENT_ID, PFS_STATUS, DAYS_TO_EVENT, YEARS_TO_EVENT, PFS_DATE, DATE_FIRST_SURGERY, PROGRESSION, DATE_FIRST_PROGRESSION, RECURRENCE, DATE_FIRST_RECURRENCE, DEATH, DATE_DEATH)

write_tsv(x = PFS_from_first_surgery,
          file = "../data/02_PFS_from_first_surgery.tsv")

PFS_from_first_surgery |> 
  group_by(PFS_STATUS) |> 
  summarise(n = n(), pct = n()/nrow(PFS_from_first_surgery))

PFS_from_first_surgery |> 
  ggplot(mapping = aes(x = YEARS_TO_EVENT,
                       fill = PFS_STATUS)) + 
  geom_histogram(color = "white") + 
  facet_wrap(~ PFS_STATUS, dir = "v") + 
  theme_minimal() + 
  theme(legend.position = "None")

ggsave(filename = "../results/02_years_to_progression_from_first_surgery.png")
```

Based on first treatment as timepoint 0.

```{r}
# Progression
  PFS_from_first_treatment <- PFS_from_first_treatment |> 
  left_join(data_timeline_progression, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_TREATMENT,
         PROGRESSION == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_PROGRESSION = min(START_DATE)) |> 
  right_join(PFS_from_first_treatment, join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = !is.na(DATE_FIRST_PROGRESSION))

# Recurrence
PFS_from_first_treatment <- PFS_from_first_treatment |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_TREATMENT) |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  filter(HAS_CANCER == "N") |> 
  select(PATIENT_ID, DATE_FIRST_TREATMENT, PROGRESSION, DATE_FIRST_PROGRESSION) |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_TREATMENT,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_RECURRENCE = min(START_DATE)) |> 
  right_join(PFS_from_first_treatment, join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_FIRST_RECURRENCE)) 

# Death
PFS_from_first_treatment <- PFS_from_first_treatment |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, OS_MONTHS) |> 
  right_join(PFS_from_first_treatment, join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censor
PFS_from_first_treatment <- PFS_from_first_treatment |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  group_by(PATIENT_ID) |> 
  mutate(PFS_STATUS = ifelse(PROGRESSION == FALSE & RECURRENCE == FALSE & DEATH == FALSE, "0:No progression", "1:Progression event"),
         PFS_DATE = ifelse(PFS_STATUS == "1:Progression event", pmin(DATE_FIRST_PROGRESSION, DATE_FIRST_RECURRENCE, DATE_DEATH, na.rm = TRUE), round(OS_MONTHS*30.44)),
         DAYS_TO_EVENT = PFS_DATE - DATE_FIRST_TREATMENT,
         YEARS_TO_EVENT = DAYS_TO_EVENT/365.25) |> 
  select(PATIENT_ID, PFS_STATUS, DAYS_TO_EVENT, YEARS_TO_EVENT, PFS_DATE, DATE_FIRST_TREATMENT, PROGRESSION, DATE_FIRST_PROGRESSION, RECURRENCE, DATE_FIRST_RECURRENCE, DEATH, DATE_DEATH)

write_tsv(x = PFS_from_first_treatment,
          file = "../data/02_PFS_from_first_treatment.tsv")

PFS_from_first_treatment |> 
  group_by(PFS_STATUS) |> 
  summarise(n = n(), pct = n()/nrow(PFS_from_first_treatment))

PFS_from_first_treatment |> 
  ggplot(mapping = aes(x = YEARS_TO_EVENT,
                       fill = PFS_STATUS)) + 
  geom_histogram(color = "white") + 
  facet_wrap(~ PFS_STATUS, dir = "v") + 
  theme_minimal() + 
  theme(legend.position = "None")

ggsave(filename = "../results/02_years_to_progression_from_first_treatment.png")
```

# Kaplan Meier plots

```{r}
PFS_from_diagnosis_bin <- PFS_from_diagnosis |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))

fit_PFS_from_diagnosis <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_bin)

ggsurvplot(
  fit = fit_PFS_from_diagnosis,
  data = PFS_from_diagnosis_bin,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Kaplan-Meier plot for Progression Free Survival from Diagnosis"
  )

ggsave(filename = "../results/02_KM_from_diagnosis.png")
```

```{r}
PFS_from_first_surgery_bin <- PFS_from_first_surgery |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))

fit_PFS_from_first_surgery <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_first_surgery_bin)

ggsurvplot(
  fit = fit_PFS_from_first_surgery,
  data = PFS_from_first_surgery_bin,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Kaplan-Meier plot for Progression Free Survival from First Surgery"
  )

ggsave(filename = "../results/02_KM_from_first_surgery.png")
```

```{r}
PFS_from_first_treatment_bin <- PFS_from_first_treatment |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))

fit_PFS_from_first_treatment <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_first_treatment_bin)

ggsurvplot(
  fit = fit_PFS_from_first_treatment,
  data = PFS_from_first_treatment_bin,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Kaplan-Meier plot for Progression Free Survival from First Treatment"
  )

ggsave(filename = "../results/02_KM_from_first_treatment.png")
```

## Per cancer type (from diagnosis)

```{r}
PFS_from_diagnosis_CRC <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))
PFS_from_diagnosis_BC <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Breast Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))
PFS_from_diagnosis_NSCLC <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Non-Small Cell Lung Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))
PFS_from_diagnosis_PaC <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Pancreatic Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))
PFS_from_diagnosis_PrC <- PFS_from_diagnosis |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Prostate Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))


fit_PFS_from_diagnosis_CRC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_CRC)
fit_PFS_from_diagnosis_BC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_BC)
fit_PFS_from_diagnosis_NSCLC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_NSCLC)
fit_PFS_from_diagnosis_PaC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_PaC)
fit_PFS_from_diagnosis_PrC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_diagnosis_PrC)

p1 <- ggsurvplot(
  fit = fit_PFS_from_diagnosis_CRC,
  data = PFS_from_diagnosis_CRC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Colorectal Cancer",
    subtitle = str_c("n = ", length(unique(PFS_from_diagnosis_CRC$PATIENT_ID)))
  )

p2 <- ggsurvplot(
  fit = fit_PFS_from_diagnosis_BC,
  data = PFS_from_diagnosis_BC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Breast Cancer",
    subtitle = str_c("n = ", length(unique(PFS_from_diagnosis_BC$PATIENT_ID)))
  )

p3 <- ggsurvplot(
  fit = fit_PFS_from_diagnosis_NSCLC,
  data = PFS_from_diagnosis_NSCLC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Non-Small Cell Lung Cancer",
    subtitle = str_c("n = ", length(unique(PFS_from_diagnosis_NSCLC$PATIENT_ID)))
  )

p4 <- ggsurvplot(
  fit = fit_PFS_from_diagnosis_PaC,
  data = PFS_from_diagnosis_PaC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Pancreatic Cancer",
    subtitle = str_c("n = ", length(unique(PFS_from_diagnosis_PaC$PATIENT_ID)))
  )

p5 <- ggsurvplot(
  fit = fit_PFS_from_diagnosis_PrC,
  data = PFS_from_diagnosis_PrC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "none"
) + 
  labs(
    title = "Prostate Cancer",
    subtitle = str_c("n = ", length(unique(PFS_from_diagnosis_PrC$PATIENT_ID)))
  )

(p1$plot + p2$plot + p3$plot) / (p4$plot + p5$plot + plot_spacer()) + 
  plot_annotation(
  title = "Kaplan-Meier plots for Progression Free Survival from Diagnosis"
  ) +
  plot_layout(
    axis_titles = "collect"
  )

ggsave(filename = "../results/02_KM_from_diagnosis_per_type.png", width = 3200, height = 2000, units = "px")
```

## CRC PFS from surgery

```{r}
PFS_from_surgery_CRC <- PFS_from_first_surgery |> 
  left_join(data_clinical_sample, join_by(PATIENT_ID)) |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))

fit_PFS_from_surgery_CRC <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = PFS_from_surgery_CRC)

ggsurvplot(
  fit = fit_PFS_from_surgery_CRC,
  data = PFS_from_surgery_CRC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  risk.table = T,
  break.time.by = 2,
  legend = "none",
  title = "Progression-free survival after CRC surgery"
) 

ggsave(filename = "../results/02_KM_CRC_PFS_from_surgery.png", width = 3200, height = 2000, units = "px")

PFS_from_surgery_CRC <- PFS_from_surgery_CRC |> 
  inner_join(data_clinical_patient, by = join_by(PATIENT_ID))

fit_by_stage <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ STAGE_HIGHEST_RECORDED, data = PFS_from_surgery_CRC)

ggsurvplot(
  fit = fit_by_stage,
  data = PFS_from_surgery_CRC,
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  legend = "bottom",
  title = "Progression-free survival after CRC surgery",         
  surv.median.line = "hv",
  pval = T,
  break.time.by = 1
) 

ggsave(filename = "../results/02_KM_CRC_PFS_by_stage.png", width = 3200, height = 2000, units = "px")
```