---
title: "Define and explore the cohort"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false
#| echo: false

library("tidyverse")
library("survival")
library("survminer")
library("janitor")
library("table1")
library("knitr")
```

```{r}
#| label: fix_mistake_survminer_package
#| echo: false

is_ggplot <- function(x)is.ggplot(x)
```

```{r}
#| label: load_data
#| message: false
#| echo: false

data_clinical_sample <- read_tsv("../data/01_data_clinical_sample.tsv")
data_timeline_surgery <- read_tsv("../data/01_data_timeline_surgery.tsv")
data_timeline_sample_acquisition <- read_tsv("../data/01_data_timeline_specimen_surgery.tsv")
data_timeline_cancer_presence <- read_tsv("../data/01_data_timeline_cancer_presence.tsv")
data_timeline_treatment  <- read_tsv("../data/01_data_timeline_treatment.tsv")
data_timeline_radiation  <- read_tsv("../data/01_data_timeline_radiation.tsv")
data_timeline_prior_meds <- read_tsv("../data/01_data_timeline_prior_meds.tsv")
data_clinical_patient <- read_tsv("../data/01_data_clinical_patient.tsv")
data_timeline_performance_status  <- read_tsv("../data/01_data_timeline_performance_status.tsv")
data_timeline_cea_labs <- read_tsv("../data/01_data_timeline_cea_labs.tsv")
data_timeline_tumor_site  <- read_tsv("../data/01_data_timeline_tumor_sites.tsv")
data_timeline_mmr  <- read_tsv("../data/01_data_timeline_mmr.tsv")
data_cna <- read_tsv("../data/01_data_cna.tsv")
data_mutations <- read_tsv("../data/01_data_mutations.tsv")
data_sv <- read_tsv("../data/01_data_sv.tsv")
```

This project uses the Memorial Sloan Kettering clinicogenomic, harmonized oncologic real-world dataset (MSK-CHORD), published by [Jee *et al*. (2024)](https://doi.org/10.1038/s41586-024-08167-5 "Automated real-world data integration improves cancer outcome prediction"). The dataset is a collection of clinicogenomic data and structured real-world data (RWD), captured from unstructured medical texts using natural language processing (NLP). MSK-CHORD includes data from non-small-cell lung, breast, colorectal, prostate and pancreatic cancers. 

The full MSK-CHORD dataset was downloaded from [cBioPortal](https://www.cbioportal.org/study/summary?id=msk_chord_2024 "cBioPortal: MSK-CHORD") on the 10th of June, 2025.

# Define inclusion and exclusion criteria

In this project, we are focused on stage I-III colorectal cancer (CRC) patients who underwent surgical resection of the tumor and surrounding tissues, which is intended to be curative. To define which patients should be included in this cohort, we use the relevant data that was collected within 180 days of the date of the first surgery recorded in MSK-CHORD. The following criteria were used to define what patients should be included.

1. Patients with a CRC diagnosis were selected.

```{r}
#| label: CRC_patients

CRC_patients <- data_clinical_sample |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()

length(CRC_patients)
```

2. Patients who underwent surgery were selected and the date of their first surgery was identified.

```{r}
#| label: surgery_dates

surgery_dates <- data_timeline_surgery |> 
  filter(SUBTYPE == "PROCEDURE",
         PATIENT_ID %in% CRC_patients) |> 
  group_by(PATIENT_ID) |> 
  summarise(SURGERY_DATE = min(START_DATE))

nrow(surgery_dates)
```

3. Patients who had a sample taken within 180 days of the surgery were included.

```{r}
#| label: sample_dates

patients_sampled <- data_timeline_sample_acquisition |> 
  right_join(surgery_dates,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 
                 180)) |> 
  select(PATIENT_ID, 
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE = START_DATE)

n_distinct(patients_sampled$PATIENT_ID)
```

4. Patients whose sample indicated a metastasis were excluded.

```{r}
#| label: exclude_metastases

patients_metastatic <- data_clinical_sample |> 
  right_join(patients_sampled,
             by = join_by(PATIENT_ID,
                          SAMPLE_ID)) |> 
  filter(SAMPLE_TYPE == "Metastasis") |> 
  pull(PATIENT_ID) |> 
  unique()

patients_included <- patients_sampled |> 
  filter(!PATIENT_ID %in% patients_metastatic)

n_distinct(patients_included$PATIENT_ID)
```

5. Patients who were diagnosed with cancers other than CRC were excluded.

```{r}
#| label: exclude_multiple_cancers

patients_multiple_cancers <- patients_included |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  filter(n > 1) |> 
  left_join(data_clinical_sample,
            by = join_by(PATIENT_ID)) |> 
  filter(!CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()

patients_included <- patients_included |> 
  filter(!PATIENT_ID %in% patients_multiple_cancers)

n_distinct(patients_included$PATIENT_ID)
```

6. Patients whose sample indicates stage IV CRC were excluded, because their surgery would not be intended to be curative.

```{r}
#| label: exclude_stage_4

patients_included <- data_clinical_sample |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID, 
                          SAMPLE_ID)) |> 
  filter(CANCER_TYPE == "Colorectal Cancer",
         SAMPLE_TYPE == "Primary",
         !str_starts(PATHOLOGICAL_GROUP, 
                     "4")) |> 
  select(PATIENT_ID,
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE)

n_distinct(patients_included$PATIENT_ID)
```

7. If a patient had imaging data from the first 6 months after the surgery, patients whose most recent scan indicates the presence of cancer were excluded, because that indicates that the surgery was not curative.

```{r}
#| label: not_cured

patients_not_cured <- patients_included |> 
  left_join(data_timeline_cancer_presence, 
            join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 0, 180)) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  filter(HAS_CANCER == "Y") |> 
  pull(PATIENT_ID)

patients_included <- patients_included |> 
  filter(!PATIENT_ID %in% patients_not_cured)

n_distinct(patients_included$PATIENT_ID)
```

In this project, we will investigate the disease-free survival (DFS) and overall survival (OS) from 180 days after the day of the first surgery. Therefor, we must define the relevant survival events, namely recurrence (which can be indicated by imaging or treatment) and death.

```{r}
# Recurrence (indicated by imaging)
data_survival <- patients_included |> 
  left_join(data_timeline_cancer_presence, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_RECURRENCE = min(START_DATE)) |> 
  right_join(patients_included, 
             join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_RECURRENCE)) 

# Recurrence (indicated by surgery)
data_survival <- data_survival |> 
  left_join(data_timeline_surgery, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         SUBTYPE == "PROCEDURE") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_NEXT_SURGERY = min(START_DATE)) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = ifelse(RECURRENCE,
                             RECURRENCE, 
                             !is.na(DATE_NEXT_SURGERY))) 

# Recurrence (indicated by treatment)
data_survival <- data_timeline_treatment |> 
  bind_rows(data_timeline_radiation) |> 
  right_join(data_survival,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         SUBTYPE %in% c("Chemo", 
                        "Targeted", 
                        "Radiation Therapy")) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_TREATMENT = min(START_DATE)) |> 
  left_join(data_timeline_treatment, 
             by = join_by(PATIENT_ID)) |> 
  filter(SUBTYPE != "Chemo" | 
           DATE_TREATMENT - STOP_DATE > 60) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_TREATMENT = min(DATE_TREATMENT)) |> 
  right_join(data_survival, 
             by = join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = ifelse(RECURRENCE,
                             RECURRENCE, 
                             !is.na(DATE_TREATMENT)))

# Death
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, 
         OS_MONTHS) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censoring
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID))|> 
  group_by(PATIENT_ID) |> 
  mutate(OS_STATUS = ifelse(DEATH == FALSE, 
                            0, 
                            1),
         DFS_STATUS = ifelse(RECURRENCE == FALSE & 
                               DEATH == FALSE, 
                             0, 
                             1),
         OS_DATE = round(OS_MONTHS*30.44),
         DFS_DATE = ifelse(DFS_STATUS == 1, 
                           pmin(DATE_RECURRENCE, 
                                DATE_DEATH, 
                                na.rm = TRUE), 
                           OS_DATE),
         DAYS_TO_OS_EVENT = OS_DATE - SURGERY_DATE,
         DAYS_TO_DFS_EVENT = DFS_DATE - SURGERY_DATE,
         YEARS_TO_OS_EVENT = DAYS_TO_OS_EVENT/365.25, 
         YEARS_TO_DFS_EVENT = DAYS_TO_DFS_EVENT/365.25) |> 
  select(PATIENT_ID, 
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE,
         OS_STATUS, 
         DAYS_TO_OS_EVENT, 
         YEARS_TO_OS_EVENT, 
         DFS_STATUS, 
         DAYS_TO_DFS_EVENT, 
         YEARS_TO_DFS_EVENT)
```

8. Only patients survived without recurrence or death for 180 days after the surgery were included.

```{r}
patients_included <- data_survival |> 
  mutate(survived_6_months = case_when(OS_STATUS == 1 ~ DAYS_TO_OS_EVENT > 180,
                                       TRUE ~ TRUE),
         no_recurrence_6_months = case_when(DFS_STATUS == 1 ~ DAYS_TO_DFS_EVENT > 180,
                                            TRUE ~ TRUE)) |> 
  filter(survived_6_months == TRUE,
         no_recurrence_6_months == TRUE) |> 
  select(- c(survived_6_months,
             no_recurrence_6_months))

n_distinct(patients_included$PATIENT_ID)
```

Total number of patients included:

```{r}
#| label: patients_included
#| echo: false

cat(nrow(patients_included))

write_tsv(x = patients_included,
          file = "../data/02_patients_included.tsv")
```

# Explore the cohort data

The next section provides an overview of the different features that are captured in the data.

## Survival

```{r}
#| label: summarize_survival data
#| warning: false
#| echo: false
#| message: false

patients_included |> 
  group_by(OS_STATUS) |> 
  summarise(n(), 
            mean(YEARS_TO_OS_EVENT)) |> 
  kable(digits = 2)

fit_OS <- survfit(Surv(YEARS_TO_OS_EVENT, 
                       OS_STATUS) ~ 1, 
                  data = patients_included)

ggsurvplot(
  fit = fit_OS,
  data = patients_included,
  title = "KM plot for OS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T,
  surv.median.line = "hv"
) 

ggsave("../results/02_KM_OS.png")


patients_included |> 
  group_by(DFS_STATUS) |> 
  summarise(n(), 
            mean(YEARS_TO_DFS_EVENT)) |> 
  kable(digits = 2)

fit_DFS <- survfit(Surv(YEARS_TO_DFS_EVENT, 
                        DFS_STATUS) ~ 1, 
                   data = patients_included)

ggsurvplot(
  fit = fit_DFS,
  data = patients_included,
  title = "KM plot for DFS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T,
  surv.median.line = "hv"
) 

ggsave("../results/02_KM_DFS.png")
```

## Clinical

### Sample data

```{r}
#| label: clean_data_clinical_sample
#| message: false
#| echo: false

data_clinical_sample_clean <- data_clinical_sample |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID, 
                          SAMPLE_ID)) |> 
  remove_constant() |> 
  # CANCER_TYPE_DETAILED was omitted, because it contains the same information as ONCOTREE_CODE.
  # Columns with descriptions and comments for extra detail were removed.
  # GLEASON_SAMPLE_LEVEL, PDL1_POSITIVE and SOMATIC_STATUS were omitted, because they were \>90% homogeneous.
  select(- c(CANCER_TYPE_DETAILED,
             MSI_COMMENT,
             CLINICAL_SUMMARY,
             ICD_O_HISTOLOGY_DESCRIPTION,
             DIAGNOSIS_DESCRIPTION,
             GLEASON_SAMPLE_LEVEL,
             PDL1_POSITIVE,
             SOMATIC_STATUS)) |>
  mutate(PRIMARY_SITE = str_to_lower(PRIMARY_SITE),
         MSI_TYPE = ifelse(MSI_TYPE == "Do not report", 
                           NA, 
                           MSI_TYPE),
         CLINICAL_GROUP_SIMPLIFIED = str_sub(CLINICAL_GROUP, 
                                             1, 
                                             1),
         PATHOLOGICAL_GROUP_SIMPLIFIED = str_sub(PATHOLOGICAL_GROUP, 
                                                 1, 
                                                 1)) 

write_tsv(x = data_clinical_sample_clean,
          file = "../data/02_data_clinical_sample_clean.tsv")
```

```{r}
#| label: table_1_clinical_sample
#| warning: false
#| echo: false

data_clinical_sample_clean$ONCOTREE_CODE <- factor(data_clinical_sample_clean$ONCOTREE_CODE)
label(data_clinical_sample_clean$ONCOTREE_CODE) <- "Oncotree code"

data_clinical_sample_clean$PATHOLOGICAL_GROUP_SIMPLIFIED <- factor(data_clinical_sample_clean$PATHOLOGICAL_GROUP_SIMPLIFIED)
label(data_clinical_sample_clean$PATHOLOGICAL_GROUP_SIMPLIFIED) <- "Pathological stage"

data_clinical_sample_clean$CLINICAL_GROUP_SIMPLIFIED <- factor(data_clinical_sample_clean$CLINICAL_GROUP_SIMPLIFIED)
label(data_clinical_sample_clean$CLINICAL_GROUP_SIMPLIFIED) <- "Clinical stage"

data_clinical_sample_clean$PRIMARY_SITE <- factor(data_clinical_sample_clean$PRIMARY_SITE)
label(data_clinical_sample_clean$PRIMARY_SITE) <- "Tumor site"

data_clinical_sample_clean$GENE_PANEL <- factor(data_clinical_sample_clean$GENE_PANEL)
label(data_clinical_sample_clean$GENE_PANEL) <- "Gene panel"

label(data_clinical_sample_clean$SAMPLE_COVERAGE) <- "Sample coverage"

label(data_clinical_sample_clean$TUMOR_PURITY) <- "Tumor purity"

label(data_clinical_sample_clean$MSI_SCORE) <- "Microsatellite instability (MSI)"

data_clinical_sample_clean$MSI_TYPE <- factor(data_clinical_sample_clean$MSI_TYPE)
label(data_clinical_sample_clean$MSI_TYPE) <- "MSI group"

label(data_clinical_sample_clean$TMB_NONSYNONYMOUS) <- "Tumor mutational burden"

table1(~ ONCOTREE_CODE +
         PATHOLOGICAL_GROUP_SIMPLIFIED +
         CLINICAL_GROUP_SIMPLIFIED +
         PRIMARY_SITE + 
         GENE_PANEL +
         SAMPLE_COVERAGE +
         TUMOR_PURITY +
         MSI_SCORE +
         MSI_TYPE +
         TMB_NONSYNONYMOUS,
       data = data_clinical_sample_clean)
```

### Patient data

```{r}
#| label: clean_data_clinical_patient
#| message: false
#| echo: false

data_clinical_patient_clean <- data_clinical_patient|> 
  right_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  remove_constant() |> 
  # The NLP-derived tumor site data will be taken from data_timeline_tumor_sites to take the timing into account.
  # GLEASON_FIRST_REPORTED, GLEASON_HIGHEST_REPORTED, HISTORY_OF_PDL1, HR and HER2 were omitted, because they were \>90% homogeneous.
  select(- c(ADRENAL_GLANDS,
             BONE,
             CNS_BRAIN,
             INTRA_ABDOMINAL,
             LIVER,
             LUNG,
             LYMPH_NODES,
             OTHER,
             PLEURA,
             REPRODUCTIVE_ORGANS,
             GLEASON_FIRST_REPORTED, 
             GLEASON_HIGHEST_REPORTED, 
             HISTORY_OF_PDL1, 
             HR,
             HER2)) |> 
  mutate(RACE_SIMPLIFIED = case_when(RACE == "White" ~ "White",
                                     RACE == "Black Or African American" ~ "Black",
                                     RACE == "Asian-Far East/Indian Subcont" ~ "Asian",
                                     RACE == "Unknown" | 
                                       RACE == "Pt Refused to Answer" ~ NA,
                                     TRUE ~ "Other"),
         ETHNICITY_SIMPLIFIED = case_when(ETHNICITY == "Non-Spanish; Non-Hispanic" ~ "Non-Spanish; Non-Hispanic",
                                          ETHNICITY == "Unknown whether Spanish or not" ~ NA,
                                          TRUE ~ "Spanish/Hispanic"),
         EARLY_ONSET = ifelse(CURRENT_AGE_DEID < 50,
                              TRUE,
                              FALSE),
         LAST_FOLLOW_UP_DATE = round(OS_MONTHS*30.44),
         AGE_SURGERY = CURRENT_AGE_DEID - round((LAST_FOLLOW_UP_DATE - SURGERY_DATE)/365.25)) |> 
  rename(AGE_LAST_FOLLOW_UP = CURRENT_AGE_DEID,
         SMOKING_STATUS_NLP = SMOKING_PREDICTIONS_3_CLASSES,
         PRIOR_MED_NLP = PRIOR_MED_TO_MSK)

write_tsv(x = data_clinical_patient_clean,
          file = "../data/02_data_clinical_patient_clean.tsv")
```

```{r}
#| label: table_1_clinical_patient
#| warning: false
#| echo: false

data_clinical_patient_clean$GENDER <- factor(data_clinical_patient_clean$GENDER)
label(data_clinical_patient_clean$GENDER) <- "Gender"

data_clinical_patient_clean$RACE_SIMPLIFIED <- factor(data_clinical_patient_clean$RACE_SIMPLIFIED)
label(data_clinical_patient_clean$RACE_SIMPLIFIED) <- "Race"

data_clinical_patient_clean$ETHNICITY_SIMPLIFIED <- factor(data_clinical_patient_clean$ETHNICITY_SIMPLIFIED)
label(data_clinical_patient_clean$ETHNICITY_SIMPLIFIED) <- "Ethnicity"

label(data_clinical_patient_clean$AGE_SURGERY) <- "Age"

data_clinical_patient_clean$EARLY_ONSET <- factor(data_clinical_patient_clean$EARLY_ONSET)
label(data_clinical_patient_clean$EARLY_ONSET) <- "Early onset"

data_clinical_patient_clean$STAGE_HIGHEST_RECORDED <- factor(data_clinical_patient_clean$STAGE_HIGHEST_RECORDED)
label(data_clinical_patient_clean$STAGE_HIGHEST_RECORDED) <- "Stage (highest recorded)"

label(data_clinical_patient_clean$NUM_ICDO_DX) <- "Number of tumor registry entries"

data_clinical_patient_clean$SMOKING_STATUS_NLP <- factor(data_clinical_patient_clean$SMOKING_STATUS_NLP)
label(data_clinical_patient_clean$SMOKING_STATUS_NLP) <- "Smoking status (NLP)"

data_clinical_patient_clean$PRIOR_MED_NLP <- factor(data_clinical_patient_clean$PRIOR_MED_NLP)
label(data_clinical_patient_clean$PRIOR_MED_NLP) <- "Prior medication (NLP)"

table1(~ GENDER +
         RACE_SIMPLIFIED +
         ETHNICITY_SIMPLIFIED +
         AGE_SURGERY +
         EARLY_ONSET +
         STAGE_HIGHEST_RECORDED +
         NUM_ICDO_DX +
         SMOKING_STATUS_NLP +
         PRIOR_MED_NLP,
       data = data_clinical_patient_clean)
```

### ECOG scores

```{r}
#| label: clean_data_timeline_performance_status
#| message: false
#| echo: false

data_timeline_performance_status_clean <- data_timeline_performance_status |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 
                 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE,
             STYLE_COLOR)) 

write_tsv(x = data_timeline_performance_status_clean,
          file = "../data/02_data_timeline_performance_status_clean.tsv")
```

Fraction of patients in the cohort with ECOG scores:

```{r}
#| label: patients_performance_status
#| echo: false

cat(str_c(round(data_timeline_performance_status_clean |> 
                  pull(PATIENT_ID) |> 
                  n_distinct()/nrow(patients_included)*100,
                digits = 1),
          "%"))
```

Number of ECOG scores per patient:

```{r}
#| label: performance_status_measures_per_patient
#| echo: false
#| warning: false

data_timeline_performance_status_clean |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  ungroup() |> 
  summarise(mean = mean(n),
            sd = sd(n),
            median = median(n)) |> 
  kable(digits = 2)
```

Distribution of ECOG scores (showing the latest score in the 180 day window around the day of surgery per patient):

```{r}
#| label: distribution_performanc_status
#| echo: false

data_timeline_performance_status_clean |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = ECOG,
                       fill = factor(ECOG))) + 
  geom_bar() +
  labs(x = "ECOG score",
       y = "n patients") + 
  theme_minimal() +
  theme(legend.position = "None")
```

## Labs

### Carcinoembryonic antigen

```{r}
#| label: clean_data_timeline_cea_labs
#| message: false
#| echo: false

data_timeline_cea_labs_clean <- data_timeline_cea_labs |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 
                 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             LR_UNIT_MEASURE)) 

write_tsv(x = data_timeline_cea_labs_clean,
          file = "../data/02_data_timeline_cea_labs_clean.tsv")
```

Fraction of patients in the cohort with CEA labs:

```{r}
#| label: patients_cea_labs
#| echo: false

cat(str_c(round(data_timeline_cea_labs_clean |> 
                  pull(PATIENT_ID) |> 
                  n_distinct()/nrow(patients_included)*100,
                digits = 1),
          "%"))
```

Number of CEA labs per patient:

```{r}
#| label: cea_labs_per_patient
#| echo: false
#| warning: false

data_timeline_cea_labs_clean |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  ungroup() |> 
  summarise(mean = mean(n),
            sd = sd(n),
            median = median(n)) |> 
  kable(digits = 2)
```

Distribution of CEA (showing the latest measurement in the 180 day window around the day of surgery per patient):

```{r}
#| label: distribution_CEA
#| echo: false
#| warning: false

data_timeline_cea_labs_clean |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = RESULT)) + 
  geom_histogram(binwidth = 10) +
  labs(x = "CEA level (ng/ml)") + 
  theme_minimal() 

data_timeline_cea_labs_clean |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = RESULT)) + 
  geom_boxplot(outlier.shape = NA) +
  xlim(0, 10) +
  labs(x = "CEA level (ng/ml)") + 
  theme_minimal() + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

## Radiology

### Tumor site (NLP)

```{r}
#| label: clean_data_timeline_tumor_site
#| message: false
#| echo: false

data_timeline_tumor_site_clean <- data_timeline_tumor_site |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 
                 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE,
             SOURCE)) 

write_tsv(x = data_timeline_tumor_site_clean,
          file = "../data/02_data_timeline_tumor_site_clean.tsv")
```

```{r}
#| label: table_1_tumor_site
#| warning: false
#| echo: false

data_timeline_tumor_site_clean$SOURCE_SPECIFIC <- factor(data_timeline_tumor_site_clean$SOURCE_SPECIFIC)
label(data_timeline_tumor_site_clean$SOURCE_SPECIFIC) <- "Source"

data_timeline_tumor_site_clean$TUMOR_SITE <- factor(data_timeline_tumor_site_clean$TUMOR_SITE)
label(data_timeline_tumor_site_clean$TUMOR_SITE) <- "Tumor site"

data_timeline_tumor_site_clean$CHEST <- factor(data_timeline_tumor_site_clean$CHEST)
label(data_timeline_tumor_site_clean$CHEST) <- "Chest involvement"

data_timeline_tumor_site_clean$ABDOMEN <- factor(data_timeline_tumor_site_clean$ABDOMEN)
label(data_timeline_tumor_site_clean$ABDOMEN) <- "Abdomen involvement"

data_timeline_tumor_site_clean$PELVIS <- factor(data_timeline_tumor_site_clean$PELVIS)
label(data_timeline_tumor_site_clean$PELVIS) <- "Pelvis involvement"

data_timeline_tumor_site_clean$HEAD <- factor(data_timeline_tumor_site_clean$HEAD)
label(data_timeline_tumor_site_clean$HEAD) <- "Head involvement"

data_timeline_tumor_site_clean$OTHER <- factor(data_timeline_tumor_site_clean$OTHER)
label(data_timeline_tumor_site_clean$OTHER) <- "Other sites"


table1(~ SOURCE_SPECIFIC +
         TUMOR_SITE +
         CHEST +
         ABDOMEN +
         PELVIS + 
         HEAD +
         OTHER,
       data = data_timeline_tumor_site_clean)
```

## Pathology

### Mismatch repair status (NLP)

```{r}
#| label: clean_data_timeline_mmr
#| message: false
#| echo: false

data_timeline_mmr_clean <- data_timeline_mmr |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 
                 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE)) 

write_tsv(x = data_timeline_mmr_clean,
          file = "../data/02_data_timeline_mmr_clean.tsv")
```

Fraction of patients in the cohort with an MMR status:

```{r}
#| label: patients_mmr_status
#| echo: false

cat(str_c(round(data_timeline_mmr_clean |> 
                  pull(PATIENT_ID) |> 
                  n_distinct()/nrow(patients_included)*100,
                digits = 1),
          "%"))
```

Number of MMR assessments per patient:

```{r}
#| label: mmr_per_patient
#| echo: false
#| warning: false

data_timeline_mmr_clean |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  ungroup() |> 
  summarise(mean = mean(n),
            sd = sd(n),
            median = median(n)) |> 
  kable(digits = 2)
```

Distribution of MMR statuses (showing latest status in the 180 day window around the day of surgery per patient):

```{r}
#| label: distribution_mmr
#| echo: false

data_timeline_mmr_clean |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = MMR_ABSENT,
                       fill = factor(MMR_ABSENT))) + 
  geom_bar() +
  labs(x = "MMR absent",
       y = "n patients") + 
  theme_minimal() +
  theme(legend.position = "None")
```

## Treatment history

```{r}
#| label: clean_data_treatment_history
#| message: false
#| echo: false

data_timeline_treatment_clean <- data_timeline_treatment |> 
  inner_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE,
         AGENT) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - STOP_DATE)

data_timeline_radiation_clean <- data_timeline_radiation |> 
  inner_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - START_DATE, 
         AGENT = NA)

data_timeline_prior_meds_clean <- data_timeline_prior_meds |> 
  inner_join(patients_included,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - START_DATE,
         AGENT = NA)

data_treament_history_clean <- rbind(data_timeline_treatment_clean,
                                     data_timeline_radiation_clean,
                                     data_timeline_prior_meds_clean)

write_tsv(x = data_treament_history_clean,
          file = "../data/02_data_treament_history_clean.tsv")
```

Fraction of patients in the cohort who started treatment before surgery:

```{r}
#| label: patients_treatment_history_before
#| echo: false

cat(str_c(round(data_treament_history_clean |> 
                  filter(START_DATE <= SURGERY_DATE) |> 
                  pull(PATIENT_ID) |> 
                  n_distinct()/nrow(patients_included)*100,
                digits = 1),
          "%"))
```

Fraction of patients in the cohort who started treatment within 6 months after surgery:

```{r}
#| label: patients_treatment_history_after
#| echo: false

cat(str_c(round(data_treament_history_clean |> 
                  filter(START_DATE > SURGERY_DATE) |> 
                  pull(PATIENT_ID) |> 
                  n_distinct()/nrow(patients_included)*100,
                digits = 1),
          "%"))
```

Number of patients receiving different types of treatment (a patient can be included in multiple categories):

```{r}
#| label: distribution_treatment_history
#| echo: false

data_treament_history_clean |> 
  group_by(TREATMENT, 
           PATIENT_ID) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = TREATMENT,
                       fill = factor(TREATMENT))) + 
  geom_bar() +
  labs(x = "Treatment type",
       y = "n patients") + 
  theme_minimal() +
  theme(legend.position = "None",
        axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## Genomic data

### Copy number alterations 

```{r}
#| label: clean_data_cna
#| echo: false

data_cna_clean <- data_cna |> 
  pivot_longer(-Hugo_Symbol, 
               names_to = "SAMPLE_ID",
               values_to = "CNA") |> 
  filter(SAMPLE_ID %in% patients_included$SAMPLE_ID) 

write_tsv(x = data_cna_clean,
          file = "../data/02_data_cna_clean.tsv")
```

Distribution of CNA calls in the data:

```{r}
#| label: distribution_cna
#| echo: false
#| warning: false

cna_labels <- tibble(
  CNA = c(-2, -1.5, 0, 2),
  meaning = c("Deep loss", "Shallow loss", "Neutral", "Gain")
)

data_cna_clean |> 
  count(CNA) |> 
  left_join(cna_labels, by = "CNA") |> 
  ggplot(aes(x = factor(CNA), 
             y = n, 
             fill = factor(CNA))) +
  geom_col() +
  geom_text(aes(label = meaning), 
            vjust = -0.5) +
  labs(x = "CNA state", 
       y = "Count of calls") +
  theme_minimal() +
  scale_fill_manual(values = c("-2" = "blue4", 
                               "-1.5" = "skyblue", 
                               "0" = "grey80", 
                               "2" = "red")) + 
  theme(legend.position = "None")

```

CNA burden per sample:

```{r}
#| label: cna_per_sample
#| echo: false

data_cna_clean |> 
  filter(!is.na(CNA) & CNA != 0)  |> 
  count(SAMPLE_ID) |> 
  group_by(n) |> 
  summarise(samples_with_n_alterations = n()) |> 
  ggplot(mapping = aes(x = n, 
                       y = samples_with_n_alterations)) +
  geom_col() +
  labs(x = "N", 
       y = "Number of samples with N copy number alterations") + 
  theme_minimal()
```

Most frequently altered genes:

```{r}
#| label: cna_per_gene
#| echo: false

data_cna_clean |> 
  filter(!is.na(CNA) & CNA != 0)  |> 
  count(Hugo_Symbol) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  ggplot(mapping = aes(x = reorder(Hugo_Symbol, n), 
                       y = n/nrow(data_clinical_sample)*100)) +
  geom_col() +
  labs(x = "Gene", 
       y = "Number of altered samples (%)") + 
  theme_minimal() + 
  coord_flip()
```

### Mutations

```{r}
#| label: clean_data_mutations
#| echo: false

data_mutations_clean <- data_mutations |> 
  select(SAMPLE_ID = Tumor_Sample_Barcode,
         Hugo_Symbol,
         Variant_Classification,
         Variant_Type) |> 
  filter(SAMPLE_ID %in% patients_included$SAMPLE_ID) |> 
  group_by(SAMPLE_ID, Hugo_Symbol) |> 
  slice(1) |> 
  ungroup()

write_tsv(x = data_mutations_clean,
          file = "../data/02_data_mutations_clean.tsv")
```

Distribution of mutation types in the data:

```{r}
#| label: distribution_mutations
#| echo: false

data_mutations_clean |> 
  count(Variant_Type) |> 
  ggplot(mapping = aes(x = Variant_Type, 
                        y = n, 
                        fill = Variant_Type)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.5) +
  labs(x = "Variant type", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None")

data_mutations_clean |> 
  count(Variant_Classification) |> 
  ggplot(mapping = aes(x = Variant_Classification, 
                        y = n, 
                        fill = Variant_Classification)) +
  geom_col() +
  geom_text(aes(label = n), 
            hjust = -0.5) +
  labs(x = "Variant classification", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None") + 
  coord_flip()
```

Mutational burden per sample

```{r}
#| label: mutations_per_sample
#| echo: false

data_mutations_clean |> 
  count(SAMPLE_ID) |> 
  ggplot(mapping = aes(x = n)) +
  geom_density() +
  labs(x = "Mutations per sample", 
       y = "Density") + 
  theme_minimal()
```

Most frequently mutated genes:

```{r}
#| label: mutated_gene
#| echo: false

data_mutations_clean |> 
  count(Hugo_Symbol) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  ggplot(mapping = aes(x = reorder(Hugo_Symbol, n), 
                       y = n/nrow(data_clinical_sample)*100)) +
  geom_col() +
  labs(x = "Gene", 
       y = "Number of mutated samples (%)") + 
  theme_minimal() + 
  coord_flip()
```

### Structural variants

```{r}
#| label: clean_data_sv
#| echo: false

data_sv_clean <- data_sv |> 
  select(SAMPLE_ID = Sample_Id,
         Site1_Hugo_Symbol,
         Site2_Hugo_Symbol,
         Class) |> 
  filter(SAMPLE_ID %in% patients_included$SAMPLE_ID)

write_tsv(x = data_sv_clean,
          file = "../data/02_data_sv_clean.tsv")
```

Distribution of structural varaints in the data:

```{r}
#| label: distribution_sv
#| echo: false

data_sv_clean |> 
  count(Class) |> 
  ggplot(mapping = aes(x = Class, 
                        y = n, 
                        fill = Class)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.5) +
  labs(x = "SV class", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None")
```

Most frequently altered genes:

```{r}
#| label: sv_gene
#| echo: false

top20 <- data_sv_clean |> 
  group_by(Site1_Hugo_Symbol) |> 
  summarise(n = n()) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  pull(Site1_Hugo_Symbol)
  
data_sv_clean |> 
  filter(Site1_Hugo_Symbol %in% top20) |> 
  ggplot(mapping = aes(x = Site1_Hugo_Symbol,
                       fill = Class)) +
  geom_bar() +
  labs(x = "Gene", 
       y = "Number of structural variants in the data") + 
  theme_minimal() + 
  coord_flip() 
```


