---
title: "Define Progression-Free Survival"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
library("tidyverse")
library("survival")
library("survminer")
library("patchwork")

# Fix mistake in survminer
is_ggplot <- function(x)is.ggplot(x)
```

# Load data

```{r}
data_timeline_progression <- read_tsv("../data/01_data_timeline_progression.tsv")
data_timeline_cancer_presence <- read_tsv("../data/01_data_timeline_cancer_presence.tsv")
data_clinical_patient <- read_tsv("../data/01_data_clinical_patient.tsv")
data_timeline_diagnosis <- read_tsv("../data/01_data_timeline_diagnosis.tsv")
data_timeline_surgery <- read_tsv("../data/01_data_timeline_surgery.tsv")
data_timeline_treatment <- read_tsv("../data/01_data_timeline_treatment.tsv")
data_clinical_sample <- read_tsv("../data/01_data_clinical_sample.tsv")
```

# Define PFS

The [NCI Dictionary of Cancer Terms](https://www.cancer.gov/publications/dictionaries/cancer-terms/def/progression-free-survival "NCI Dictionary of Cancer Terms: Progression-free survival") defined PFS as follows:

> The length of time during and after the treatment of a disease, such as cancer, that a patient lives with the disease but it does not get worse. In a clinical trial, measuring the progression-free survival is one way to see how well a new treatment works. Also called PFS.

I define PFS based on a chosen time point 0 and a relevant PFS end-point.
In the timeline data, time point 0 is defined as the time of sequencing. This can be far after sample acquisition, diagnosis or surgery, so it is not necessarily a good time point 0 for the purposes of this project. I define time point 0 as the date of the first surgery, defined by data_timeline_surgery. Relevant PFS end-points include:
1. Progression, defined by data_timeline_progression;
2. Recurrence, defined by data_timeline_cancer_presence;
3. Death, defined by data_clinical_patient;
4. Censored or last follow-up, defined by data_clinical patient.

# Prepare PFS data

First, identify the CRC patients:

```{r}
CRC_patients <- data_clinical_sample |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()
```

Next, find the date of the first surgery:

```{r}
first_surgery_dates <- data_timeline_surgery |> 
  filter(SUBTYPE == "PROCEDURE",
         PATIENT_ID %in% CRC_patients) |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  select(PATIENT_ID,
         DATE_FIRST_SURGERY = START_DATE)
```

Then define the PFS time based on the different end points:

```{r}
# Progression
data_PFS <- first_surgery_dates |> 
  left_join(data_timeline_progression, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         PROGRESSION == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_PROGRESSION = min(START_DATE)) |> 
  right_join(first_surgery_dates, join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = !is.na(DATE_FIRST_PROGRESSION)) 

# Recurrence
data_PFS <- data_PFS |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  filter(HAS_CANCER == "N") |> 
  select(PATIENT_ID, DATE_FIRST_SURGERY, PROGRESSION, DATE_FIRST_PROGRESSION) |> 
  left_join(data_timeline_cancer_presence, join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_RECURRENCE = min(START_DATE)) |> 
  right_join(data_PFS, join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_FIRST_RECURRENCE)) 

# Death
data_PFS <- data_PFS |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, OS_MONTHS) |> 
  right_join(data_PFS, join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censor
data_PFS <- data_PFS |> 
  left_join(data_clinical_patient, join_by(PATIENT_ID)) |> 
  group_by(PATIENT_ID) |> 
  mutate(PFS_STATUS = ifelse(PROGRESSION == FALSE & RECURRENCE == FALSE & DEATH == FALSE, "0:No progression", "1:Progression event"),
         PFS_DATE = ifelse(PFS_STATUS == "1:Progression event", pmin(DATE_FIRST_PROGRESSION, DATE_FIRST_RECURRENCE, DATE_DEATH, na.rm = TRUE), round(OS_MONTHS*30.44)),
         DAYS_TO_EVENT = PFS_DATE - DATE_FIRST_SURGERY,
         YEARS_TO_EVENT = DAYS_TO_EVENT/365.25) |> 
  select(PATIENT_ID, PFS_STATUS, DAYS_TO_EVENT, YEARS_TO_EVENT, PFS_DATE, DATE_FIRST_SURGERY, PROGRESSION, DATE_FIRST_PROGRESSION, RECURRENCE, DATE_FIRST_RECURRENCE, DEATH, DATE_DEATH)

write_tsv(x = data_PFS,
          file = "../data/02_data_PFS.tsv")

data_PFS |> 
  group_by(PFS_STATUS) |> 
  summarise(n = n(), pct = n()/nrow(data_PFS))

data_PFS |> 
  ggplot(mapping = aes(x = YEARS_TO_EVENT,
                       fill = PFS_STATUS)) + 
  geom_histogram(color = "white") + 
  facet_wrap(~ PFS_STATUS, dir = "v") + 
  theme_minimal() + 
  theme(legend.position = "None")

ggsave(filename = "../results/02_years_to_progression_event.png")
```

# Kaplan Meier plots

```{r}
data_PFS_bin <- data_PFS |> 
  mutate(PFS_STATUS = ifelse(PFS_STATUS == "0:No progression", 0, 1))

fit_PFS <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ 1, data = data_PFS_bin)

ggsurvplot(
  fit = fit_PFS,
  data = data_PFS_bin,
  title = "KM plot for PFS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T
) 

ggsave(filename = "../results/02_KM.png")
```

## By stage

```{r}
data_PFS_bin <- data_PFS_bin |> 
  inner_join(data_clinical_patient, by = join_by(PATIENT_ID))

fit_PFS_by_stage <- survfit(Surv(YEARS_TO_EVENT, PFS_STATUS) ~ STAGE_HIGHEST_RECORDED, data = data_PFS_bin)

ggsurvplot(
  fit = fit_PFS_by_stage,
  data = data_PFS_bin,
  title = "Progression-free survival after CRC surgery",         
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 1,
  legend = "top",
  surv.median.line = "hv",
  pval = T,
) 

ggsave(filename = "../results/02_KM_by_stage.png")
```