---
title: "Define cohort"
author: "Puck Quarles van Ufford"
date: today
format: html
---

# Set up

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("rsample")
```

```{r}
#| label: seed
#| echo: true

set.seed(8)
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_clinical_sample <- read_tsv(file = "../data/01_data_clinical_sample.tsv")
data_timeline_surgery <- read_tsv(file = "../data/01_data_timeline_surgery.tsv")
data_timeline_sample_acquisition <- read_tsv(file = "../data/01_data_timeline_specimen_surgery.tsv")
data_timeline_cancer_presence <- read_tsv(file = "../data/01_data_timeline_cancer_presence.tsv")
data_timeline_treatment <- read_tsv(file = "../data/01_data_timeline_treatment.tsv")
data_timeline_radiation <- read_tsv(file = "../data/01_data_timeline_radiation.tsv")
data_clinical_patient <- read_tsv(file = "../data/01_data_clinical_patient.tsv")
```

# Define inclusion and exclusion criteria

In this project, we are focused on stage I-III colorectal cancer (CRC) patients who underwent surgical resection of the tumor and surrounding tissues, which is intended to be curative. To define which patients should be included in this cohort, we use the relevant data that was collected within 180 days of the date of the first surgery recorded in MSK-CHORD. The following criteria were used to define what patients should be included.

1. Patients with a CRC diagnosis were selected.

```{r}
#| label: crc_patients
#| echo: true

crc_patients <- data_clinical_sample |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()

length(crc_patients)
```

2. Patients who underwent surgery were selected and the date of their first surgery was identified.

```{r}
#| label: surgery_dates
#| echo: true

surgery_dates <- data_timeline_surgery |> 
  filter(SUBTYPE == "PROCEDURE",
         PATIENT_ID %in% crc_patients) |> 
  group_by(PATIENT_ID) |> 
  summarise(SURGERY_DATE = min(START_DATE))

nrow(surgery_dates)
```

3. Patients who had a sample taken within 180 days of the surgery were included.

```{r}
#| label: sample_dates
#| echo: true

patients_sampled <- data_timeline_sample_acquisition |> 
  right_join(surgery_dates,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 180)) |> 
  select(PATIENT_ID, 
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE = START_DATE)

n_distinct(patients_sampled$PATIENT_ID)
```

4. Patients whose sample indicated a metastasis were excluded.

```{r}
#| label: exclude_metastases
#| echo: true

patients_metastatic <- data_clinical_sample |> 
  right_join(patients_sampled,
             by = join_by(PATIENT_ID,
                          SAMPLE_ID)) |> 
  filter(SAMPLE_TYPE == "Metastasis") |> 
  pull(PATIENT_ID) |> 
  unique()

patients_included <- patients_sampled |> 
  filter(!PATIENT_ID %in% patients_metastatic)

n_distinct(patients_included$PATIENT_ID)
```

5. Patients who were diagnosed with cancers other than CRC were excluded.

```{r}
#| label: exclude_multiple_cancers
#| echo: true

patients_multiple_cancers <- patients_included |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  filter(n > 1) |> 
  left_join(data_clinical_sample,
            by = join_by(PATIENT_ID)) |> 
  filter(!CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()

patients_included <- patients_included |> 
  filter(!PATIENT_ID %in% patients_multiple_cancers)

n_distinct(patients_included$PATIENT_ID)
```

6. Patients whose sample indicates stage IV CRC were excluded, because their surgery would not be intended to be curative.

```{r}
#| label: exclude_stage_4
#| echo: true

patients_included <- data_clinical_sample |> 
  right_join(patients_included,
             by = join_by(PATIENT_ID, 
                          SAMPLE_ID)) |> 
  filter(CANCER_TYPE == "Colorectal Cancer",
         SAMPLE_TYPE == "Primary",
         !str_starts(PATHOLOGICAL_GROUP, 
                     "4")) |> 
  select(PATIENT_ID,
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE)

n_distinct(patients_included$PATIENT_ID)
```

7. If a patient had imaging data from the first 6 months after the surgery, patients whose most recent scan indicates the presence of cancer were excluded, because that indicates that the surgery was not curative.

```{r}
#| label: not_cured
#| echo: true

patients_not_cured <- patients_included |> 
  left_join(data_timeline_cancer_presence, 
            join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 0, 180)) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  filter(HAS_CANCER == "Y") |> 
  pull(PATIENT_ID)

patients_included <- patients_included |> 
  filter(!PATIENT_ID %in% patients_not_cured)

n_distinct(patients_included$PATIENT_ID)
```

In this project, we will investigate the disease-free survival (DFS) and overall survival (OS) from 180 days after the day of the first surgery. Therefor, we must define the relevant survival events, namely recurrence (which can be indicated by imaging or treatment) and death.

```{r}
#| label: define_survival_data
#| echo: true

# Recurrence (indicated by imaging)
data_survival <- patients_included |> 
  left_join(data_timeline_cancer_presence, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_RECURRENCE = min(START_DATE)) |> 
  right_join(patients_included, 
             join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_RECURRENCE)) 

# Recurrence (indicated by surgery)
data_survival <- data_survival |> 
  left_join(data_timeline_surgery, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         SUBTYPE == "PROCEDURE") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_NEXT_SURGERY = min(START_DATE)) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = ifelse(RECURRENCE,
                             RECURRENCE, 
                             !is.na(DATE_NEXT_SURGERY))) 

# Recurrence (indicated by treatment)
data_survival <- data_timeline_treatment |> 
  bind_rows(data_timeline_radiation) |> 
  right_join(data_survival,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE > SURGERY_DATE + 180,
         SUBTYPE %in% c("Chemo", 
                        "Targeted", 
                        "Radiation Therapy")) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_TREATMENT = min(START_DATE)) |> 
  left_join(data_timeline_treatment, 
            by = join_by(PATIENT_ID)) |> 
  filter(SUBTYPE != "Chemo" | 
           DATE_TREATMENT - STOP_DATE > 60) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_TREATMENT = min(DATE_TREATMENT)) |> 
  right_join(data_survival, 
             by = join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = ifelse(RECURRENCE,
                             RECURRENCE, 
                             !is.na(DATE_TREATMENT)))

# Death
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, 
         OS_MONTHS) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censoring
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID))|> 
  group_by(PATIENT_ID) |> 
  mutate(OS_STATUS = ifelse(DEATH == FALSE, 
                            0, 
                            1),
         DFS_STATUS = ifelse(RECURRENCE == FALSE & 
                               DEATH == FALSE, 
                             0, 
                             1),
         OS_DATE = round(OS_MONTHS*30.44),
         DFS_DATE = ifelse(DFS_STATUS == 1, 
                           pmin(DATE_RECURRENCE, 
                                DATE_DEATH, 
                                na.rm = TRUE), 
                           OS_DATE),
         DAYS_TO_OS_EVENT = OS_DATE - SURGERY_DATE,
         DAYS_TO_DFS_EVENT = DFS_DATE - SURGERY_DATE,
         YEARS_TO_OS_EVENT = DAYS_TO_OS_EVENT/365.25, 
         YEARS_TO_DFS_EVENT = DAYS_TO_DFS_EVENT/365.25) |> 
  select(PATIENT_ID, 
         SAMPLE_ID,
         SURGERY_DATE,
         SAMPLE_DATE,
         OS_STATUS, 
         DAYS_TO_OS_EVENT, 
         YEARS_TO_OS_EVENT, 
         DFS_STATUS, 
         DAYS_TO_DFS_EVENT, 
         YEARS_TO_DFS_EVENT)
```

8. Only patients survived without recurrence or death for 180 days after the surgery were included.

```{r}
#| label: exclude_no_DFS_180_days
#| echo: true

patients_included <- data_survival |> 
  mutate(survived_6_months = case_when(OS_STATUS == 1 ~ DAYS_TO_OS_EVENT > 180,
                                       TRUE ~ TRUE),
         no_recurrence_6_months = case_when(DFS_STATUS == 1 ~ DAYS_TO_DFS_EVENT > 180,
                                            TRUE ~ TRUE)) |> 
  filter(survived_6_months == TRUE,
         no_recurrence_6_months == TRUE) |> 
  select(- c(survived_6_months,
             no_recurrence_6_months))

n_distinct(patients_included$PATIENT_ID)
```

# Split data

The data will be split 80/20 into a training and testing set, controlling for their survival status.

```{r}
#| label: split_data
#| echo: true

split <- patients_included |> 
  mutate(strata = interaction(OS_STATUS,
                              DFS_STATUS)) |> 
  initial_split(prop = 0.8, 
                strata = strata)

data_train <- training(split) |> 
  select(-strata)

data_test <- testing(split) |> 
  select(-strata)

print(str_c("# patients in training set: ", 
            nrow(data_train)))

print(str_c("# patients in test set: ", 
            nrow(data_test)))

write_tsv(x = data_train,
          file = "../data/02_data_train.tsv")

write_tsv(x = data_test,
          file = "../data/02_data_test.tsv")
```
