---
title: "Prepare surival data"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: seed

set.seed(8)
```

```{r}
#| label: packages
#| message: false

library("tidyverse")
library("survival")
library("survminer")
library("rsample")
```

```{r}
#| label: fix_mistake_survminer_package
#| echo: false

is_ggplot <- function(x)is.ggplot(x)
```

# Load data

```{r}
#| label: load_data
#| message: false

data_clinical_sample <- read_tsv("../data/01_data_clinical_sample.tsv")
data_timeline_surgery <- read_tsv("../data/01_data_timeline_surgery.tsv")
data_timeline_diagnosis <- read_tsv("../data/01_data_timeline_diagnosis.tsv")
data_timeline_progression <- read_tsv("../data/01_data_timeline_progression.tsv")
data_timeline_treatment <- read_tsv("../data/01_data_timeline_treatment.tsv")
data_timeline_radiation <- read_tsv("../data/01_data_timeline_radiation.tsv")
data_timeline_cancer_presence <- read_tsv("../data/01_data_timeline_cancer_presence.tsv")
data_clinical_patient <- read_tsv("../data/01_data_clinical_patient.tsv")
```

# Prepare survival data

For this project, I am selecting colorectal cancer patients who had stage I-III CRC at the time of surgery, and whose surgery was considered curative.

First, I will select patients who had CRC:

```{r}
#| label: select_patients

CRC_patients <- data_clinical_sample |> 
  filter(CANCER_TYPE == "Colorectal Cancer") |> 
  pull(PATIENT_ID) |> 
  unique()
```

Then, I find the date of the first surgery:

```{r}
#| label: first_surgery_date

first_surgery_dates <- data_timeline_surgery |> 
  filter(SUBTYPE == "PROCEDURE",
         PATIENT_ID %in% CRC_patients) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_SURGERY = min(START_DATE))
```

Next, I need to confirm the cancer stage at the time of surgery:

```{r}
#| label: cancer_stage

CRC_patients <- data_timeline_diagnosis |> 
  inner_join(first_surgery_dates, 
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY,
         SUBTYPE == "Primary") |> 
  group_by(PATIENT_ID) |> 
  arrange(START_DATE) |> 
  slice(1) |> 
  filter(STAGE_CDM_DERIVED == "Stage 1-3") |> 
  pull(PATIENT_ID) |> 
  unique()

length(CRC_patients)
```

Finally, I define the survival data based on the different end points:

```{r}
#| label: survival_data

# Progression
data_survival <- first_surgery_dates |> 
  left_join(data_timeline_progression, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         PROGRESSION == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_PROGRESSION = min(START_DATE)) |> 
  right_join(first_surgery_dates, 
             join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = !is.na(DATE_FIRST_PROGRESSION)) 

# Progression indicated by treatment
# If chemotherapy, targeted treatment, or radiation started 180 days after surgery and 60 days after the last chemotherapy treatment ended, it indicates that the tumor has progressed after surgery.
data_survival <- data_survival |> 
  left_join(data_timeline_treatment,
            by = join_by(PATIENT_ID)) |> 
  bind_rows(data_timeline_radiation) |> 
  filter(START_DATE > DATE_FIRST_SURGERY + 180,
         SUBTYPE %in% c("Chemo", 
                        "Targeted", 
                        "Radiation Therapy")) |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_NEXT_TREATMENT = min(START_DATE)) |> 
  left_join(data_timeline_treatment, 
             by = join_by(PATIENT_ID)) |> 
  filter(DATE_NEXT_TREATMENT - STOP_DATE > 60,
         SUBTYPE == "Chemo") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_NEXT_TREATMENT = min(DATE_NEXT_TREATMENT)) |> 
  right_join(data_survival, 
             by = join_by(PATIENT_ID)) |> 
  mutate(PROGRESSION = ifelse(PROGRESSION, 
                              PROGRESSION, 
                              !is.na(DATE_NEXT_TREATMENT)))

# Recurrence
data_survival <- data_survival |> 
  left_join(data_timeline_cancer_presence, 
            join_by(PATIENT_ID)) |> 
  filter(START_DATE > DATE_FIRST_SURGERY,
         HAS_CANCER == "Y") |> 
  group_by(PATIENT_ID) |> 
  summarise(DATE_FIRST_RECURRENCE = min(START_DATE)) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(RECURRENCE = !is.na(DATE_FIRST_RECURRENCE)) 

# Death
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID)) |> 
  filter(OS_STATUS == "1:DECEASED") |> 
  group_by(PATIENT_ID) |> 
  slice(1) |> 
  select(PATIENT_ID, 
         OS_MONTHS) |> 
  right_join(data_survival, 
             join_by(PATIENT_ID)) |> 
  mutate(DEATH = !is.na(OS_MONTHS),
         DATE_DEATH = round(OS_MONTHS*30.44)) |> 
  select(-OS_MONTHS)

# Censor
data_survival <- data_survival |> 
  left_join(data_clinical_patient, 
            join_by(PATIENT_ID))|> 
  filter(PATIENT_ID %in% CRC_patients) |> 
  group_by(PATIENT_ID) |> 
  mutate(OS_STATUS = ifelse(DEATH == FALSE, 
                            0, 
                            1),
         PFS_STATUS = ifelse(PROGRESSION == FALSE & 
                               RECURRENCE == FALSE & 
                               DEATH == FALSE, 
                             0, 
                             1),
         DFS_STATUS = ifelse(RECURRENCE == FALSE & 
                               DEATH == FALSE, 
                             0, 
                             1),
         OS_DATE = round(OS_MONTHS*30.44),
         PFS_DATE = ifelse(PFS_STATUS == 1, 
                           pmin(DATE_FIRST_PROGRESSION, 
                                DATE_NEXT_TREATMENT,
                                DATE_FIRST_RECURRENCE, 
                                DATE_DEATH, 
                                na.rm = TRUE), 
                           round(OS_MONTHS*30.44)),
         DFS_DATE = ifelse(DFS_STATUS == 1, 
                           pmin(DATE_FIRST_RECURRENCE, 
                                DATE_DEATH, 
                                na.rm = TRUE), 
                           round(OS_MONTHS*30.44)),
         DAYS_TO_OS_EVENT = OS_DATE - DATE_FIRST_SURGERY,
         DAYS_TO_PFS_EVENT = PFS_DATE - DATE_FIRST_SURGERY,
         DAYS_TO_DFS_EVENT = DFS_DATE - DATE_FIRST_SURGERY,
         YEARS_TO_OS_EVENT = DAYS_TO_OS_EVENT/365.25, 
         YEARS_TO_PFS_EVENT = DAYS_TO_PFS_EVENT/365.25, 
         YEARS_TO_DFS_EVENT = DAYS_TO_DFS_EVENT/365.25) |> 
  select(PATIENT_ID, 
         OS_STATUS, 
         DAYS_TO_OS_EVENT, 
         YEARS_TO_OS_EVENT, 
         PFS_STATUS, 
         DAYS_TO_PFS_EVENT, 
         YEARS_TO_PFS_EVENT, 
         DFS_STATUS, 
         DAYS_TO_DFS_EVENT, 
         YEARS_TO_DFS_EVENT, 
         DATE_FIRST_SURGERY)

head(data_survival)

write_tsv(x = data_survival,
          file = "../data/02_data_survival.tsv")
```

# Summarize survival data

```{r}
#| label: summarize_survival data

data_survival |> 
  group_by(OS_STATUS) |> 
  summarise(n(), 
            mean(YEARS_TO_OS_EVENT))

data_survival |> 
  group_by(PFS_STATUS) |> 
  summarise(n(), 
            mean(YEARS_TO_PFS_EVENT))

data_survival |> 
  group_by(DFS_STATUS) |> 
  summarise(n(), 
            mean(YEARS_TO_DFS_EVENT))
```

```{r}
#| label: km_curves
#| message: false

fit_OS <- survfit(Surv(YEARS_TO_OS_EVENT, 
                       OS_STATUS) ~ 1, 
                  data = data_survival)

ggsurvplot(
  fit = fit_OS,
  data = data_survival,
  title = "KM plot for OS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T,
  surv.median.line = "hv"
) 

ggsave("../results/02_KM_OS.png")

fit_PFS <- survfit(Surv(YEARS_TO_PFS_EVENT, 
                        PFS_STATUS) ~ 1, 
                   data = data_survival)

ggsurvplot(
  fit = fit_PFS,
  data = data_survival,
  title = "KM plot for PFS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T,
  surv.median.line = "hv"
) 

ggsave("../results/02_KM_PFS.png")

fit_DFS <- survfit(Surv(YEARS_TO_DFS_EVENT, 
                        DFS_STATUS) ~ 1, 
                   data = data_survival)

ggsurvplot(
  fit = fit_DFS,
  data = data_survival,
  title = "KM plot for DFS after first CRC surgery",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "none",
  risk.table = T,
  surv.median.line = "hv"
) 

ggsave("../results/02_KM_DFS.png")
```

# Split data

I will reserve 20% of the data for model testing. I use stratified splitting to ensure that both the training and testing set are balanced in terms of survival outcomes.

```{r}
#| label: split_data

split <- data_survival |> 
  mutate(strata = interaction(OS_STATUS, 
                              PFS_STATUS,
                              DFS_STATUS)) |> 
  initial_split(prop = 0.8, 
                strata = strata)

data_train <- training(split) |> 
  select(-strata)

data_test <- testing(split) |> 
  select(-strata)

write_tsv(x = data_train,
          file = "../data/02_data_train.tsv")

write_tsv(x = data_test,
          file = "../data/02_data_test.tsv")

nrow(data_train)
nrow(data_test)
```

Evaluate if the training and test set are resembling each other:

```{r}
#| label: evaluate_split
#| warning: false
#| message: false

data_train <- data_train |> 
  mutate(split = "Train")

data_test <- data_test |> 
  mutate(split = "Test") 

data_split <- bind_rows(data_train, 
                        data_test)

fit_OS_split <- survfit(Surv(YEARS_TO_OS_EVENT, 
                             OS_STATUS) ~ split, 
                        data = data_split)

ggsurvplot(
  fit = fit_OS_split,
  data = data_split,
  title = "KM plot for OS by data split",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "top",
  legend.labs = c("test", "train"),
  surv.median.line = "hv",
  pval = TRUE,
  pval.method = TRUE
) 

ggsave("../results/02_KM_OS_split.png")

fit_PFS_split <- survfit(Surv(YEARS_TO_PFS_EVENT, 
                              PFS_STATUS) ~ split, 
                         data = data_split)

ggsurvplot(
  fit = fit_PFS_split,
  data = data_split,
  title = "KM plot for PFS by data split",
  xlab = "Time (years)",
  ylab = "PFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "top",
  legend.labs = c("test", "train"),
  surv.median.line = "hv",
  pval = TRUE,
  pval.method = TRUE
) 

ggsave("../results/02_KM_PFS_split.png")

fit_DFS_split <- survfit(Surv(YEARS_TO_DFS_EVENT, 
                              DFS_STATUS) ~ split, 
                         data = data_split)

ggsurvplot(
  fit = fit_DFS_split,
  data = data_split,
  title = "KM plot for DFS by data split",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "top",
  legend.labs = c("test", "train"),
  surv.median.line = "hv",
  pval = TRUE,
  pval.method = TRUE
) 

ggsave("../results/02_KM_DFS_split.png")
```
