---
title: "Clean data"
author: "Puck Quarles van Ufford"
date: today
format: html
---

# Set up

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("janitor")
library("fastDummies")
library("survival")
library("caret")
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train <- read_tsv(file = "../data/02_data_train.tsv")
data_clinical_sample <- read_tsv(file = "../data/01_data_clinical_sample.tsv")
data_clinical_patient <- read_tsv(file = "../data/01_data_clinical_patient.tsv")
data_timeline_performance_status <- read_tsv(file = "../data/01_data_timeline_performance_status.tsv")
data_timeline_cea_labs <- read_tsv(file = "../data/01_data_timeline_cea_labs.tsv")
data_timeline_tumor_sites <- read_tsv(file = "../data/01_data_timeline_tumor_sites.tsv")
data_timeline_mmr <- read_tsv(file = "../data/01_data_timeline_mmr.tsv")
data_timeline_treatment <- read_tsv(file = "../data/01_data_timeline_treatment.tsv")
data_timeline_radiation <- read_tsv(file = "../data/01_data_timeline_radiation.tsv")
data_timeline_prior_meds <- read_tsv(file = "../data/01_data_timeline_prior_meds.tsv")
data_cna <- read_tsv(file = "../data/01_data_cna.tsv")
data_mutations <- read_tsv(file = "../data/01_data_mutations.tsv")
data_sv <- read_tsv(file = "../data/01_data_sv.tsv")
```

# Clean individual datasets

## Clinical sample

Select training data, select relevant features, and simplify clinical and pathological groups.

```{r}
#| label: clean_data_clinical_sample
#| message: false
#| echo: true

data_clinical_sample_clean <- data_clinical_sample |> 
  right_join(data_train,
             by = join_by(PATIENT_ID, 
                          SAMPLE_ID)) |> 
  remove_constant() |> 
  # CANCER_TYPE_DETAILED was omitted, because it contains the same information as ONCOTREE_CODE.
  # Columns with descriptions and comments for extra detail were removed.
  # GLEASON_SAMPLE_LEVEL and PDL1_POSITIVE were omitted, because they were \>90% homogeneous.
  select(- c(CANCER_TYPE_DETAILED,
             MSI_COMMENT,
             CLINICAL_SUMMARY,
             ICD_O_HISTOLOGY_DESCRIPTION,
             DIAGNOSIS_DESCRIPTION,
             GLEASON_SAMPLE_LEVEL,
             PDL1_POSITIVE)) |>
  mutate(PRIMARY_SITE = str_to_lower(PRIMARY_SITE),
         MSI_TYPE = ifelse(MSI_TYPE == "Do not report", 
                           NA, 
                           MSI_TYPE),
         CLINICAL_GROUP_SIMPLIFIED = as.integer(str_sub(CLINICAL_GROUP,
                                                        1, 1)),
         PATHOLOGICAL_GROUP_SIMPLIFIED = as.integer(str_sub(PATHOLOGICAL_GROUP,
                                                            1, 1))) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## Clinical patient

Select training data, select relevant features, simplify race and ethnicity, and calculate age at surgery.

```{r}
#| label: clean_data_clinical_patient
#| message: false
#| echo: true

data_clinical_patient_clean <- data_clinical_patient|> 
  right_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  remove_constant() |> 
  # The NLP-derived tumor site data will be taken from data_timeline_tumor_sites to take the timing into account.
  # GLEASON_FIRST_REPORTED, GLEASON_HIGHEST_REPORTED, HISTORY_OF_PDL1, HR and HER2 were omitted, because they were \>90% homogeneous.
  select(- c(ADRENAL_GLANDS,
             BONE,
             CNS_BRAIN,
             INTRA_ABDOMINAL,
             LIVER,
             LUNG,
             LYMPH_NODES,
             OTHER,
             PLEURA,
             REPRODUCTIVE_ORGANS,
             GLEASON_FIRST_REPORTED, 
             GLEASON_HIGHEST_REPORTED, 
             HISTORY_OF_PDL1, 
             HR,
             HER2,
             OS_STATUS.x,
             OS_STATUS.y)) |> 
  mutate(RACE_SIMPLIFIED = case_when(RACE == "White" ~ "White",
                                     RACE == "Black Or African American" ~ "Black",
                                     RACE == "Asian-Far East/Indian Subcont" ~ "Asian",
                                     RACE == "Unknown" | 
                                       RACE == "Pt Refused to Answer" ~ NA,
                                     TRUE ~ "Other"),
         ETHNICITY_SIMPLIFIED = case_when(ETHNICITY == "Non-Spanish; Non-Hispanic" ~ "Non-Spanish; Non-Hispanic",
                                          ETHNICITY == "Unknown whether Spanish or not" ~ NA,
                                          TRUE ~ "Spanish/Hispanic"),
         EARLY_ONSET = ifelse(CURRENT_AGE_DEID < 50,
                              TRUE,
                              FALSE),
         LAST_FOLLOW_UP_DATE = round(OS_MONTHS*30.44),
         AGE_SURGERY = CURRENT_AGE_DEID - round((LAST_FOLLOW_UP_DATE - SURGERY_DATE)/365.25)) |> 
  rename(AGE_LAST_FOLLOW_UP = CURRENT_AGE_DEID,
         SMOKING_STATUS_NLP = SMOKING_PREDICTIONS_3_CLASSES,
         PRIOR_MED_NLP = PRIOR_MED_TO_MSK) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## Performance status

Select training data, filter dates, and select relevant columns.

```{r}
#| label: clean_data_timeline_performance_status
#| message: false
#| echo: true

data_timeline_performance_status_clean <- data_timeline_performance_status |> 
  right_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE,
             STYLE_COLOR)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## CEA labs

Select training data, filter dates, and select relevant columns.

```{r}
#| label: clean_data_timeline_cea_labs
#| message: false
#| echo: true

data_timeline_cea_labs_clean <- data_timeline_cea_labs |> 
  right_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             LR_UNIT_MEASURE,
             TEST)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## Tumor sites

Select training data, filter dates, and select relevant columns.

```{r}
#| label: clean_data_timeline_tumor_sites
#| message: false
#| echo: true

data_timeline_tumor_sites_clean <- data_timeline_tumor_sites |> 
  right_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE,
             SOURCE)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## MMR

Select training data, filter dates, and select relevant columns.

```{r}
#| label: clean_data_timeline_mmr
#| message: false
#| echo: true

data_timeline_mmr_clean <- data_timeline_mmr |> 
  right_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(between(START_DATE - SURGERY_DATE, 
                 -180, 180)) |> 
  select(- c(STOP_DATE,
             EVENT_TYPE,
             SUBTYPE)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## Treatment, radiation and prior meds

For treatment data, radiation data and prior meds data, select training data, filter dates, and select relevant columns. Then combine them into treatment history data.

```{r}
#| label: clean_data_treatment_history
#| message: false
#| echo: true

data_timeline_treatment_clean <- data_timeline_treatment |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE,
         AGENT) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - STOP_DATE)

data_timeline_radiation_clean <- data_timeline_radiation |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - START_DATE, 
         AGENT = NA)

data_timeline_prior_meds_clean <- data_timeline_prior_meds |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= SURGERY_DATE + 180) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         SURGERY_DATE,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = SURGERY_DATE - START_DATE,
         AGENT = NA)

data_treament_history_clean <- rbind(data_timeline_treatment_clean,
                                     data_timeline_radiation_clean,
                                     data_timeline_prior_meds_clean) 
```

## CNA

Pivot data to create a sample ID column, select training data, and select relevant columns.

```{r}
#| label: clean_data_cna
#| echo: true

data_cna_clean <- data_cna |> 
  pivot_longer(-Hugo_Symbol, 
               names_to = "SAMPLE_ID",
               values_to = "CNA") |> 
  right_join(data_train,
             by = join_by(SAMPLE_ID)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## Mutations

Select relevant columns, select training data, and select one row per sample-gene combination.

```{r}
#| label: clean_data_mutations
#| echo: true

data_mutations_clean <- data_mutations |> 
  select(SAMPLE_ID = Tumor_Sample_Barcode,
         Hugo_Symbol,
         Variant_Classification,
         Variant_Type) |> 
  right_join(data_train,
             by = join_by(SAMPLE_ID)) |> 
  group_by(SAMPLE_ID, Hugo_Symbol) |> 
  slice(1) |> 
  ungroup() |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

## SV

Select relevant columns and select training data.

```{r}
#| label: clean_data_sv
#| echo: true

data_sv_clean <- data_sv |> 
  select(SAMPLE_ID = Sample_Id,
         Site1_Hugo_Symbol,
         Site2_Hugo_Symbol,
         Class) |> 
  inner_join(data_train,
             by = join_by(SAMPLE_ID)) |> 
  select(- any_of(setdiff(colnames(data_train), "PATIENT_ID")))
```

# Combine data

The combined datasets contains one row per patient and the following columns:

- The columns from the clinical sample and patient data.
- Columns containing the latest ECOG score, CEA level and MMR assessment in the +- 180 day window around the surgery.
- A summary of all tumor site data in the 180 day window, with one column for each tumor site in the data, with a 1 if it was detected in a patient at any point in that window, and a 0 if not.
- A summary of the treatment history, with a column per treatment type and a 1 if the patient has a history of that treatment or a 0 if not.
- A summary of the CNA, mutations and SV data, with one column per gene. The columns indicate the type of genomic data with the prefixes CNA_, Mut_ and SV_, respectively.

```{r}
#| label: summarize_data_per_patient
#| echo: true

data_timeline_tumor_sites_summary <- data_timeline_tumor_sites_clean |> 
  group_by(PATIENT_ID) |> 
  summarize(
    CHEST   = as.integer(any(CHEST == 1)),
    ABDOMEN = as.integer(any(ABDOMEN == 1)),
    PELVIS  = as.integer(any(PELVIS == 1)),
    HEAD    = as.integer(any(HEAD == 1)),
    OTHER   = as.integer(any(OTHER == 1)),
    .groups = "drop"
  )  |> 
  left_join(data_timeline_tumor_sites_clean |> 
              distinct(PATIENT_ID, TUMOR_SITE) |> 
              mutate(has_site = 1) |> 
              pivot_wider(names_from = TUMOR_SITE,
                          values_from = has_site,
                          values_fill = 0),
            by = join_by(PATIENT_ID)) |> 
  mutate(other = ifelse(Other == 1 | OTHER == 1,
                        1, 0)) |> 
  select(- c(Other, OTHER))

data_treament_history_summary <- data_treament_history_clean |> 
  select(- c(AGENT,
             SURGERY_DATE,
             START_DATE,
             STOP_DATE,
             DAYS_BEFORE_SURGERY)) |> 
  distinct(PATIENT_ID, TREATMENT, .keep_all = TRUE) |> 
  mutate(has_treatment = 1) |> 
  pivot_wider(id_cols = c(PATIENT_ID, everything()),
              names_from = TREATMENT,
              values_from = has_treatment,
              values_fill = 0) 

data_cna_summary <- data_cna_clean |> 
  pivot_wider(names_from = Hugo_Symbol,
              values_from = CNA,
              names_prefix = "CNA_")

data_mutations_summary <- data_mutations_clean |> 
  select(- Variant_Type) |> 
  pivot_wider(names_from = Hugo_Symbol,
              values_from = Variant_Classification,
              names_prefix = "Mut_")

data_sv_summary <- data_sv_clean |> 
  select(- Site2_Hugo_Symbol) |> 
  pivot_wider(names_from = Site1_Hugo_Symbol,
              values_from = Class,
              names_prefix = "SV_")
```

```{r}
#| label: combine_data
#| echo: true

data_train_complete <- data_train |> 
  left_join(data_clinical_sample_clean,
            by = join_by(PATIENT_ID)) |> 
  left_join(data_clinical_patient_clean,
            by = join_by(PATIENT_ID)) |> 
  select(-OS_MONTHS) |> 
  left_join(data_timeline_performance_status_clean,
            by = join_by(PATIENT_ID)) |> 
  # Take latest ECOG in the +-180 day window
  rename(DATE_ECOG = START_DATE) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(DATE_ECOG)) |> 
  slice(1) |> 
  ungroup() |> 
  left_join(data_timeline_cea_labs_clean,
            by = join_by(PATIENT_ID)) |>
  # Take latest CEA in the +-180 day window
  rename(DATE_CEA = START_DATE,
         CEA = RESULT) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(DATE_CEA)) |> 
  slice(1) |> 
  ungroup() |> 
  left_join(data_timeline_tumor_sites_summary,
            by = join_by(PATIENT_ID)) |> 
  left_join(data_timeline_mmr_clean,
            by = join_by(PATIENT_ID)) |> 
  # Take latest MMR in the +-180 day window
  rename(DATE_MMR = START_DATE) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(DATE_MMR)) |> 
  slice(1) |> 
  ungroup() |> 
  left_join(data_treament_history_summary,
            by = join_by(PATIENT_ID)) |> 
  left_join(data_cna_summary,
            by = join_by(PATIENT_ID)) |> 
  left_join(data_mutations_summary,
            by = join_by(PATIENT_ID)) |> 
  left_join(data_sv_summary,
            by = join_by(PATIENT_ID)) 

data_train_complete <- data_train_complete |> 
  remove_empty(which = "cols") |> 
  remove_constant() |> 
  clean_names()
```

# Encode data

In this dataset, we need to encode features differently, based on whether they are numerical, nominal or ordinal. We also need to identify missing data and impute and encode that properly. Finally, certain columns need to be removed from the dataset.

```{r}
#| label: sort_features
#| echo: true

cna_cols <- data_train_complete |> 
  select(starts_with("cna_")) |>
  colnames()

mut_cols <- data_train_complete |> 
  select(starts_with("mut_")) |>
  colnames()

sv_cols <- data_train_complete |> 
  select(starts_with("sv_")) |>
  colnames()

cols_to_exclude <- c("surgery_date",
                     "sample_date",
                     "last_follow_up_date",
                     "date_ecog",
                     "date_cea",
                     "date_mmr",
                     "days_to_os_event",
                     "days_to_dfs_event",
                     "clinical_group",
                     "pathological_group",
                     "race",
                     "ethnicity",
                     "age_last_follow_up",
                     "stage_highest_recorded")

numeric_cols <- c("tmb_nonsynonymous",
                  "sample_coverage",
                  "num_icdo_dx",
                  "age_surgery")

id_cols <- c("patient_id",
             "sample_id")

outcome_cols <- c("os_status",
                  "dfs_status",
                  "years_to_os_event",
                  "years_to_dfs_event")

numeric_cols_with_NA <- c("tumor_purity",
                          "msi_score",
                          "cea")

ordinal_cols <- c(cna_cols,
                  "pathological_group_simplified")

ordinal_cols_with_NA <- c("clinical_group_simplified",
                          "ecog")

nominal_cols <- c("msi_type",
                  "race_simplified",
                  "ethnicity_simplified",
                  "smoking_status_nlp",
                  "prior_med_nlp",
                  "primary_site",
                  "gene_panel",
                  "oncotree_code",
                  "chest",
                  "abdomen",
                  "pelvis",
                  "head",
                  "other",
                  "lymph_nodes",
                  "liver",
                  "bone",
                  "lung",
                  "reproductive_organs",
                  "intra_abdominal",
                  "cns_brain",
                  "mmr_absent",
                  "chemo",
                  "investigational",
                  "hormone",
                  "biologic",
                  "bone_treatment",
                  "targeted",
                  "radiation_therapy",
                  "prior_medications_to_msk")

binary_cols <- c("gender",
                 "early_onset")

# Check if all is accounted for
setdiff(colnames(data_train_complete),
        c(id_cols, 
          outcome_cols, 
          mut_cols, 
          sv_cols,
          cols_to_exclude, 
          numeric_cols, 
          numeric_cols_with_NA, 
          ordinal_cols, 
          ordinal_cols_with_NA, 
          nominal_cols, 
          binary_cols))
```

First, excluded features were removed.

```{r}
#| label: remove_excluded_features
#| echo: true

data_train_complete <- data_train_complete |> 
  select(-all_of(cols_to_exclude)) 
```

Next, for the numerical and ordinal features with missing data, the missing data was imputed with the median and mode, respectively. Additional columns were created to indicate that the original value was NA.

```{r}
#| label: impute_missing_data
#| echo: true

data_train_complete <- data_train_complete |> 
  mutate(
    across(all_of(c(numeric_cols_with_NA,
                    ordinal_cols_with_NA)),
           ~ as.integer(is.na(.)),
           .names = "{.col}_na"),
    across(all_of(numeric_cols_with_NA),
           ~ if_else(is.na(.),
                     median(., na.rm = TRUE),
                     .)),
    across(all_of(ordinal_cols_with_NA),
           ~ {
             this_mode <- names(sort(table(.),
                                     decreasing = TRUE))[1]
             if_else(is.na(.),
                     as.numeric(this_mode),
                     .)}))
```

Nominal and binary features were one-hot encoded and if they contained missing data, it would be treated as its own category.

```{r}
#| label: encode_binary_and_nominal_data
#| echo: true

data_train_complete <- data_train_complete |> 
  mutate(
    across(starts_with("mut_"), 
           ~ if_else(is.na(.x), 0, 1)),
    across(starts_with("sv_"), 
           ~ if_else(is.na(.x), 0, 1)),
    gender = if_else(gender == "Male", 0, 1),
    early_onset = if_else(early_onset, 1, 0)
    ) |> 
  dummy_cols(select_columns = nominal_cols,
             remove_selected_columns = TRUE) |> 
  mutate(across(starts_with(nominal_cols),
                ~ replace_na(., 0)))
```

Finally, ordinal data is turned into a numerical representation, but to not force equal spacing, optimal scaling is used to maximize the likelihood function for the Cox proportional hazards model. This means that the scaling is not equal for estimating OS or DFS.

```{r}
#| label: encode_ordinal_data
#| warning: false
#| echo: true

get_log_likelihood <- function(levels, 
                               data, 
                               var, 
                               time_col, 
                               event_col) {
  
  # Sort the original values for the ordinal levels
  levels_vec <- sort(unique(data[[var]]))
  
  # Create a new variable with the new values for the ordinal levels
  data[[paste0(var, "_scaled")]] <- levels[match(data[[var]], 
                                                 levels_vec) ]
  
  # Fit a univariate Cox model and return the log likelihood
  surv_formula <- as.formula(
    paste0("Surv(", time_col, ", ", event_col, ") ~ ", var, "_scaled")
  )

  fit <- coxph(formula = surv_formula,
               data = data)
  
  return(logLik(fit))
}

optimimal_scaling <- function(data, 
                              var, 
                              time_col, 
                              event_col) {
  
  # Sort the original values for the ordinal levels and initiate the optimal scaling with 1, 2, 3, 4, ...
  levels_vec <- sort(unique(data[[var]]))
  
  n_levels <- length(levels_vec)

  init_levels <- seq_len(n_levels)
  
  # Optimize the scaling using the log likelihood of the Cox model
  opt <- optim(
    par = init_levels,
    fn = function(par) -get_log_likelihood(par, 
                                           data, 
                                           var, 
                                           time_col, 
                                           event_col),
    method = "L-BFGS-B",
    lower = rep(0, n_levels),
    control = list(fnscale = 1)
  )

  data.frame(value = levels_vec,
             optimal_score = opt$par)
}

rescale_ordinal_vars <- function(data, 
                                 ordinal_vars, 
                                 time_col, 
                                 event_col) {

  mappings <- list()

  for (var in ordinal_vars) {
    
    # Rescale the values assigned to the levels for each ordinal variable
    mapping <- optimimal_scaling(data, 
                                 var, 
                                 time_col, 
                                 event_col)
    
    mappings[[var]] <- mapping

    data[[paste0(var, "_scaled")]] <-
      mapping$optimal_score[ match(data[[var]], 
                                   mapping$value) ]
  }

  return(list(data = data,
              mappings = mappings))
}

data_rescaled_os <- rescale_ordinal_vars(data = data_train_complete,
                                         ordinal_vars = c(ordinal_cols, 
                                                          ordinal_cols_with_NA),
                                         time_col = "years_to_os_event",
                                         event_col = "os_status")

saveRDS(object = data_rescaled_os$mappings,
        file = "../results/03_map_ordinal_vars_OS.rds")

data_train_os <- data_rescaled_os$data |> 
  select(-c(ordinal_cols, 
            ordinal_cols_with_NA)) |> 
  clean_names()

data_rescaled_dfs <- rescale_ordinal_vars(data = data_train_complete,
                                          ordinal_vars = c(ordinal_cols, 
                                                           ordinal_cols_with_NA),
                                          time_col = "years_to_dfs_event",
                                          event_col = "dfs_status")

saveRDS(object = data_rescaled_dfs$mappings,
        file = "../results/03_map_ordinal_vars_DFS.rds")

data_train_dfs <- data_rescaled_dfs$data |> 
  select(-c(ordinal_cols, 
            ordinal_cols_with_NA)) |> 
  clean_names()
```

# Filter data

Remove features that are linear combinations of other features.

```{r}
#| label: lc_filter
#| echo: true

# OS
feature_data_os <- data_train_os |> 
    select(- all_of(c(outcome_cols, 
                      id_cols)))

mm <- model.matrix(~ . -1, 
                   data = feature_data_os)

assign_vec <- attr(mm, 
                   "assign")

term_labels <- attr(terms(~ . -1, 
                          data = feature_data_os), 
                    "term.labels")

lc <- findLinearCombos(mm)

cols_to_remove <- unique(term_labels[assign_vec[lc$remove]])

data_train_os <- data_train_os |> 
  select(- all_of(cols_to_remove))

# DFS
feature_data_dfs <- data_train_dfs |> 
    select(- all_of(c(outcome_cols, 
                      id_cols)))

mm <- model.matrix(~ . -1, 
                   data = feature_data_dfs)

assign_vec <- attr(mm, 
                   "assign")

term_labels <- attr(terms(~ . -1, 
                          data = feature_data_dfs), 
                    "term.labels")

lc <- findLinearCombos(mm)

cols_to_remove <- unique(term_labels[assign_vec[lc$remove]])

data_train_dfs <- data_train_dfs |> 
  select(- all_of(cols_to_remove))
```

For the categorical features, we need to remove the features that have a category with zero events or zero controls.

```{r}
#| label: exclude_no_level_variation
#| echo: true

# OS
features_os <- setdiff(colnames(data_train_os),
                       c(outcome_cols,
                         id_cols))

to_test_os <- setdiff(features_os,
                      c(numeric_cols,
                        numeric_cols_with_NA))

to_remove_os <- c()

for(var in to_test_os){
  test <- data_train_os |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

data_train_os <- data_train_os |> 
  select(- all_of(to_remove_os)) |>
  select(- c("dfs_status",
             "years_to_dfs_event"))

# DFS
features_dfs <- setdiff(colnames(data_train_dfs),
                        c(outcome_cols,
                          id_cols))

to_test_dfs <- setdiff(features_dfs,
                       c(numeric_cols,
                         numeric_cols_with_NA))

to_remove_dfs <- c()

for(var in to_test_dfs){
  test <- data_train_dfs |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

data_train_dfs <- data_train_dfs |> 
  select(- all_of(to_remove_dfs)) |>
  select(- c("os_status",
             "years_to_os_event"))
```

```{r}
#| label: save_data
#| echo: true

write_tsv(x = data_train_os,
          file = "../data/03_data_train_os.tsv")

write_tsv(x = data_train_dfs,
          file = "../data/03_data_train_dfs.tsv")
```

# Feature types

In the analyses, we want to distinguish between data that is routinely collected, NLP-derived, and from genomic testing. Thus, we create data subsets with the relevant features.

```{r}
#| label: define_feature_types
#| echo: true

all_colnames <- union(colnames(data_train_os),
                      colnames(data_train_os))

cna_features <- all_colnames[startsWith(all_colnames, "cna")] 

mut_features <- all_colnames[startsWith(all_colnames, "mut")] 

sv_features <- all_colnames[startsWith(all_colnames, "sv")] 

test <- all_colnames[startsWith(all_colnames, "cns")] 


radiology_prefixes <- c("chest",
                        "abdomen",
                        "pelvis",
                        "head",
                        "other",
                        "lymph_nodes",
                        "liver",
                        "bone",
                        "lung",
                        "reproductive_organs",
                        "intra_abdominal",
                        "cns",
                        "pleura")

radiology_features <- all_colnames[grepl(paste0("^(", 
                                                paste(radiology_prefixes, 
                                                      collapse = "|"), ")"), 
                                         all_colnames)]

treatment_prefixes <- c("biologic",
                        "chemo",
                        "hormone",
                        "investigational",
                        "prior",
                        "radiation",
                        "targeted")

treatment_features <- all_colnames[grepl(paste0("^(", 
                                                paste(treatment_prefixes, 
                                                      collapse = "|"), ")"), 
                                         all_colnames)]

nlp_features <- c("smoking_status_nlp_former_current_smoker",
                  "smoking_status_nlp_never",
                  "smoking_status_nlp_unknown",
                  "prior_med_nlp_no_prior_medications",
                  "prior_med_nlp_prior_medications_to_msk",
                  "prior_med_nlp_unknown",
                  "mmr_absent_false",
                  "mmr_absent_na",
                  "mmr_absent_true")

routine_features <- setdiff(all_colnames,
                            c(id_cols, 
                              outcome_cols, 
                              cna_features, 
                              mut_features, 
                              sv_features, 
                              nlp_features,
                              radiology_features)) 
```

```{r}
#| label: save_data_subsets
#| echo: true

data_train_os_noNLPgenomic <- data_train_os |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features)))

data_train_dfs_noNLPgenomic <- data_train_dfs |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features)))

write_tsv(x = data_train_os_noNLPgenomic,
          file = "../data/03_data_train_os_noNLPgenomic.tsv")

write_tsv(x = data_train_dfs_noNLPgenomic,
          file = "../data/03_data_train_dfs_noNLPgenomic.tsv")

data_train_os_noNLP <- data_train_os |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features,
                  cna_features,
                  mut_features,
                  sv_features)))

data_train_dfs_noNLP <- data_train_dfs |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features,
                  cna_features,
                  mut_features,
                  sv_features)))

write_tsv(x = data_train_os_noNLP,
          file = "../data/03_data_train_os_noNLP.tsv")

write_tsv(x = data_train_dfs_noNLP,
          file = "../data/03_data_train_dfs_noNLP.tsv")

data_train_os_nogenomic <- data_train_os |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features,
                  nlp_features,
                  radiology_features)))

data_train_dfs_nogenomic <- data_train_dfs |> 
  select(any_of(c(id_cols,
                  outcome_cols,
                  routine_features,
                  nlp_features,
                  radiology_features)))

write_tsv(x = data_train_os_nogenomic,
          file = "../data/03_data_train_os_nogenomic.tsv")

write_tsv(x = data_train_dfs_nogenomic,
          file = "../data/03_data_train_dfs_nogenomic.tsv")
```
