---
title: "Encode features"
author: "Puck Quarles van Ufford"
format: html
---

# Packages

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("janitor")
library("fastDummies")
library("survival")
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train <- read_tsv("../data/02_data_train_complete.tsv")
```

# Clean data

Empty and constant columns were removed from the data set.

```{r}
#| label: clean_data

data_train <- data_train |> 
  remove_empty(which = "cols") |> 
  remove_constant() |> 
  clean_names()
```

# Sort features

In this dataset, we need to encode features differently, based on whether they are numerical, nominal or ordinal. We also need to identify missing data and impute and encode that properly. Finally, certain columns need to be removed from the dataset.

```{r}
#| label: sort_features
#| echo: true

cols_to_exclude <- c("surgery_date",
                     "sample_date",
                     "last_follow_up_date",
                     "date_ecog",
                     "date_cea",
                     "date_mmr",
                     "days_to_os_event",
                     "days_to_dfs_event",
                     "clinical_group",
                     "pathological_group",
                     "race",
                     "ethnicity",
                     "age_last_follow_up",
                     "stage_highest_recorded")

numeric_cols <- c("tmb_nonsynonymous",
                  "sample_coverage",
                  "num_icdo_dx",
                  "age_surgery")

numeric_cols_with_NA <- c("tumor_purity",
                          "msi_score",
                          "cea")

cna_cols <- data_train |> 
  select(starts_with("cna_")) |>
  colnames()

ordinal_cols <- c(cna_cols,
                  "pathological_group_simplified")

ordinal_cols_with_NA <- c("clinical_group_simplified",
                          "ecog")

nominal_cols <- c("msi_type",
                  "race_simplified",
                  "ethnicity_simplified",
                  "smoking_status_nlp",
                  "prior_med_nlp",
                  "primary_site",
                  "gene_panel",
                  "oncotree_code",
                  "chest",
                  "abdomen",
                  "pelvis",
                  "head",
                  "other",
                  "lymph_nodes",
                  "other_2",
                  "liver",
                  "bone",
                  "lung",
                  "reproductive_organs",
                  "intra_abdominal",
                  "cns_brain",
                  "pleura",
                  "mmr_absent",
                  "chemo",
                  "investigational",
                  "hormone",
                  "biologic",
                  "bone_treatment",
                  "targeted",
                  "radiation_therapy",
                  "prior_medications_to_msk")
```

# Encode features

First, excluded features were removed.

```{r}
#| label: remove_excluded_features
#| echo: true

data_train <- data_train |> 
  select(-all_of(cols_to_exclude)) 
```

Next, for the numerical and ordinal features with missing data, the missing data was imputed with the median and mode, respectively. Additional columns were created to indicate that the original value was NA.

```{r}
#| label: impute_missing_data
#| echo: true

data_train <- data_train |> 
  mutate(
    across(all_of(c(numeric_cols_with_NA,
                    ordinal_cols_with_NA)),
           ~ as.integer(is.na(.)),
           .names = "{.col}_na"),
    across(all_of(numeric_cols_with_NA),
           ~ if_else(is.na(.),
                     median(., na.rm = TRUE),
                     .)),
    across(all_of(ordinal_cols_with_NA),
           ~ {
             this_mode <- names(sort(table(.),
                                     decreasing = TRUE))[1]
             if_else(is.na(.),
                     as.numeric(this_mode),
                     .)}))
```

Nominal and binary features were one-hot encoded and if they contained missing data, it would be treated as its own category.

```{r}
#| label: encode_binary_and_nominal_data
#| echo: true

data_train <- data_train |> 
  mutate(
    across(starts_with("mut_"), 
           ~ if_else(is.na(.x), 0, 1)),
    across(starts_with("sv_"), 
           ~ if_else(is.na(.x), 0, 1)),
    gender = if_else(gender == "Male", 0, 1),
    early_onset = if_else(early_onset, 1, 0)
    ) |> 
  dummy_cols(select_columns = nominal_cols,
             remove_selected_columns = TRUE) |> 
  mutate(across(starts_with(nominal_cols),
                ~ replace_na(., 0)))
```

Finally, ordinal data is turned into numerical, but to not force equal spacing, optimal scaling is used to maximize the likelihood function for the Cox proportional hazards model. This means that the scaling is not equal for estimating overall survival or disease free survival.

```{r}
#| label: encode_ordinal_data
#| warning: false
#| echo: true

get_log_likelihood <- function(scores, 
                               data, 
                               var, 
                               time_col, 
                               event_col) {
  
  levels_vec <- sort(unique(data[[var]]))
  
  data[[paste0(var, "_scaled")]] <- scores[match(data[[var]], 
                                                 levels_vec) ]
  
  surv_formula <- as.formula(
    paste0("Surv(", time_col, ", ", event_col, ") ~ ", var, "_scaled")
  )

  fit <- coxph(formula = surv_formula,
               data = data)
  
  return(logLik(fit))
}

optimimal_scaling <- function(data, 
                              var, 
                              time_col, 
                              event_col) {
  
  levels_vec <- sort(unique(data[[var]]))
  
  n_levels <- length(levels_vec)

  init_scores <- seq_len(n_levels)

  opt <- optim(
    par = init_scores,
    fn = function(par) -get_log_likelihood(par, 
                                           data, 
                                           var, 
                                           time_col, 
                                           event_col),
    method = "L-BFGS-B",
    lower = rep(0, 
                n_levels),
    control = list(fnscale = 1)
  )

  data.frame(
    value = levels_vec,
    optimal_score = opt$par
  )
}

rescale_ordinal_vars <- function(data, 
                                 ordinal_vars, 
                                 time_col, 
                                 event_col) {

  mappings <- list()

  for (var in ordinal_vars) {

    mapping <- optimimal_scaling(data, 
                                 var, 
                                 time_col, 
                                 event_col)
    
    mappings[[var]] <- mapping

    data[[paste0(var, "_scaled")]] <-
      mapping$optimal_score[ match(data[[var]], 
                                   mapping$value) ]
  }

  return(list(
    data = data,
    mappings = mappings
  ))
}

data_rescaled_os <- rescale_ordinal_vars(
  data = data_train,
  ordinal_vars = c(ordinal_cols, 
                   ordinal_cols_with_NA),
  time_col = "years_to_os_event",
  event_col = "os_status"
)

data_train_os <- data_rescaled_os$data |> 
  select(-c(ordinal_cols, 
            ordinal_cols_with_NA)) |> 
  clean_names()

data_rescaled_dfs <- rescale_ordinal_vars(
  data = data_train,
  ordinal_vars = c(ordinal_cols, 
                   ordinal_cols_with_NA),
  time_col = "years_to_dfs_event",
  event_col = "dfs_status"
)

data_train_dfs <- data_rescaled_dfs$data |> 
  select(-c(ordinal_cols, 
            ordinal_cols_with_NA)) |> 
  clean_names()
```

# Save data

```{r}
#| label: save_data

write_tsv(x = data_train_os,
          file = "../data/03_data_train_os.tsv")

write_tsv(x = data_train_dfs,
          file = "../data/03_data_train_dfs.tsv")
```
