---
title: "Prepare and explore explanatory variables"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false

library("tidyverse")
library("survival")
library("survminer")
library("table1")
library("patchwork")
```

```{r}
#| label: fix_mistake_survminer_package
#| echo: false

is_ggplot <- function(x)is.ggplot(x)
```

# Load data

```{r}
#| label: load_data
#| message: false

data_train <- read_tsv("../data/02_data_train.tsv")
data_clinical_patient <- read_tsv("../data/01_data_clinical_patient.tsv")
data_clinical_sample <- read_tsv("../data/01_data_clinical_sample.tsv")
data_timeline_sample_acquisition <- read_tsv("../data/01_data_timeline_specimen_surgery.tsv")
data_labs_ca_15_3 <- read_tsv("../data/01_data_timeline_ca_15_3_labs.tsv")
data_labs_ca_19_9 <- read_tsv("../data/01_data_timeline_ca_19_9_labs.tsv")
data_labs_cea  <- read_tsv("../data/01_data_timeline_cea_labs.tsv")
data_labs_psa  <- read_tsv("../data/01_data_timeline_psa_labs.tsv")
data_tumor_gleason  <- read_tsv("../data/01_data_timeline_gleason.tsv")
data_tumor_mmr  <- read_tsv("../data/01_data_timeline_mmr.tsv")
data_tumor_pdl1  <- read_tsv("../data/01_data_timeline_pdl1.tsv")
data_tumor_site  <- read_tsv("../data/01_data_timeline_tumor_sites.tsv")
data_performance_status  <- read_tsv("../data/01_data_timeline_performance_status.tsv")
data_treatment  <- read_tsv("../data/01_data_timeline_treatment.tsv")
data_radiation  <- read_tsv("../data/01_data_timeline_radiation.tsv")
data_prior_meds  <- read_tsv("../data/01_data_timeline_prior_meds.tsv")
data_cna <- read_tsv("../data/01_data_cna.tsv")
data_mutations <- read_tsv("../data/01_data_mutations.tsv")
data_sv <- read_tsv("../data/01_data_sv.tsv")
```

# Clinical data

## Clinical patient data

Note: the NLP-derived tumor site data will be taken from data_timeline_tumor_sites. This is done to ensure we only use information that was available at the time of the surgery.

### Clean the data

```{r}
#| label: clean_data_clinical_patient

data_clinical_patient_clean <- data_clinical_patient |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  mutate(RACE = case_when(RACE == "White" ~ "White",
                          RACE == "Black Or African American" ~ "Black",
                          RACE == "Asian-Far East/Indian Subcont" ~ "Asian",
                          RACE == "Unknown" | RACE == "Pt Refused to Answer" ~ NA,
                          TRUE ~ "Other"),
         ETHNICITY = case_when(ETHNICITY == "Non-Spanish; Non-Hispanic" ~ "Non-Spanish; Non-Hispanic",
                               ETHNICITY == "Unknown whether Spanish or not" ~ NA,
                               TRUE ~ "Spanish/Hispanic"),
         AGE_BRACKET = case_when(CURRENT_AGE_DEID < 40 ~ "<40",
                                 CURRENT_AGE_DEID < 50 ~ "40-49",
                                 CURRENT_AGE_DEID < 60 ~ "50-59",
                                 CURRENT_AGE_DEID < 70 ~ "60-69",
                                 CURRENT_AGE_DEID < 80 ~ "70-79",
                                 CURRENT_AGE_DEID < 90 ~ "80-89",
                                 TRUE ~ "90+"),
         SMOKING = ifelse(SMOKING_PREDICTIONS_3_CLASSES == "Unknown", 
                          NA, 
                          SMOKING_PREDICTIONS_3_CLASSES),
         PRIOR_TREATMENT = ifelse(PRIOR_MED_TO_MSK == "Unknown", 
                                  NA, 
                                  PRIOR_MED_TO_MSK)) |> 
  select(PATIENT_ID,
         GENDER,
         RACE,
         ETHNICITY,
         AGE = CURRENT_AGE_DEID,
         AGE_BRACKET,
         SMOKING,
         PRIOR_TREATMENT
         )

write_tsv(x = data_clinical_patient_clean,
          file = "../data/03_data_clinical_patient_clean.tsv")

data_clinical_patient <- data_clinical_patient_clean |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID))
```

### Table 1

```{r}
#| label: table_1_clinical_patient
#| warning: false

table1(~ factor(GENDER) + 
         factor(RACE) + 
         factor(ETHNICITY) +
         AGE + 
         factor(AGE_BRACKET) +
         factor(SMOKING) +
         factor(PRIOR_TREATMENT), 
       data = data_clinical_patient_clean)
```

### Variable effects on OS and DFS

#### Gender

```{r}
#| label: Gender
#| warning: false
#| message: false

fit_OS_GENDER <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ GENDER, 
                         data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ GENDER, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_GENDER,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Female", "Male"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_GENDER <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ GENDER, 
                         data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ GENDER, 
      data = data_clinical_patient)

p2 <- ggsurvplot(
  fit = fit_DFS_GENDER,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Female", "Male"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Race

```{r}
#| label: Race
#| warning: false
#| message: false

fit_OS_RACE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ RACE,
                       data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ RACE, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_RACE,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Asian", "Black", "Other", "White"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_RACE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ RACE, 
                        data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ RACE, 
      data = data_clinical_patient)

p2 <- ggsurvplot(
  fit = fit_DFS_RACE,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Asian", "Black", "Other", "White"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Ethnicity

```{r}
#| label: Ethnicity
#| warning: false
#| message: false

fit_OS_ETHNICITY <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ ETHNICITY,
                            data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ ETHNICITY, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_ETHNICITY,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Non-Spanish; Non-Hispanic", "Spanish/Hispanic"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_ETHNICITY <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ ETHNICITY, 
                             data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ ETHNICITY, 
      data = data_clinical_patient)

p2 <- ggsurvplot(
  fit = fit_DFS_ETHNICITY,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Non-Spanish; Non-Hispanic", "Spanish/Hispanic"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Age

```{r}
#| label: Age
#| warning: false
#| message: false

fit_OS_AGE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ AGE_BRACKET,
                      data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ AGE, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_AGE,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("< 40", "40-49", "50-59", "60-69", "70-79", "80-89"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_AGE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ AGE_BRACKET, 
                       data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ AGE, 
      data = data_clinical_patient)


p2 <- ggsurvplot(
  fit = fit_DFS_AGE,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("< 40", "40-49", "50-59", "60-69", "70-79", "80-89"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Smoking status (NLP)

```{r}
#| label: Smoking_status
#| warning: false
#| message: false

fit_OS_SMOKING <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SMOKING, 
                         data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SMOKING, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_SMOKING,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Former/current smoker", "Never smoked"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_SMOKING <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SMOKING,
                          data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SMOKING, 
      data = data_clinical_patient)

p2 <- ggsurvplot(
  fit = fit_DFS_SMOKING,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Former/current smoker", "Never smoked"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Prior treatment before MSK (NLP)

```{r}
#| label: Prior_treament
#| warning: false
#| message: false

fit_OS_PRIOR_TREATMENT <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PRIOR_TREATMENT,
                                   data = data_clinical_patient)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PRIOR_TREATMENT, 
      data = data_clinical_patient)

p1 <- ggsurvplot(
  fit = fit_OS_PRIOR_TREATMENT,
  data = data_clinical_patient,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("No prior treatments", "Prior treatments"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_PRIOR_TREATMENT <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PRIOR_TREATMENT,
                                    data = data_clinical_patient)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PRIOR_TREATMENT, 
      data = data_clinical_patient)

p2 <- ggsurvplot(
  fit = fit_DFS_PRIOR_TREATMENT,
  data = data_clinical_patient,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("No prior treatments", "Prior treatments"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Clinical sample data

### Select samples

```{r}
#| label: select_data_clinical_sample

## Select only patients who were samples before or during the surgery
patients_sampled <- data_train |> 
  inner_join(data_timeline_sample_acquisition,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  pull(PATIENT_ID)

## Identify patients with multiple samples
patients_2_samples <- data_timeline_sample_acquisition |> 
  filter(PATIENT_ID %in% patients_sampled) |> 
  group_by(PATIENT_ID) |> 
  summarise(n = n()) |> 
  filter(n > 1) |> 
  pull(PATIENT_ID)

## Select their CRC sample
patients_2_samples <- data_clinical_sample |> 
  filter(PATIENT_ID %in% patients_2_samples,
         CANCER_TYPE == "Colorectal Cancer")  |> 
  select(PATIENT_ID, SAMPLE_ID)

## Filter the sample data
data_clinical_sample_selected <- data_clinical_sample |> 
  filter(PATIENT_ID %in% patients_sampled) |> 
  left_join(patients_2_samples,
            by = join_by(PATIENT_ID),
            suffix = c("", "_SELECTED")) |> 
  filter(is.na(SAMPLE_ID_SELECTED) | SAMPLE_ID == SAMPLE_ID_SELECTED) |> 
  select(-SAMPLE_ID_SELECTED)

map_sample_patient <- data_clinical_sample_selected |> 
  select(SAMPLE_ID, PATIENT_ID)

write_tsv(x = map_sample_patient,
          file = "../data/03_map_sample_patient.tsv")
```

### Clean the data

```{r}
#| label: clean_data_clinical_sample

data_clinical_sample_clean <- data_clinical_sample_selected |> 
  mutate(SAMPLE_TYPE = ifelse(SAMPLE_TYPE == "Unknown", 
                              NA,
                              SAMPLE_TYPE),
         PRIMARY_SITE = case_when(PRIMARY_SITE == "Ascending colon" | PRIMARY_SITE == "Ascending Colon" ~ "Ascending Colon",
                                  PRIMARY_SITE == "Bowel" ~ "Bowel",
                                  PRIMARY_SITE == "Cecum" ~ "Cecum",
                                  PRIMARY_SITE == "colon" | PRIMARY_SITE == "Colon" ~ "Colon",
                                  PRIMARY_SITE == "Descending Colon" ~ "Descending Colon",
                                  PRIMARY_SITE == "Hepatic flexure" ~ "Hepatic Flexure",
                                  PRIMARY_SITE == "Rectosigmoid Colon" | PRIMARY_SITE == "Rectosigmoid Colon,Bowel" ~ "Rectosigmoid Colon",
                                  PRIMARY_SITE == "Rectum" ~ "Rectum",
                                  PRIMARY_SITE == "Sigmoid colon" | PRIMARY_SITE == "Sigmoid Colon" ~ "Sigmoid Colon",
                                  PRIMARY_SITE == "Transverse colon" | PRIMARY_SITE == "Transverse Colon" ~ "Transverse Colon"),
         MSI_TYPE = ifelse(MSI_TYPE == "Do not report", 
                           NA, 
                           MSI_TYPE),
         SAMPLE_COVERAGE_CAT = ifelse(SAMPLE_COVERAGE < median(SAMPLE_COVERAGE, 
                                                               na.rm = TRUE),
                                      "low",
                                      "high"),
         TUMOR_PURITY_CAT = ifelse(TUMOR_PURITY < median(TUMOR_PURITY, 
                                                         na.rm = TRUE),
                                   "low",
                                   "high"),
         CLINICAL_GROUP_SIMPLIFIED = case_when(CLINICAL_GROUP == "0" ~ "0",
                                               CLINICAL_GROUP == "1" ~ "1",
                                               CLINICAL_GROUP %in% c("2", "2A", "2B", "2C") ~ "2",
                                               CLINICAL_GROUP %in% c("3", "3A", "3B", "3C") ~ "3"),
         PATHOLOGICAL_GROUP_SIMPLIFIED = case_when(PATHOLOGICAL_GROUP == "0" ~ "0",
                                                   PATHOLOGICAL_GROUP == "1" ~ "1",
                                                   PATHOLOGICAL_GROUP %in% c("2", "2A", "2B", "2C") ~ "2",
                                                   PATHOLOGICAL_GROUP %in% c("3", "3A", "3B", "3C") ~ "3"),
         CLINICAL_SUMMARY_SIMPLIFIED = case_when(CLINICAL_SUMMARY %in% c("Distant", "Distant Metastases/Systemic Disease") ~ "Distant",
                                                 CLINICAL_SUMMARY %in% c("Regional Both 2 and 3", "Regional by Direct Extension", "Regional Nos", "Regional To Lymph Nodes", "Regional,Direct Extension", "Regional,Extension and Nodes", "Regional,Lymph Nodes Only") ~ "Regional",
                                                 CLINICAL_SUMMARY %in% c("Unknown/Unstaged", "Unstaged Unknown") ~ NA,
                                                 TRUE ~ CLINICAL_SUMMARY),
         TMB_CAT = ifelse(TMB_NONSYNONYMOUS < median(TMB_NONSYNONYMOUS, 
                                                     na.rm = TRUE),
                                      "low",
                                      "high")) |> 
  select(PATIENT_ID,
         SAMPLE_ID,
         SAMPLE_TYPE,
         PRIMARY_SITE,
         GENE_PANEL,
         SAMPLE_COVERAGE,
         SAMPLE_COVERAGE_CAT,
         TUMOR_PURITY,
         TUMOR_PURITY_CAT,
         ONCOTREE_CODE,
         MSI_SCORE,
         MSI_TYPE,
         CLINICAL_GROUP,
         CLINICAL_GROUP_SIMPLIFIED,
         PATHOLOGICAL_GROUP,
         PATHOLOGICAL_GROUP_SIMPLIFIED,
         CLINICAL_SUMMARY,
         CLINICAL_SUMMARY_SIMPLIFIED,
         DIAGNOSIS_DESCRIPTION,
         TMB_NONSYNONYMOUS,
         TMB_CAT)

write_tsv(x = data_clinical_sample_clean,
          file = "../data/03_data_clinical_sample_clean.tsv")

data_clinical_sample <- data_clinical_sample_clean |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID))
```

### Table 1

```{r}
#| label: table_1_clinical_sample
#| warning: false

table1(~ factor(SAMPLE_TYPE) + 
         factor(PRIMARY_SITE) + 
         factor(GENE_PANEL) +
         SAMPLE_COVERAGE +
         TUMOR_PURITY +
         factor(ONCOTREE_CODE) +
         MSI_SCORE +
         factor(MSI_TYPE) +
         factor(CLINICAL_GROUP) +
         factor(PATHOLOGICAL_GROUP) +
         factor(CLINICAL_SUMMARY) +
         factor(DIAGNOSIS_DESCRIPTION) +
         TMB_NONSYNONYMOUS, 
       data = data_clinical_sample_clean)
```

### Variable effects on OS and DFS

#### Sample type

```{r}
#| label: Sample_type
#| warning: false
#| message: false

fit_OS_SAMPLE_TYPE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SAMPLE_TYPE, 
                              data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SAMPLE_TYPE, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_SAMPLE_TYPE,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Local Recurrence", "Metastasis", "Primary"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_SAMPLE_TYPE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SAMPLE_TYPE, 
                               data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SAMPLE_TYPE, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_SAMPLE_TYPE,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Local Recurrence", "Metastasis", "Primary"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Primary site

```{r}
#| label: Primary_site
#| warning: false
#| message: false

fit_OS_PRIMARY_SITE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PRIMARY_SITE, 
                               data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PRIMARY_SITE, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_PRIMARY_SITE,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Ascending Colon", "Bowel", "Cecum", "Colon", "Descending Colon", "Hepatic Flexure", "Rectosigmoid Colon", "Rectum", "Sigmoid Colon", "Transverse Colon"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_PRIMARY_SITE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PRIMARY_SITE,
                                data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PRIMARY_SITE, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_PRIMARY_SITE,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Ascending Colon", "Bowel", "Cecum", "Colon", "Descending Colon", "Hepatic Flexure", "Rectosigmoid Colon", "Rectum", "Sigmoid Colon", "Transverse Colon"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Gene panel

```{r}
#| label: Gene_panel
#| warning: false
#| message: false

fit_OS_GENE_PANEL <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ GENE_PANEL, 
                             data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ GENE_PANEL, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_GENE_PANEL,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("IMPACT341", "IMPACT410", "IMPACT468", "IMPACT505"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_GENE_PANEL <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ GENE_PANEL,
                              data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ GENE_PANEL, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_GENE_PANEL,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("IMPACT341", "IMPACT410", "IMPACT468", "IMPACT505"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Sample coverage

```{r}
#| label: Sample_coverage
#| warning: false
#| message: false

fit_OS_SAMPLE_COVERAGE_CAT <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SAMPLE_COVERAGE_CAT, 
                                      data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ SAMPLE_COVERAGE_CAT, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_SAMPLE_COVERAGE_CAT,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_SAMPLE_COVERAGE_CAT <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SAMPLE_COVERAGE_CAT,
                                       data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ SAMPLE_COVERAGE_CAT, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_SAMPLE_COVERAGE_CAT,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Tumor purity

```{r}
#| label: Tumor_purity
#| warning: false
#| message: false

fit_OS_TUMOR_PURITY_CAT <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ TUMOR_PURITY_CAT, 
                                   data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ TUMOR_PURITY_CAT, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_TUMOR_PURITY_CAT,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_TUMOR_PURITY_CAT <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ TUMOR_PURITY_CAT,
                                    data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ TUMOR_PURITY_CAT, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_TUMOR_PURITY_CAT,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Oncotree code

```{r}
#| label: Oncotree code
#| warning: false
#| message: false

fit_OS_ONCOTREE_CODE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ ONCOTREE_CODE, 
                                data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ ONCOTREE_CODE, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_ONCOTREE_CODE,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("CMC", "COAD", "COADREAD", "MACR", "READ", "SRCCR"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_ONCOTREE_CODE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ ONCOTREE_CODE,
                                 data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ ONCOTREE_CODE, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_ONCOTREE_CODE,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("CMC", "COAD", "COADREAD", "MACR", "READ", "SRCCR"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### MSI

```{r}
#| label: MSI
#| warning: false
#| message: false

fit_OS_MSI_TYPE <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ MSI_TYPE, 
                           data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ MSI_TYPE, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_MSI_TYPE,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Indeterminate", "Instable", "Stable"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_MSI_TYPE <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ MSI_TYPE,
                            data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ MSI_TYPE, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_MSI_TYPE,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Indeterminate", "Instable", "Stable"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Clinical stage

```{r}
#| label: Clinical stage
#| warning: false
#| message: false

fit_OS_CLINICAL_GROUP <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_GROUP,
                                 data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_GROUP, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_CLINICAL_GROUP,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "2A", "2B", "2C", "3A", "3B", "3C"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_CLINICAL_GROUP <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_GROUP,
                                  data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_GROUP, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_CLINICAL_GROUP,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "2A", "2B", "2C", "3A", "3B", "3C"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r}
#| label: Clinical_stage_simplified
#| warning: false
#| message: false

fit_OS_CLINICAL_GROUP_SIMPLIFIED <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_GROUP_SIMPLIFIED,
                                            data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_GROUP_SIMPLIFIED, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_CLINICAL_GROUP_SIMPLIFIED,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "3"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_CLINICAL_GROUP_SIMPLIFIED <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_GROUP_SIMPLIFIED,
                                             data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_GROUP_SIMPLIFIED, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_CLINICAL_GROUP_SIMPLIFIED,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "3"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Pathology stage

```{r}
#| label: Pathology_stage
#| warning: false
#| message: false

fit_OS_PATHOLOGICAL_GROUP <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PATHOLOGICAL_GROUP, 
                                     data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PATHOLOGICAL_GROUP, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_PATHOLOGICAL_GROUP,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "2A", "2B", "2C", "3", "3A", "3B", "3C"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_PATHOLOGICAL_GROUP <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PATHOLOGICAL_GROUP,
                                      data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PATHOLOGICAL_GROUP, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_PATHOLOGICAL_GROUP,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "2A", "2B", "2C", "3", "3A", "3B", "3C"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r}
#| label: Pathology_stage_simplified
#| warning: false
#| message: false

fit_OS_PATHOLOGICAL_GROUP_SIMPLIFIED <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PATHOLOGICAL_GROUP_SIMPLIFIED,
                                            data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ PATHOLOGICAL_GROUP_SIMPLIFIED, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_PATHOLOGICAL_GROUP_SIMPLIFIED,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "3"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_PATHOLOGICAL_GROUP_SIMPLIFIED <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PATHOLOGICAL_GROUP_SIMPLIFIED,
                                             data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ PATHOLOGICAL_GROUP_SIMPLIFIED, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_PATHOLOGICAL_GROUP_SIMPLIFIED,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("0", "1", "2", "3"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Clinical summary

```{r}
#| label: Clinical_summary
#| warning: false
#| message: false

fit_OS_CLINICAL_SUMMARY_SIMPLIFIED <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_SUMMARY_SIMPLIFIED, 
                                              data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ CLINICAL_SUMMARY_SIMPLIFIED, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_CLINICAL_SUMMARY_SIMPLIFIED,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Distant", "In Situ", "Localized", "Regional"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_CLINICAL_SUMMARY_SIMPLIFIED <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_SUMMARY_SIMPLIFIED,
                                               data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ CLINICAL_SUMMARY_SIMPLIFIED, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_CLINICAL_SUMMARY_SIMPLIFIED,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Distant", "In Situ", "Localized", "Regional"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Diagnosis description

```{r}
#| label: Diagnosis description
#| warning: false
#| message: false

fit_OS_DIAGNOSIS_DESCRIPTION <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ DIAGNOSIS_DESCRIPTION, 
                                        data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ DIAGNOSIS_DESCRIPTION, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_DIAGNOSIS_DESCRIPTION,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Anus/Anal canal/Anorectum", "Ascending colon", "Cecum", "Descending colon", "Hepatic Felxure", "Colon NOS", "Rectosigmoid junction", "Rectum", "Sigmoid colon", "Splenic flexure", "Transverse colon"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_DIAGNOSIS_DESCRIPTION <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ DIAGNOSIS_DESCRIPTION,
                                         data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ DIAGNOSIS_DESCRIPTION, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_DIAGNOSIS_DESCRIPTION,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("Anus/Anal canal/Anorectum", "Ascending colon", "Cecum", "Descending colon", "Hepatic Felxure", "Colon NOS", "Rectosigmoid junction", "Rectum", "Sigmoid colon", "Splenic flexure", "Transverse colon"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Tumor mutational burden

```{r}
#| label: TMB
#| warning: false
#| message: false

fit_OS_TMB_CAT <- survfit(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ TMB_CAT, 
                          data = data_clinical_sample)

coxph(Surv(YEARS_TO_OS_EVENT, OS_STATUS) ~ TMB_CAT, 
      data = data_clinical_sample)

p1 <- ggsurvplot(
  fit = fit_OS_TMB_CAT,
  data = data_clinical_sample,
  title = "OS",
  xlab = "Time (years)",
  ylab = "OS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

fit_DFS_TMB_CAT <- survfit(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ TMB_CAT,
                           data = data_clinical_sample)

coxph(Surv(YEARS_TO_DFS_EVENT, DFS_STATUS) ~ TMB_CAT, 
      data = data_clinical_sample)

p2 <- ggsurvplot(
  fit = fit_DFS_TMB_CAT,
  data = data_clinical_sample,
  title = "DFS",
  xlab = "Time (years)",
  ylab = "DFS probability",
  xlim = c(0, 10), 
  ylim = c(0, 1),
  break.time.by = 2,
  legend = "bottom",
  legend.labs = c("high", "low"),
  pval = TRUE,
  pval.method = TRUE,
  censor = FALSE
) 

(p1$plot + p2$plot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Most recent labs

### Clean the data

```{r}
#| label: clean_labs

data_labs_ca_15_3_clean <- data_labs_ca_15_3 |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST,
         RESULT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_labs_ca_19_9_clean <- data_labs_ca_19_9 |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST,
         RESULT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_labs_cea_clean <- data_labs_cea |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST,
         RESULT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_labs_psa_clean <- data_labs_psa |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST,
         RESULT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_labs_clean <- rbind(data_labs_ca_15_3_clean,
                         data_labs_ca_19_9_clean,
                         data_labs_cea_clean,
                         data_labs_psa_clean)

write_tsv(x = data_labs_clean,
          file = "../data/03_data_labs_clean.tsv")
```

### Distribution

```{r}
#| label: distribution_labs

p1 <- data_labs_clean |> 
  filter(TEST == "CA_15-3") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of CA 15-3 tests: ", 
                     nrow(data_labs_ca_15_3_clean),
                     " (",
                     round(nrow(data_labs_ca_15_3_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p2 <- data_labs_clean |> 
  filter(TEST == "CA_19-9") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of CA 19-9 tests: ", 
                     nrow(data_labs_ca_19_9_clean),
                     " (",
                     round(nrow(data_labs_ca_19_9_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p3 <- data_labs_clean |> 
  filter(TEST == "CEA") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of CEA tests: ", 
                     nrow(data_labs_cea_clean),
                     " (",
                     round(nrow(data_labs_cea_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p4 <- data_labs_clean |> 
  filter(TEST == "PSA") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of PSA tests: ", 
                     nrow(data_labs_psa_clean),
                     " (",
                     round(nrow(data_labs_psa_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

## Most recent tumor data

### Clean the data

```{r}
#| label: clean_tumor_data

data_tumor_gleason_clean <- data_tumor_gleason |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST = SUBTYPE,
         RESULT = GLEASON_SCORE) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_tumor_mmr_clean <- data_tumor_mmr |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST = SUBTYPE,
         RESULT = MMR_ABSENT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_tumor_pdl1_clean <- data_tumor_pdl1 |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST = SUBTYPE,
         RESULT = PDL1_POSITIVE) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_tumor_site_clean <- data_tumor_site |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         TEST = SUBTYPE,
         RESULT = TUMOR_SITE) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

data_tumor_clean <- rbind(data_tumor_gleason_clean,
                         data_tumor_mmr_clean,
                         data_tumor_pdl1_clean,
                         data_tumor_site_clean)

write_tsv(x = data_tumor_clean,
          file = "../data/03_data_tumor_clean.tsv")
```

### Distribution

```{r}
#| label: distribution_tumor_data

p1 <- data_tumor_clean |> 
  filter(TEST == "Gleason Score") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of Gleason scores: ", 
                     nrow(data_tumor_gleason_clean),
                     " (",
                     round(nrow(data_tumor_gleason_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p2 <- data_tumor_clean |> 
  filter(TEST == "MMR (IHC)") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of MMR tests: ", 
                     nrow(data_tumor_mmr_clean),
                     " (",
                     round(nrow(data_tumor_mmr_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p3 <- data_tumor_clean |> 
  filter(TEST == "PD-L1 Positive") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of PD-L1 tests: ", 
                     nrow(data_tumor_pdl1_clean),
                     " (",
                     round(nrow(data_tumor_pdl1_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p4 <- data_tumor_clean |> 
  filter(TEST == "Tumor Sites") |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Number of tumor sites: ", 
                     nrow(data_tumor_site_clean),
                     " (",
                     round(nrow(data_tumor_site_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

## Performance status (ECOG)

### Clean the data

```{r}
#| label: clean_performance_status_data

data_performance_status_clean <- data_performance_status |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  group_by(PATIENT_ID) |> 
  arrange(desc(START_DATE)) |> 
  slice(1) |> 
  ungroup() |> 
  select(PATIENT_ID,
         START_DATE,
         DATE_FIRST_SURGERY,
         ECOG) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE)

write_tsv(x = data_performance_status_clean,
          file = "../data/03_data_performance_status_clean.tsv")
```

### Distribution

```{r}
#| label: distribution_performance_status

p1 <- data_performance_status_clean |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with ECOG score: ", 
                     nrow(data_performance_status_clean),
                     " (",
                     round(nrow(data_performance_status_clean)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery", 
       y = "Number of patients tested") +
  theme_minimal()

p2 <- data_performance_status_clean |> 
  group_by(ECOG) |> 
  summarise(n = n()) |> 
  ggplot(mapping = aes(x = ECOG,
                       y = n)) +
  geom_col() +
  labs(x = "ECOG score", 
       y = "Number of patients") +
  theme_minimal()

p1 + p2
```

## Treatment history

### Clean the data

```{r}
#| label: clean_treatment_history

data_treatment_clean <- data_treatment |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         DATE_FIRST_SURGERY,
         TREATMENT = SUBTYPE,
         AGENT) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - STOP_DATE)

data_radiation_clean <- data_radiation |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         DATE_FIRST_SURGERY,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE, 
         AGENT = NA)

data_prior_meds_clean <- data_prior_meds |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID)) |> 
  filter(START_DATE <= DATE_FIRST_SURGERY) |> 
  select(PATIENT_ID,
         START_DATE,
         STOP_DATE,
         DATE_FIRST_SURGERY,
         TREATMENT = SUBTYPE) |> 
  mutate(DAYS_BEFORE_SURGERY = DATE_FIRST_SURGERY - START_DATE,
         AGENT = NA)

data_treament_history_clean <- rbind(data_treatment_clean,
                                     data_radiation_clean,
                                     data_prior_meds_clean)

write_tsv(x = data_treament_history_clean,
          file = "../data/03_data_treament_history_clean.tsv")
```

### Distribution

```{r}
#| label: distribution_treatment_history

p1 <- data_treament_history_clean |> 
  filter(TREATMENT == "Chemo") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior chemotherapy: ", 
                     filter(data_treament_history_clean, TREATMENT == "Chemo") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Chemo") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most treament chemo round ended", 
       y = "Number of patients") +
  theme_minimal()

p2 <- data_treament_history_clean |> 
  filter(TREATMENT == "Biologic") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior biological therapy: ", 
                     filter(data_treament_history_clean, TREATMENT == "Biologic") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Biologic") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

p3 <- data_treament_history_clean |> 
  filter(TREATMENT == "Investigational") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior investigational treament: ", 
                     filter(data_treament_history_clean, TREATMENT == "Investigational") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Investigational") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treament round ended", 
       y = "Number of patients") +
  theme_minimal()

p4 <- data_treament_history_clean |> 
  filter(TREATMENT == "Targeted") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior targeted treament: ", 
                     filter(data_treament_history_clean, TREATMENT == "Targeted") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Targeted") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

p5 <- data_treament_history_clean |> 
  filter(TREATMENT == "Bone Treatment") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior bone treatment: ", 
                     filter(data_treament_history_clean, TREATMENT == "Bone Treatment") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Bone Treatment") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

p6 <- data_treament_history_clean |> 
  filter(TREATMENT == "Hormone") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior hormone treament: ", 
                     filter(data_treament_history_clean, TREATMENT == "Hormone") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Hormone") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treament round ended", 
       y = "Number of patients") +
  theme_minimal()

p7 <- data_treament_history_clean |> 
  filter(TREATMENT == "Immuno") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior immunotherapy: ", 
                     filter(data_treament_history_clean, TREATMENT == "Immuno") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Immuno") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

p8 <- data_treament_history_clean |> 
  filter(TREATMENT == "Radiation Therapy") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior radiation: ", 
                     filter(data_treament_history_clean, TREATMENT == "Radiation Therapy") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Radiation Therapy") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

p9 <- data_treament_history_clean |> 
  filter(TREATMENT == "Prior Medications to MSK") |> 
  group_by(PATIENT_ID) |> 
  arrange(DAYS_BEFORE_SURGERY) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot(mapping = aes(x = DAYS_BEFORE_SURGERY)) +
  geom_histogram(binwidth = 30) +
  labs(title = str_c("Patients with prior medication to MSK: ", 
                     filter(data_treament_history_clean, TREATMENT == "Prior Medications to MSK") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n),
                     " (",
                     round(filter(data_treament_history_clean, TREATMENT == "Prior Medications to MSK") |> summarise(n = n_distinct(PATIENT_ID)) |>  pull(n)/nrow(data_clinical_patient)*100, 2),
                     "%)"),
       x = "Days before surgery that most recent treatment round ended", 
       y = "Number of patients") +
  theme_minimal()

(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

# Genomic data

## Copy number alterations data 

### Clean the data

```{r}
#| label: clean_data_cna

data_cna_clean <- data_cna |> 
  pivot_longer(-Hugo_Symbol, 
               names_to = "SAMPLE_ID",
               values_to = "CNA") |> 
  filter(SAMPLE_ID %in% data_clinical_sample$SAMPLE_ID) 

write_tsv(x = data_cna_clean,
          file = "../data/03_data_cna_clean.tsv")
```

### Distribution 

```{r}
#| label: distribution_cna

data_cna_clean |> 
  count(CNA) |> 
  ggplot(mapping = aes(x = factor(CNA), 
                        y = n, 
                        fill = factor(CNA))) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.5) +
  labs(x = "CNA state", 
       y = "Count of calls") +
  theme_minimal() +
  scale_fill_manual(values = c("-2" = "blue4", 
                               "-1.5" = "skyblue", 
                               "0" = "grey80", 
                               "2" = "red")) + 
  theme(legend.position = "None")

```

### CNA burden per sample

```{r}
#| label: cna_per_sample

data_cna_clean |> 
  filter(!is.na(CNA) & CNA != 0)  |> 
  count(SAMPLE_ID) |> 
  group_by(n) |> 
  summarise(samples_with_n_alterations = n()) |> 
  ggplot(mapping = aes(x = n, 
                       y = samples_with_n_alterations)) +
  geom_col() +
  labs(x = "N", 
       y = "Number of samples with N alterations") + 
  theme_minimal()
```

### Frequently altered genes

```{r}
#| label: cna_per_gene

data_cna_clean |> 
  filter(!is.na(CNA) & CNA != 0)  |> 
  count(Hugo_Symbol) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  ggplot(mapping = aes(x = reorder(Hugo_Symbol, n), 
                       y = n/nrow(data_clinical_sample)*100)) +
  geom_col() +
  labs(x = "Gene", 
       y = "Number of altered samples (%)") + 
  theme_minimal() + 
  coord_flip()
```

### PCA

```{r}
#| label: cna_pca

cna_matrix <- data_cna_clean |> 
  pivot_wider(names_from = "Hugo_Symbol",
              values_from = "CNA") |> 
  column_to_rownames("SAMPLE_ID") |> 
  as.matrix()

cna_matrix[is.na(cna_matrix)] <- 0 

gene_var <- apply(cna_matrix, 
                  2, 
                  var, 
                  na.rm = TRUE)

keep_genes_cna <- names(gene_var)[gene_var > 0]

cna_matrix <- cna_matrix[, 
                         keep_genes_cna, 
                         drop = FALSE]

pca_res <- prcomp(cna_matrix,
                  scale. = TRUE,
                  center = TRUE)

pca_df <- as.data.frame(pca_res$x) |> 
  rownames_to_column("SAMPLE_ID") |> 
  inner_join(map_sample_patient,
             by = join_by(SAMPLE_ID)) |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID))

var_expl <- summary(pca_res)$importance

pct1 <- round(100 * var_expl[2,1], 1)

pct2 <- round(100 * var_expl[2,2], 1)

p1 <- ggplot(data = pca_df, 
       mapping = aes(x = PC1, 
                     y = PC2,
                     color = factor(OS_STATUS))) +
  geom_point(size = 3, 
             alpha = 0.8) +
  labs(x = paste0("PC1 (", pct1, "%)"),
       y = paste0("PC2 (", pct2, "%)"),
       color = "OS") +
  theme_minimal() + 
  theme(legend.position = "top")

p2 <- ggplot(data = pca_df, 
       mapping = aes(x = PC1, 
                     y = PC2,
                     color = factor(DFS_STATUS))) +
  geom_point(size = 3, 
             alpha = 0.8) +
  labs(x = paste0("PC1 (", pct1, "%)"),
       y = paste0("PC2 (", pct2, "%)"),
       color = "DFS") +
  theme_minimal() + 
  theme(legend.position = "top")

p1 + p2
```

## Mutations data

### Clean the data

```{r}
#| label: clean_data_mutations

data_mutations_clean <- data_mutations |> 
  select(SAMPLE_ID = Tumor_Sample_Barcode,
         Hugo_Symbol,
         Variant_Classification,
         Variant_Type) |> 
  filter(SAMPLE_ID %in% data_clinical_sample$SAMPLE_ID) |> 
  group_by(SAMPLE_ID, Hugo_Symbol) |> 
  slice(1) |> 
  ungroup()

write_tsv(x = data_mutations_clean,
          file = "../data/03_data_mutations_clean.tsv")
```

### Distribution 

```{r}
#| label: distribution_mutations

data_mutations_clean |> 
  count(Variant_Type) |> 
  ggplot(mapping = aes(x = Variant_Type, 
                        y = n, 
                        fill = Variant_Type)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.5) +
  labs(x = "Variant type", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None")

data_mutations_clean |> 
  count(Variant_Classification) |> 
  ggplot(mapping = aes(x = Variant_Classification, 
                        y = n, 
                        fill = Variant_Classification)) +
  geom_col() +
  geom_text(aes(label = n), 
            hjust = -0.5) +
  labs(x = "Variant classification", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None") + 
  coord_flip()
```

### Mutational burden per sample

```{r}
#| label: mutations_per_sample

data_mutations_clean |> 
  count(SAMPLE_ID) |> 
  ggplot(mapping = aes(x = n)) +
  geom_density() +
  labs(x = "Mutations per sample", 
       y = "Density") + 
  theme_minimal()
```

### Frequently mutated genes

```{r}
#| label: mutated_gene

data_mutations_clean |> 
  count(Hugo_Symbol) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  ggplot(mapping = aes(x = reorder(Hugo_Symbol, n), 
                       y = n/nrow(data_clinical_sample)*100)) +
  geom_col() +
  labs(x = "Gene", 
       y = "Number of mutated samples (%)") + 
  theme_minimal() + 
  coord_flip()
```

### PCA

```{r}
#| label: mutations_pca

mutations_matrix <- data_mutations_clean |> 
  select(Hugo_Symbol,
         SAMPLE_ID) |> 
  mutate(value = 1) |> 
  pivot_wider(names_from = "Hugo_Symbol",
              values_from = "value",
              values_fill = list(value = 0)) |> 
  column_to_rownames("SAMPLE_ID") |> 
  as.matrix()

pca_res <- prcomp(mutations_matrix,
                  scale. = TRUE,
                  center = TRUE)

pca_df <- as.data.frame(pca_res$x) |> 
  rownames_to_column("SAMPLE_ID") |> 
  inner_join(map_sample_patient,
             by = join_by(SAMPLE_ID)) |> 
  inner_join(data_train,
             by = join_by(PATIENT_ID))

var_expl <- summary(pca_res)$importance

pct1 <- round(100 * var_expl[2,1], 1)

pct2 <- round(100 * var_expl[2,2], 1)

p1 <- ggplot(data = pca_df, 
       mapping = aes(x = PC1, 
                     y = PC2,
                     color = factor(OS_STATUS))) +
  geom_point(size = 3, 
             alpha = 0.8) +
  labs(x = paste0("PC1 (", pct1, "%)"),
       y = paste0("PC2 (", pct2, "%)"),
       color = "OS") +
  theme_minimal() + 
  theme(legend.position = "top")

p2 <- ggplot(data = pca_df, 
       mapping = aes(x = PC1, 
                     y = PC2,
                     color = factor(DFS_STATUS))) +
  geom_point(size = 3, 
             alpha = 0.8) +
  labs(x = paste0("PC1 (", pct1, "%)"),
       y = paste0("PC2 (", pct2, "%)"),
       color = "DFS") +
  theme_minimal() + 
  theme(legend.position = "top")

p1 + p2
```

## Structural variants data

### Clean the data

```{r}
#| label: clean_data_sv

data_sv_clean <- data_sv |> 
  select(SAMPLE_ID = Sample_Id,
         Site1_Hugo_Symbol,
         Site2_Hugo_Symbol,
         Class) |> 
  filter(SAMPLE_ID %in% data_clinical_sample$SAMPLE_ID)

write_tsv(x = data_sv_clean,
          file = "../data/03_data_sv_clean.tsv")
```

### Distribution 

```{r}
#| label: distribution_sv

data_sv_clean |> 
  count(Class) |> 
  ggplot(mapping = aes(x = Class, 
                        y = n, 
                        fill = Class)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.5) +
  labs(x = "SV class", 
       y = "Count of calls") +
  theme_minimal() +
  theme(legend.position = "None")
```

### Frequently alterd genes

```{r}
#| label: sv_gene

data_sv_clean |> 
  group_by(Class) |> 
  count(Site1_Hugo_Symbol) |>
  ungroup() |> 
  mutate(value = 1) |> 
  pivot_wider(names_from = Class,
              values_from = "value",
              values_fill = list(value = 0)) |> 
  arrange(desc(n)) |> 
  slice_head(n = 20) |> 
  pivot_longer(cols = c("DELETION", "DUPLICATION", "INVERSION", "TRANSLOCATION", "NA"),
               names_to = "Class",
               values_to = "value") |> 
  ggplot(mapping = aes(x = reorder(Site1_Hugo_Symbol, n), 
                       y = n/nrow(data_clinical_sample)*100,
                       fill = Class)) +
  geom_col() +
  labs(x = "Gene", 
       y = "Number of SV samples (%)") + 
  theme_minimal() + 
  coord_flip()
```
