---
title: "Regularized Cox proportional hazards models"
author: "Puck Quarles van Ufford"
format: html
---

In this analysis, the regularized Cox proportional hazards model is used to assess the predictive value of the different features in our dataset for overall survival and disease-free survival. The following steps were taken.

1.  The glmnet R package was used for applying elastic-net regularization to the Cox regression. The optimal value for lambda was found using 10-fold cross-validation.

2.  The regularized Cox regression was used to exclude features for the univariate Cox models, if their beta was reduced to 0.

3.  The univariate Cox models were used to further reduce the number of features for a multivariate Cox model, using a p-value threshold of 0.05.

4.  The multivariate Cox model was used to identify features predictive of the survival outcomes.

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("janitor")
library("glmnet")
library("survival")
library("survminer")
library("knitr")
library("broom")
library("patchwork")
library("fastDummies")
library("caret")
```

```{r}
#| label: fix_mistake_survminer_package
#| echo: true

is_ggplot <- function(x)is.ggplot(x)
```

```{r}
#| label: seed
#| echo: true

set.seed(1)
```

# Settings

```{r}
#| label: settings

regularization <- TRUE
alpha <- 0.5 # 0.5 for elastic-net, 1 for lasso, 0 for ridge

prevalence_filter <- TRUE
prevalence_threshold <- 0.95
```

# Encoding the variables

In order to use the glmnet package, the variables need to be encoded, because the package cannot handle missing values and non-numerical features. The non-numerical features and features with NAs are one-hot encoded.

Missing values: - For categorical data, it is one-hot encoded and there is a column for missing values. - For numerical data: median imputation and a column for missing values. - ordinal categories; ordinal factor + mode imputation

```{r}
#| label: encode_data
#| message: false
#| echo: true

data_train <- read_tsv("../data/02_data_train_complete.tsv")

cols_to_exclude <- c("surgery_date",
                     "sample_date",
                     "last_follow_up_date",
                     "date_ecog",
                     "date_cea",
                     "date_mmr",
                     "days_to_os_event",
                     "days_to_dfs_event",
                     "clinical_group",
                     "pathological_group",
                     "race",
                     "ethnicity",
                     "age_last_follow_up",
                     "stage_highest_recorded")

numeric_cols <- c("tmb_nonsynonymous",
                  "sample_coverage",
                  "num_icdo_dx",
                  "age_surgery")

numeric_cols_with_NA <- c("tumor_purity",
                          "msi_score",
                          "cea")

cna_cols <- data_train |> 
  select(starts_with("cna_")) |>
  colnames()

ordinal_cols <- c(cna_cols,
                  "pathological_group_simplified")

ordinal_cols_with_NA <- c("clinical_group_simplified",
                          "ecog")

cols_to_one_hot_encode <- c("msi_type",
                            "race_simplified",
                            "ethnicity_simplified",
                            "smoking_status_nlp",
                            "prior_med_nlp",
                            "primary_site",
                            "gene_panel",
                            "oncotree_code",
                            "chest",
                            "abdomen",
                            "pelvis",
                            "head",
                            "other",
                            "lymph_nodes",
                            "other_2",
                            "liver",
                            "bone",
                            "lung",
                            "reproductive_organs",
                            "intra_abdominal",
                            "cns_brain",
                            "pleura",
                            "mmr_absent",
                            "chemo",
                            "investigational",
                            "hormone",
                            "biologic",
                            "bone_treatment",
                            "targeted",
                            "radiation_therapy",
                            "prior_medications_to_msk")


data_train <- data_train |> 
  select(-all_of(cols_to_exclude)) |> 
  mutate(
    across(all_of(c(numeric_cols_with_NA,
                    ordinal_cols_with_NA)),
           # indicate NAs
           ~ as.integer(is.na(.)),
           .names = "{.col}_na"),
    across(all_of(numeric_cols_with_NA),
           # impute NAs with the median
           ~ if_else(is.na(.),
                     median(., na.rm = TRUE),
                     .)),
    across(all_of(ordinal_cols_with_NA),
           #impute NAs with the mode
           ~ {
             this_mode <- names(sort(table(.),
                                     decreasing = TRUE))[1]
             if_else(is.na(.),
                     as.numeric(this_mode),
                     .)}),
    across(all_of(c(ordinal_cols,
                    ordinal_cols_with_NA)),
           # convert to ordered factors
           ~ factor(., 
                    levels = sort(unique(.)),
                    ordered = TRUE)),
    across(starts_with("mut_"), 
           # convert to binary
           ~ if_else(is.na(.x), 0, 1)),
    across(starts_with("sv_"), 
           # convert to binary
           ~ if_else(is.na(.x), 0, 1)),
    gender = if_else(gender == "Male", 0, 1),
    early_onset = if_else(early_onset, 1, 0)
    ) |> 
  dummy_cols(select_columns = cols_to_one_hot_encode,
             remove_selected_columns = TRUE) |> 
  mutate(across(starts_with(cols_to_one_hot_encode),
                ~ replace_na(., 0))) |> 
  clean_names() |> 
  remove_empty(which = "cols") |> 
  remove_constant()
```

# Filter variables

For the non-numerical explanatory variables, we need to remove the features that have a level or category with zero events or zero controls, because the Cox model won't work for those.

```{r}
#| label: prevalence_filter
#| message: false
#| echo: true

if(prevalence_filter){
  col_prevalence <- data_train |> 
    select(-c(patient_id,
              sample_id,
              contains("status"),
              contains("event"))) |> 
    summarise(across(everything(),
                     ~ max(table(.)/n()))) |> 
    pivot_longer(everything(),
                 names_to = "column",
                 values_to = "prevalence")
  
  cols_to_keep <- col_prevalence |> 
    filter(prevalence <= prevalence_threshold) |> 
    pull(column)
    
  data_train <- data_train |> 
    select(patient_id, 
           sample_id, 
           os_status,
           years_to_os_event,
           dfs_status,
           years_to_dfs_event,
           all_of(cols_to_keep))
}
```

```{r}
#| label: linear_combo_filter
#| echo: true

explanatory_vars <- setdiff(colnames(data_train),
                            c("patient_id",
                              "sample_id",
                              "os_status",
                              "years_to_os_event",
                              "dfs_status",
                              "years_to_dfs_event"))

explanatory_data <- data_train |> 
  select(all_of(explanatory_vars))

mm <- model.matrix(~ . -1, 
                   data = explanatory_data)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

cols_to_remove <- str_remove(cols_to_remove, 
                             "\\..*$")

data_train <- data_train |> 
  select(-all_of(cols_to_remove))
```

```{r}
#| label: exclude_vars_no_event_levels
#| echo: true

explanatory_vars <- setdiff(colnames(data_train),
                            c("patient_id",
                              "sample_id",
                              "os_status",
                              "years_to_os_event",
                              "dfs_status",
                              "years_to_dfs_event"))

to_test <- setdiff(explanatory_vars, c(numeric_cols, 
                                       numeric_cols_with_NA))

to_remove_os <- c()

for(var in to_test){
  test <- data_train |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

print(paste("Number of features to remove due to no OS variation:", 
            length(to_remove_os)))

to_remove_dfs <- c()

for(var in to_test){
  test <- data_train |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

print(paste("Number of features to remove due to no DFS variation:", 
            length(to_remove_dfs)))
```

# Regularized CPH models

## OS

```{r}
#| label: regularized_cph_os
#| echo: true

x_os <- data_train |> 
  select(-c(patient_id, 
            sample_id,
            contains("status"),
            contains("event"),
            all_of(to_remove_os))) |> 
  as.matrix()

y_os <- Surv(data_train$years_to_os_event,
             data_train$os_status)

if (regularization) {
  cvfit_os <- cv.glmnet(x_os, 
                        y_os,
                        family = "cox",
                        type.measure = "C",
                        alpha = alpha)
  
  lambda_os <- cvfit_os$lambda.min
  print(paste("Lambda:", lambda_os))
  
  betas_os <- coef(cvfit_os, 
                   s = lambda_os)
  beta_vec_os <- as.numeric(betas_os)
  nonzero_features_os <- rownames(betas_os)[which(beta_vec_os != 0)]
  
  print(paste("n non-zero features:",
              length(nonzero_features_os)))
  
  plot(cvfit_os, 
       xlab = "Log(lambda)")
  abline(v = log(lambda_os), lty = 2, col = "red")
  title("Effect of penalty strength lambda on the C-index for predicting OS.", line = 2.5)
  
} else {
  nonzero_features_os <- colnames(x_os)
  
  print(paste("n features:", length(nonzero_features_os)))
}
```

## DFS

```{r}
#| label: regularized_cph_dfs
#| echo: true

x_dfs <- data_train |> 
  select(-c(patient_id, 
            sample_id,
            contains("status"),
            contains("event"),
            all_of(to_remove_dfs))) |> 
  as.matrix()

y_dfs <- Surv(data_train$years_to_dfs_event,
              data_train$dfs_status)

if (regularization) {
  cvfit_dfs <- cv.glmnet(x_dfs, 
                         y_dfs,
                         family = "cox",
                         type.measure = "C",
                         alpha = alpha)
  
  lambda_dfs <- cvfit_dfs$lambda.min
  print(paste("Lambda:", lambda_dfs))
  
  betas_dfs <- coef(cvfit_dfs, s = lambda_dfs)
  beta_vec_dfs <- as.numeric(betas_dfs)
  nonzero_features_dfs <- rownames(betas_dfs)[which(beta_vec_dfs != 0)]
  
  print(paste("n non-zero features:", 
              length(nonzero_features_dfs)))
  
  plot(cvfit_dfs, 
       xlab = "Log(lambda)")
  abline(v = log(lambda_dfs), lty = 2, col = "red")
  title("Effect of penalty strength lambda on the C-index for predicting DFS.", line = 2.5)
  
} else {
  nonzero_features_dfs <- colnames(x_dfs)
  
  print(paste("n features:", length(nonzero_features_dfs)))
}
```

# Univariate CPH models

### OS

```{r}
#| label: univ_cph_OS
#| echo: true
#| warning: false

formulas_OS <- sapply(nonzero_features_os,
                      function(x){
                        as.formula(paste("Surv(years_to_os_event, os_status) ~", 
                                         x))
                        })

models_OS <- lapply(formulas_OS,
                    function(x){
                      coxph(x, 
                            data = data_train)
                      })

res_OS <- lapply(models_OS,
                 function(x){
                   x <- summary(x)
                   p <- signif(x$wald["pvalue"],
                               digits = 3)
                   wald <- signif(x$wald["test"],
                                  digits = 3)
                   beta <- signif(x$coef[1],
                                  digits = 3)
                   hr <- signif(x$coef[2],
                                digits = 3)
                   res <- c(beta, hr, wald, p)
                   names(res) <- c("beta", "hr", "wald_test", "p_value")
                   return(res)
                   }) |> 
  as.data.frame() |> 
  t() |> 
  as_tibble(rownames = NA) |> 
  rownames_to_column("var")

significant_vars_OS <- res_OS |> 
  arrange(p_value) |> 
  filter(p_value < 0.05) |> 
  pull(var)

res_OS |> 
  arrange(p_value) |> 
  slice_head(n = 10) |> 
  kable(digits = 5,
        caption = "Top 10 most significant variables for predicing OS")

print(paste("Number of significant features: ",
            length(significant_vars_OS)))
```

### DFS

```{r}
#| label: univ_cph_DFS
#| echo: true
#| warning: false

formulas_DFS <- sapply(nonzero_features_dfs,
                       function(x){
                         as.formula(paste("Surv(years_to_dfs_event, dfs_status) ~", 
                                          x))
                         })

models_DFS <- lapply(formulas_DFS,
                     function(x){
                       coxph(x, 
                             data = data_train)
                       })

res_DFS <- lapply(models_DFS,
                  function(x){
                    x <- summary(x)
                    p <- signif(x$wald["pvalue"],
                                digits = 3)
                    wald <- signif(x$wald["test"],
                                   digits = 3)
                    beta <- signif(x$coef[1],
                                   digits = 3)
                    hr <- signif(x$coef[2],
                                 digits = 3)
                    res <- c(beta, hr, wald, p)
                    names(res) <- c("beta", "hr", "wald_test", "p_value")
                    return(res)
                    }) |> 
  as.data.frame() |> 
  t() |> 
  as_tibble(rownames = NA) |> 
  rownames_to_column("var")

significant_vars_DFS <- res_DFS |> 
  arrange(p_value) |> 
  filter(p_value < 0.05) |> 
  pull(var)

res_DFS |> 
  arrange(p_value) |> 
  slice_head(n = 10) |> 
  kable(digits = 5,
        caption = "Top 10 most significant variables for predicting DFS")

print(paste("Number of significant features: ",
            length(significant_vars_DFS)))
```

# Multivariate CPH

## OS

```{r}
#| label: multivariate_cph_OS
#| echo: true
#| warning: false
#| fig.width: 10
#| fig.height: 30

formula_multiv_os <- as.formula(str_c("Surv(years_to_os_event, os_status) ~ ",
                                      paste(significant_vars_OS,
                                            collapse = " + ")))

model_multiv_os <- coxph(formula_multiv_os, 
                         data = data_train)

print(paste("C index:", summary(model_multiv_os)$concordance[1] ))
print(paste("p value:", summary(model_multiv_os)$logtest["pvalue"] ))


model_multiv_os_tidy <- tidy(model_multiv_os, 
                             exponentiate = TRUE, 
                             conf.int = TRUE)

model_multiv_os_tidy |> 
  mutate(significant = ifelse(p.value < 0.05, "Yes", "No")) |> 
  ggplot(mapping = aes(x = estimate, 
                       y = term,
                       color = significant)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low,
                     xmax = conf.high)) +  
  geom_vline(xintercept = 1, 
             linetype = "dashed") +  
  scale_x_log10() + 
  scale_color_manual(values = c("Yes" = "red", "No" = "grey40")) +
  labs(
    x = "Hazard Ratio (95% CI, log scale)",
    y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 12)
  )

significant_mutiv_os <- model_multiv_os_tidy |> 
  filter(p.value < 0.05) |> 
  pull(term)

print(paste("Number of significant features:", 
            length(significant_mutiv_os)))

print(significant_mutiv_os)
```

## DFS

```{r}
#| label: multivariate_cph_DFS
#| echo: true
#| warning: false
#| fig.width: 10
#| fig.height: 15

formula_multiv_dfs <- as.formula(str_c("Surv(years_to_dfs_event, dfs_status) ~ ",
                                       paste(significant_vars_DFS,
                                             collapse = " + ")))

model_multiv_dfs <- coxph(formula_multiv_dfs,
                          data = data_train)

print(paste("C index:", summary(model_multiv_dfs)$concordance[1] ))
print(paste("p value:", summary(model_multiv_dfs)$logtest["pvalue"] ))

model_multiv_dfs_tidy <- tidy(model_multiv_dfs, 
                              exponentiate = TRUE, 
                              conf.int = TRUE)

model_multiv_dfs_tidy |> 
  mutate(significant = ifelse(p.value < 0.05, "Yes", "No")) |> 
  ggplot(mapping = aes(x = estimate, 
                       y = term,
                       color = significant)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low,
                     xmax = conf.high)) +  
  geom_vline(xintercept = 1, 
             linetype = "dashed") +  
  scale_x_log10() + 
  scale_color_manual(values = c("Yes" = "red", "No" = "grey40")) +
  labs(
    x = "Hazard Ratio (95% CI, log scale)",
    y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 12)
  )

significant_mutiv_dfs <- model_multiv_dfs_tidy |> 
  filter(p.value < 0.05) |> 
  pull(term)

print(paste("Number of significant features:", 
            length(significant_mutiv_dfs)))

print(significant_mutiv_dfs)
```
