---
title: "Cox proportional hazards models"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("glmnet")
library("survival")
library("caret")
```

```{r}
#| label: seed
#| echo: true

set.seed(1)
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train_os <- read_tsv("../data/03_data_train_os.tsv")

data_train_dfs <- read_tsv("../data/03_data_train_dfs.tsv")
```

Define feature types.

```{r}
#| label: feature_sets

id_cols <- c("patient_id",
             "sample_id")

outcome_cols <- c("os_status",
                  "years_to_os_event",
                  "dfs_status",
                  "years_to_dfs_event")

cna_features <- data_train_os |> 
  select(starts_with("cna")) |> 
  colnames()

mut_features <- data_train_os |> 
  select(starts_with("mut")) |> 
  colnames()

sv_features <- data_train_os |> 
  select(starts_with("sv")) |> 
  colnames()

radiology_features <- data_train_os |> 
  select(starts_with("chest"),
         starts_with("abdomen"),
         starts_with("pelvis"),
         starts_with("head"),
         starts_with("other"),
         starts_with("lymph_nodes"),
         starts_with("liver"),
         starts_with("bone"),
         starts_with("lung"),
         starts_with("reproductive_organs"),
         starts_with("intra_abdominal"),
         starts_with("cns"),
         starts_with("pleura")) |> 
  colnames()

treatment_features <- data_train_os |> 
  select(starts_with("biologic"),
         starts_with("chemo"),
         starts_with("hormone"),
         starts_with("investigational"),
         starts_with("prior"),
         starts_with("radiation"),
         starts_with("targeted")) |> 
  colnames()

nlp_features <- c("smoking_status_nlp_former_current_smoker",
                  "smoking_status_nlp_never",
                  "smoking_status_nlp_unknown",
                  "prior_med_nlp_no_prior_medications",
                  "prior_med_nlp_prior_medications_to_msk",
                  "prior_med_nlp_unknown",
                  "mmr_absent_false",
                  "mmr_absent_na",
                  "mmr_absent_true")

clinical_features <- setdiff(colnames(data_train_os),
        c(id_cols, 
          outcome_cols, 
          cna_features, 
          mut_features, 
          sv_features, 
          nlp_features,
          radiology_features, 
          treatment_features)) 
```

# Settings

```{r}
#| label: settings


features_included <- c(clinical_features, 
                       #radiology_features,
                       treatment_features)#,
                       #nlp_features,
                       #cna_features,
                       #mut_features,
                       #sv_features)

caption <- "noNLPgenomic" 

regularization <- FALSE
alpha <- 0.5 # 0.5 for elastic-net, 1 for lasso, 0 for ridge

prevalence_filter <- TRUE
prevalence_threshold <- 0.95
```

# Prepare data

Select included features.

```{r}
#| label: select_feature_sets

data_train_os <- data_train_os |> 
  select(all_of(c(outcome_cols,
                  features_included))) 

data_train_dfs <- data_train_dfs |> 
  select(all_of(c(outcome_cols,
                  features_included))) 
```

Optional prevalence filter.

```{r}
#| label: prevalence_filter
#| message: false
#| echo: true

if(prevalence_filter){
  # OS
  col_prevalence_os <- data_train_os |> 
    select(-all_of(outcome_cols)) |> 
    summarise(across(everything(),
                     ~ max(table(.)/n()))) |> 
    pivot_longer(everything(),
                 names_to = "column",
                 values_to = "prevalence")
  
  cols_to_keep_os <- col_prevalence_os |> 
    filter(prevalence <= prevalence_threshold) |> 
    pull(column)
    
  data_train_os <- data_train_os |> 
    select(all_of(c(outcome_cols, 
                    cols_to_keep_os)))
  
  # DFS
  col_prevalence_dfs <- data_train_dfs |> 
    select(-all_of(outcome_cols)) |> 
    summarise(across(everything(),
                     ~ max(table(.)/n()))) |> 
    pivot_longer(everything(),
                 names_to = "column",
                 values_to = "prevalence")
  
  cols_to_keep_dfs <- col_prevalence_dfs |> 
    filter(prevalence <= prevalence_threshold) |> 
    pull(column)
    
  data_train_dfs <- data_train_dfs |> 
    select(all_of(c(outcome_cols, 
                    cols_to_keep_dfs)))
}
```

Remove features that are linear combinations of other features.

```{r}
#| label: linear_combo_filter
#| echo: true

# OS
feature_data_os <- data_train_os |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_os)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_os <- data_train_os |> 
  select(-all_of(cols_to_remove))

# DFS
feature_data_dfs <- data_train_dfs |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_dfs)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_dfs <- data_train_dfs |> 
  select(-all_of(cols_to_remove))
```

For the categorical features, we need to remove the features that have a category with zero events or zero controls.

```{r}
#| label: exclude_vars_no_event_levels
#| echo: true

numerical_featurs <- c("tmb_nonsynonymous",
                       "sample_coverage",
                       "num_icdo_dx",
                       "age_surgery",
                       "tumor_purity",
                       "msi_score",
                       "cea")

# OS
explanatory_vars_os <- setdiff(colnames(data_train_os),
                               outcome_cols)

to_test_os <- setdiff(explanatory_vars_os,
                      numerical_featurs)

to_remove_os <- c()

for(var in to_test_os){
  test <- data_train_os |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

data_train_os <- data_train_os |> 
  select(-all_of(to_remove_os)) |>
  select(-c("dfs_status", 
            "years_to_dfs_event"))

# DFS
explanatory_vars_dfs <- setdiff(colnames(data_train_dfs),
                                outcome_cols)

to_test_dfs <- setdiff(explanatory_vars_dfs,
                       numerical_featurs)

to_remove_dfs <- c()

for(var in to_test_dfs){
  test <- data_train_dfs |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

data_train_dfs <- data_train_dfs |> 
  select(-all_of(to_remove_dfs)) |>
  select(-c("os_status", 
            "years_to_os_event"))
```

# Cox model workflow

Set up cross-validation folds.

```{r}
#| label: 5_fold_CV
#| echo: true

folds_os <- createFolds(data_train_os$os_status,
                        k = 5,
                        list = TRUE,
                        returnTrain = FALSE)

folds_dfs <- createFolds(data_train_dfs$dfs_status,
                         k = 5,
                         list = TRUE,
                         returnTrain = FALSE)
```

Function for the Cox model feature selection workflow.

```{r}
#| label: cox_workflow_function
#| echo: true

run_cox_workflow <- function(data,
                             time_var,
                             status_var,
                             p_cutoff = 0.05) {
  
  outcome_cols <- c(time_var, 
                    status_var)
  
  features_all <- setdiff(colnames(data), 
                          outcome_cols)
  
  # Regularized Cox model
  x <- data |> 
    select(all_of(features_all)) |> 
    model.matrix(~ . -1, 
                 data = _)

  y <- Surv(data[[time_var]], 
            data[[status_var]])

  if (regularization) {

    cvfit <- cv.glmnet(x,
                       y,
                       family = "cox",
                       type.measure = "C",
                       alpha = alpha)

    lambda_opt <- cvfit$lambda.min

    betas <- coef(cvfit, 
                  s = lambda_opt)
    
    beta_vec <- as.numeric(betas)

    selected_features <- rownames(betas)[beta_vec != 0]

  } else {
    selected_features <- colnames(x)
  }

  # Univariate Cox model

  univ_results <- map_dfr(
    selected_features,
    function(var) {

      f <- as.formula(
        paste0("Surv(", time_var, ", ", status_var, ") ~ ", 
               var)
      )

      fit <- coxph(f, 
                   data = data)
      
      s <- summary(fit)

      tibble(
        variable = var,
        p_value = s$wald["pvalue"]
      )
    }
  )

  significant_vars <- univ_results |> 
    arrange(p_value) |> 
    filter(p_value < p_cutoff) |> 
    pull(variable)

  # Multivariate Cox model
  multiv_formula <- as.formula(
      paste0(
        "Surv(", time_var, ", ", status_var, ") ~ ",
        paste(significant_vars, 
              collapse = " + ")
      )
    )

    coxph(multiv_formula, 
          data = data)
}
```

## Predicting overall survival

```{r}
#| label: cph_os_cv
#| echo: true
#| warning: false

res_os <- map_dbl(folds_os,
                  ~ {
                    
        test_idx  <- .x
        train_idx <- setdiff(seq_len(nrow(data_train_os)), 
                             test_idx)

        train_data <- data_train_os[train_idx, ]
        test_data  <- data_train_os[test_idx, ]

        fit <- run_cox_workflow(train_data,
                                "years_to_os_event",
                                "os_status")

        pred <- predict(fit, 
                        newdata = test_data)
        
        risk <- -pred
        
        concordance(Surv(test_data$years_to_os_event, 
                         test_data$os_status) ~ risk)$concordance
      }
    )

summary_os <- tibble(mean_c_index = mean(res_os),
                     sd_c_index = sd(res_os),
                     lower_95 = quantile(res_os, 0.025),
                     upper_95 = quantile(res_os, 0.975)) |> 
  mutate(model = "Cox",
         features = caption,
         outcome = "OS",
         setting = ifelse(regularization,
                          case_when(alpha == 1 ~ "lasso",
                                    alpha == 0.5 ~ "elastic-net",
                                    TRUE ~ "ERROR"),
                          ifelse(prevalence_filter,
                                 "prevalence_filter",
                                 "None"))) |> 
  select(c(model, features, outcome, 
           mean_c_index, sd_c_index,
           lower_95, upper_95, setting))

summary_os

write_tsv(summary_os,
          file = "../results/04_Cox_5CV_model_results.tsv",
          append = TRUE)
```

## Predicting disease-free survival

```{r}
#| label: cph_dfs_cv
#| echo: true
#| warning: false

res_dfs <- map_dbl(folds_dfs,
                   ~ {
                    
        test_idx  <- .x
        train_idx <- setdiff(seq_len(nrow(data_train_dfs)), 
                             test_idx)

        train_data <- data_train_dfs[train_idx, ]
        test_data  <- data_train_dfs[test_idx, ]

        fit <- run_cox_workflow(train_data,
                                "years_to_dfs_event",
                                "dfs_status")

        pred <- predict(fit, 
                        newdata = test_data)
        
        risk <- -pred
        
        concordance(Surv(test_data$years_to_dfs_event, 
                         test_data$dfs_status) ~ risk)$concordance
      }
    )

summary_dfs <- tibble(mean_c_index = mean(res_dfs),
                      sd_c_index = sd(res_dfs),
                      lower_95 = quantile(res_dfs, 0.025),
                      upper_95 = quantile(res_dfs, 0.975)) |> 
  mutate(model = "Cox",
         features = caption,
         outcome = "DFS",
         setting = ifelse(regularization,
                          case_when(alpha == 1 ~ "lasso",
                                    alpha == 0.5 ~ "elastic-net",
                                    TRUE ~ "ERROR"),
                          ifelse(prevalence_filter,
                                 "prevalence_filter",
                                 "none"))) |> 
  select(c(model, features, outcome, 
           mean_c_index, sd_c_index,
           lower_95, upper_95, setting))

summary_dfs

write_tsv(summary_dfs,
          file = "../results/04_Cox_5CV_model_results.tsv",
          append = TRUE)
```
