---
title: "Step-wise Cox proportional hazards models"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("survival")
library("caret")
```

```{r}
#| label: seed
#| echo: true

set.seed(1)
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train_os <- read_tsv("../data/03_data_train_os.tsv")

data_train_dfs <- read_tsv("../data/03_data_train_dfs.tsv")
```

```{r}
#| label: feature_sets

id_cols <- c("patient_id",
             "sample_id")

outcome_cols <- c("os_status",
                  "years_to_os_event",
                  "dfs_status",
                  "years_to_dfs_event")

cna_features <- data_train_os |> 
  select(starts_with("cna")) |> 
  colnames()

mut_features <- data_train_os |> 
  select(starts_with("mut")) |> 
  colnames()

sv_features <- data_train_os |> 
  select(starts_with("sv")) |> 
  colnames()

radiology_features <- data_train_os |> 
  select(starts_with("chest"),
         starts_with("abdomen"),
         starts_with("pelvis"),
         starts_with("head"),
         starts_with("other"),
         starts_with("lymph_nodes"),
         starts_with("liver"),
         starts_with("bone"),
         starts_with("lung"),
         starts_with("reproductive_organs"),
         starts_with("intra_abdominal"),
         starts_with("cns"),
         starts_with("pleura")) |> 
  colnames()

treatment_features <- data_train_os |> 
  select(starts_with("biologic"),
         starts_with("chemo"),
         starts_with("hormone"),
         starts_with("investigational"),
         starts_with("prior"),
         starts_with("radiation"),
         starts_with("targeted")) |> 
  colnames()

nlp_features <- c("smoking_status_nlp_former_current_smoker",
                  "smoking_status_nlp_never",
                  "smoking_status_nlp_unknown",
                  "prior_med_nlp_no_prior_medications",
                  "prior_med_nlp_prior_medications_to_msk",
                  "prior_med_nlp_unknown",
                  "mmr_absent_false",
                  "mmr_absent_na",
                  "mmr_absent_true")

clinical_features <- setdiff(colnames(data_train_os),
        c(id_cols, 
          outcome_cols, 
          cna_features, 
          mut_features, 
          sv_features, 
          nlp_features,
          radiology_features, 
          treatment_features)) 
```

# Settings

```{r}
#| label: settings

features_included <- c(clinical_features, 
                       #radiology_features,
                       treatment_features)#,
                       #nlp_features,
                       #cna_features,
                       #mut_features,
                       #sv_features)

caption <- "clinical, treatment." # copy number variations, mutations, structural variants, radiology, NLP-derived

file_name <- "noNLPgenomic_100max"

max_iter <- 100
```

# Filter variables

Select included features.

```{r}
#| label: select_feature_sets

data_train_os <- data_train_os |> 
  select(all_of(c(outcome_cols,
                  features_included))) 

data_train_dfs <- data_train_dfs |> 
  select(all_of(c(outcome_cols,
                  features_included))) 
```

Remove features that are linear combinations of other features

```{r}
#| label: linear_combo_filter
#| echo: true

feature_data_os <- data_train_os |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_os)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_os <- data_train_os |> 
  select(-all_of(cols_to_remove))

feature_data_dfs <- data_train_dfs |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_dfs)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_dfs <- data_train_dfs |> 
  select(-all_of(cols_to_remove))
```

For the categorical features, we need to remove the features that have a category with zero events or zero controls.

```{r}
#| label: exclude_vars_no_event_levels
#| echo: true

numerical_featurs <- c("tmb_nonsynonymous",
                       "sample_coverage",
                       "num_icdo_dx",
                       "age_surgery",
                       "tumor_purity",
                       "msi_score",
                       "cea")

explanatory_vars_os <- setdiff(colnames(data_train_os),
                               outcome_cols)

to_test_os <- setdiff(explanatory_vars_os,
                      numerical_featurs)

to_remove_os <- c()

for(var in to_test_os){
  test <- data_train_os |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

data_train_os <- data_train_os |> 
  select(-all_of(to_remove_os))

explanatory_vars_dfs <- setdiff(colnames(data_train_dfs),
                                outcome_cols)

to_test_dfs <- setdiff(explanatory_vars_dfs,
                       numerical_featurs)

to_remove_dfs <- c()

for(var in to_test_dfs){
  test <- data_train_dfs |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

data_train_dfs <- data_train_dfs |> 
  select(-all_of(to_remove_dfs))
```

# Cox propostional hazards models with step-wise forward feature selection

## Predicting overall survival

```{r}
#| label: cph_OS
#| echo: true
#| warning: false

features_os <- setdiff(colnames(data_train_os), 
                       outcome_cols)

max_iter_os <- ifelse(length(features_os) < max_iter,
                      length(features_os),
                      max_iter)

added_features_os <- c()

iter <- 1

result_os <- data.frame(iteration = integer(),
                        added_feature = character(),
                        C_index = double())

while (iter != max_iter_os + 1) {
  c_indeces <- data.frame(feature = character(),
                          C_index = double())
  
  for(f in features_os){
    formula <- as.formula(paste("Surv(years_to_os_event, os_status) ~",
                                paste(c(added_features_os, f),
                                      collapse = " + ")))
    
    model <- coxph(formula,
                   data = data_train_os)
    
    c_index <- summary(model)$concordance[1]
    
    c_indeces <- add_row(c_indeces,
                         feature = f, 
                         C_index = c_index)
  }

  best <- c_indeces$feature[which.max(c_indeces$C_index)]
  
  result_os <- add_row(result_os,
                       iteration = iter,
                       added_feature = best,
                       C_index = max(c_indeces$C_index))
  
  iter <- iter + 1
  features_os <- setdiff(features_os, best)
  added_features_os <- c(added_features_os, best)
}

write_tsv(x = result_os,
         file = paste0("../results/05_result_os_",
                       file_name, 
                       ".tsv"))

result_os |> 
  ggplot(mapping = aes(x = iteration,
                       y = C_index)) + 
  geom_line() +
  labs(
    x = "n features included",
    y = "C index",
    title = "Concordance of step-wise Cox model for predicing OS",
    subtitle = paste("Feature sets included: ", 
                     caption)
  ) + 
  theme_minimal()

ggsave(paste0("../results/05_C_index_os_",
                       file_name, 
                       ".png"))
```

## Predicting disease-free survival

```{r}
#| label: cph_DFS
#| echo: true
#| warning: false

features_dfs <- setdiff(colnames(data_train_dfs), 
                        outcome_cols)

max_iter_dfs <- ifelse(length(features_dfs) < max_iter,
                       length(features_dfs),
                       max_iter)

added_features_dfs <- c()

iter <- 1

result_dfs <- data.frame(iteration = integer(),
                     added_feature = character(),
                     C_index = double())

while (iter != max_iter_dfs + 1) {
  c_indeces <- data.frame(feature = character(),
                          C_index = double())
  
  for(f in features_dfs){
    formula <- as.formula(paste("Surv(years_to_dfs_event, dfs_status) ~",
                                paste(c(added_features_dfs, f),
                                      collapse = " + ")))
    
    model <- coxph(formula,
                   data = data_train_dfs)
    
    c_index <- summary(model)$concordance[1]
    
    c_indeces <- add_row(c_indeces,
                         feature = f, 
                         C_index = c_index)
  }

  best <- c_indeces$feature[which.max(c_indeces$C_index)]
  
  result_dfs <- add_row(result_dfs,
                       iteration = iter,
                       added_feature = best,
                       C_index = max(c_indeces$C_index))
  
  iter <- iter + 1
  features_dfs <- setdiff(features_dfs, best)
  added_features_dfs <- c(added_features_dfs, best)
}

write_tsv(x = result_dfs,
         file = paste0("../results/05_result_dfs_",
                       file_name, 
                       ".tsv"))

result_dfs |> 
  ggplot(mapping = aes(x = iteration,
                       y = C_index)) + 
  geom_line() +
  labs(
    x = "n features included",
    y = "C index",
    title = "Concordance of step-wise Cox model for predicing DFS",
    subtitle = paste("Feature sets included: ", 
                     caption)
  ) + 
  theme_minimal()

ggsave(paste0("../results/05_C_index_dfs_",
                       file_name, 
                       ".png"))
```
