---
title: "Step-wise Cox proportional hazards models"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("survival")
library("caret")
```

```{r}
#| label: seed
#| echo: true

set.seed(1)
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train_os <- read_tsv("../data/03_data_train_os.tsv")

data_train_dfs <- read_tsv("../data/03_data_train_dfs.tsv")
```

Define feature types.

```{r}
#| label: feature_sets

id_cols <- c("patient_id",
             "sample_id")

outcome_cols <- c("os_status",
                  "years_to_os_event",
                  "dfs_status",
                  "years_to_dfs_event")

cna_features <- data_train_os |> 
  select(starts_with("cna")) |> 
  colnames()

mut_features <- data_train_os |> 
  select(starts_with("mut")) |> 
  colnames()

sv_features <- data_train_os |> 
  select(starts_with("sv")) |> 
  colnames()

radiology_features <- data_train_os |> 
  select(starts_with("chest"),
         starts_with("abdomen"),
         starts_with("pelvis"),
         starts_with("head"),
         starts_with("other"),
         starts_with("lymph_nodes"),
         starts_with("liver"),
         starts_with("bone"),
         starts_with("lung"),
         starts_with("reproductive_organs"),
         starts_with("intra_abdominal"),
         starts_with("cns"),
         starts_with("pleura")) |> 
  colnames()

treatment_features <- data_train_os |> 
  select(starts_with("biologic"),
         starts_with("chemo"),
         starts_with("hormone"),
         starts_with("investigational"),
         starts_with("prior"),
         starts_with("radiation"),
         starts_with("targeted")) |> 
  colnames()

nlp_features <- c("smoking_status_nlp_former_current_smoker",
                  "smoking_status_nlp_never",
                  "smoking_status_nlp_unknown",
                  "prior_med_nlp_no_prior_medications",
                  "prior_med_nlp_prior_medications_to_msk",
                  "prior_med_nlp_unknown",
                  "mmr_absent_false",
                  "mmr_absent_na",
                  "mmr_absent_true")

clinical_features <- setdiff(colnames(data_train_os),
        c(id_cols, 
          outcome_cols, 
          cna_features, 
          mut_features, 
          sv_features, 
          nlp_features,
          radiology_features, 
          treatment_features)) 
```

# Settings

```{r}
#| label: settings

features_included <- c(clinical_features, 
                       radiology_features,
                       treatment_features,
                       nlp_features,
                       cna_features,
                       mut_features,
                       sv_features)

caption <- "clinical, treatment, radiology, NLP-derived, copy number variations, mutations, structural variants."

file_name <- "05_Forward_Cox_5CV_all_100max"

max_iter <- 100
```

# Prepare data

Select included features.

```{r}
#| label: select_feature_sets

data_train_os <- data_train_os |> 
  select(all_of(c(outcome_cols,
                  features_included))) 

data_train_dfs <- data_train_dfs |> 
  select(all_of(c(outcome_cols,
                  features_included))) 
```

Remove features that are linear combinations of other features.

```{r}
#| label: linear_combo_filter
#| echo: true

# OS
feature_data_os <- data_train_os |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_os)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_os <- data_train_os |> 
  select(-all_of(cols_to_remove))

# DFS
feature_data_dfs <- data_train_dfs |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_dfs)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_dfs <- data_train_dfs |> 
  select(-all_of(cols_to_remove))
```

For the categorical features, we need to remove the features that have a category with zero events or zero controls.

```{r}
#| label: exclude_vars_no_event_levels
#| echo: true

numerical_featurs <- c("tmb_nonsynonymous",
                       "sample_coverage",
                       "num_icdo_dx",
                       "age_surgery",
                       "tumor_purity",
                       "msi_score",
                       "cea")

# OS
explanatory_vars_os <- setdiff(colnames(data_train_os),
                               outcome_cols)

to_test_os <- setdiff(explanatory_vars_os,
                      numerical_featurs)

to_remove_os <- c()

for(var in to_test_os){
  test <- data_train_os |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

data_train_os <- data_train_os |> 
  select(-all_of(to_remove_os)) |>
  select(-c("dfs_status", 
            "years_to_dfs_event"))

# DFS
explanatory_vars_dfs <- setdiff(colnames(data_train_dfs),
                                outcome_cols)

to_test_dfs <- setdiff(explanatory_vars_dfs,
                       numerical_featurs)

to_remove_dfs <- c()

for(var in to_test_dfs){
  test <- data_train_dfs |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

data_train_dfs <- data_train_dfs |> 
  select(-all_of(to_remove_dfs)) |>
  select(-c("os_status", 
            "years_to_os_event"))
```

# Cox propostional hazards models with step-wise forward feature selection

Set up cross-validation folds.

```{r}
#| label: 5_fold_CV
#| echo: true

folds_os <- createFolds(data_train_os$os_status,
                        k = 5,
                        list = TRUE,
                        returnTrain = FALSE)

folds_dfs <- createFolds(data_train_dfs$dfs_status,
                         k = 5,
                         list = TRUE,
                         returnTrain = FALSE)
```

Forward step-wise Cox proportional hazards model function.

```{r}
#| label: Cox_function
#| echo: true

forward_cox_cv <- function(data, 
                           time_var, 
                           status_var, 
                           folds, 
                           max_iter) {

  outcome_cols <- c(time_var, 
                    status_var)
  
  features_all <- setdiff(colnames(data), 
                          outcome_cols)

  results <- list()

  for (k in seq_along(folds)) {

    test_idx  <- folds[[k]]
    train_idx <- setdiff(seq_len(nrow(data)),
                         test_idx)

    data_train <- data[train_idx, ]
    data_test  <- data[test_idx, ]

    features <- features_all
    added_features <- character()

    max_iter_k <- min(length(features), 
                      max_iter)

    for (iter in seq_len(max_iter_k)) {

      c_indices <- tibble(feature = character(),
                          c_index = numeric())

      for (f in features) {

        formula <- as.formula(paste0(
          "Surv(", time_var, ", ", status_var,  ") ~ ",
          paste(c(added_features,
                  f), 
                collapse = " + ")))

        model <- coxph(formula, 
                       data = data_train)

        lp <- predict(model, 
                      newdata = data_test, 
                      type = "lp")

        c_index <- survConcordance(as.formula(paste0(
          "Surv(", time_var, ", ", status_var, ") ~ lp")),
          data = data_test)$concordance

        c_indices <- add_row(c_indices,
                             feature = f,
                             c_index = c_index)
      }

      best <- c_indices$feature[which.max(c_indices$c_index)]

      results[[length(results) + 1]] <- tibble(
        fold = k,
        iteration = iter,
        added_feature = best,
        c_index = max(c_indices$c_index)
      )

      added_features <- c(added_features, 
                          best)
      features <- setdiff(features, 
                          best)
    }
  }

  bind_rows(results)
}
```

## Predicting overall survival

```{r}
#| label: cph_OS
#| echo: true
#| warning: false

cv_os <- forward_cox_cv(
  data      = data_train_os,
  time_var  = "years_to_os_event",
  status_var = "os_status",
  folds     = folds_os,
  max_iter  = max_iter
)

write_tsv(x = cv_os,
         file = paste0("../results/05_",
                       file_name,
                       "_results_OS.tsv"))

summary_os <- cv_os |> 
  group_by(iteration) |> 
  summarise(
    mean_c_index = mean(c_index),
    lower_95 = quantile(c_index, 0.025),
    upper_95 = quantile(c_index, 0.975),
    .groups = "drop"
  )

summary_os |> 
  ggplot(mapping = aes(x = iteration, 
                       y = mean_c_index)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_95, 
                  ymax = upper_95),
              alpha = 0.2) +
  labs(
    x = "Number of features included",
    y = "C-index (95% CI)",
    title = "5-fold CV step-wise Cox model (OS)",
    subtitle = paste("Feature sets included:", caption)
  ) +
  theme_minimal()

ggsave(paste0("../results/05_",
                       file_name,
                       "_concordance_OS.png"))
```

## Predicting disease-free survival

```{r}
#| label: cph_DFS
#| echo: true
#| warning: false

cv_dfs <- forward_cox_cv(
  data       = data_train_dfs,
  time_var   = "years_to_dfs_event",
  status_var = "dfs_status",
  folds      = folds_dfs,
  max_iter   = max_iter
)

write_tsv(x = cv_dfs,
         file = paste0("../results/05_",
                       file_name,
                       "_results_DFS.tsv"))


summary_dfs <- cv_dfs |> 
  group_by(iteration) |> 
  summarise(
    mean_c_index = mean(c_index),
    lower_95 = quantile(c_index, 0.025),
    upper_95 = quantile(c_index, 0.975),
    .groups = "drop"
  )

summary_dfs |> 
  ggplot(mapping = aes(x = iteration, 
                       y = mean_c_index)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_95, 
                  ymax = upper_95),
              alpha = 0.2) +
  labs(
    x = "Number of features included",
    y = "C-index (95% CI)",
    title = "5-fold CV step-wise Cox model (DFS)",
    subtitle = paste("Feature sets included:", caption)
  ) +
  theme_minimal()

ggsave(paste0("../results/05_",
                       file_name,
                       "_concordance_DFS.png"))
```
