---
title: "Random survival forest models"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: packages
#| message: false
#| echo: true

library("tidyverse")
library("caret")
library("randomForestSRC")
library("survival")
```

```{r}
#| label: seed
#| echo: true

set.seed(1)
```

# Load data

```{r}
#| label: load_data
#| message: false
#| echo: true

data_train_os <- read_tsv("../data/03_data_train_os.tsv")

data_train_dfs <- read_tsv("../data/03_data_train_dfs.tsv")
```

Define feature types.

```{r}
#| label: feature_sets

id_cols <- c("patient_id",
             "sample_id")

outcome_cols <- c("os_status",
                  "years_to_os_event",
                  "dfs_status",
                  "years_to_dfs_event")

cna_features <- data_train_os |> 
  select(starts_with("cna")) |> 
  colnames()

mut_features <- data_train_os |> 
  select(starts_with("mut")) |> 
  colnames()

sv_features <- data_train_os |> 
  select(starts_with("sv")) |> 
  colnames()

radiology_features <- data_train_os |> 
  select(starts_with("chest"),
         starts_with("abdomen"),
         starts_with("pelvis"),
         starts_with("head"),
         starts_with("other"),
         starts_with("lymph_nodes"),
         starts_with("liver"),
         starts_with("bone"),
         starts_with("lung"),
         starts_with("reproductive_organs"),
         starts_with("intra_abdominal"),
         starts_with("cns"),
         starts_with("pleura")) |> 
  colnames()

treatment_features <- data_train_os |> 
  select(starts_with("biologic"),
         starts_with("chemo"),
         starts_with("hormone"),
         starts_with("investigational"),
         starts_with("prior"),
         starts_with("radiation"),
         starts_with("targeted")) |> 
  colnames()

nlp_features <- c("smoking_status_nlp_former_current_smoker",
                  "smoking_status_nlp_never",
                  "smoking_status_nlp_unknown",
                  "prior_med_nlp_no_prior_medications",
                  "prior_med_nlp_prior_medications_to_msk",
                  "prior_med_nlp_unknown",
                  "mmr_absent_false",
                  "mmr_absent_na",
                  "mmr_absent_true")

clinical_features <- setdiff(colnames(data_train_os),
        c(id_cols, 
          outcome_cols, 
          cna_features, 
          mut_features, 
          sv_features, 
          nlp_features,
          radiology_features, 
          treatment_features)) 
```

# Settings

```{r}
#| label: settings

features_included <- c(clinical_features, 
                       #radiology_features,
                       treatment_features)#,
                       #nlp_features,
                       #cna_features,
                       #mut_features,
                       #sv_features)

caption <- "noNLPgenomic"

file_name <- "06_RSF_5CV_noNLPgenomic_1000trees_bs"

trees <- 1000

rule <- "bs.gradient" # logrank (default), bs.gradient, or logrankscore

tree_grid <- c(1, seq(50, trees, by = 50))
```

# Prepare data

Select included features.

```{r}
#| label: select_feature_sets

data_train_os <- data_train_os |> 
  select(all_of(c(outcome_cols,
                  features_included))) 

data_train_dfs <- data_train_dfs |> 
  select(all_of(c(outcome_cols,
                  features_included))) 
```

Remove features that are linear combinations of other features.

```{r}
#| label: linear_combo_filter
#| echo: true

# OS
feature_data_os <- data_train_os |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_os)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_os <- data_train_os |> 
  select(-all_of(cols_to_remove))

# DFS
feature_data_dfs <- data_train_dfs |> 
    select(-all_of(outcome_cols))

mm <- model.matrix(~ . -1, 
                   data = feature_data_dfs)

lc <- findLinearCombos(mm)

cols_to_remove <- colnames(mm)[lc$remove]

data_train_dfs <- data_train_dfs |> 
  select(-all_of(cols_to_remove)) 
```

For the categorical features, we need to remove the features that have a category with zero events or zero controls.

```{r}
#| label: exclude_vars_no_event_levels
#| echo: true

numerical_featurs <- c("tmb_nonsynonymous",
                       "sample_coverage",
                       "num_icdo_dx",
                       "age_surgery",
                       "tumor_purity",
                       "msi_score",
                       "cea")

# OS
explanatory_vars_os <- setdiff(colnames(data_train_os),
                               outcome_cols)

to_test_os <- setdiff(explanatory_vars_os,
                      numerical_featurs)

to_remove_os <- c()

for(var in to_test_os){
  test <- data_train_os |> 
    group_by(.data[[var]]) |> 
    summarise(has_os1 = any(os_status == 1),
              has_os0 = any(os_status == 0)) |> 
    filter(!has_os1 | !has_os0)
  
  if(nrow(test) > 0){
    to_remove_os <- c(to_remove_os, var)
  }
}

data_train_os <- data_train_os |> 
  select(-all_of(to_remove_os)) |>
  select(-c("dfs_status", 
            "years_to_dfs_event"))

# DFS
explanatory_vars_dfs <- setdiff(colnames(data_train_dfs),
                                outcome_cols)

to_test_dfs <- setdiff(explanatory_vars_dfs,
                       numerical_featurs)

to_remove_dfs <- c()

for(var in to_test_dfs){
  test <- data_train_dfs |> 
    group_by(.data[[var]]) |> 
    summarise(has_dfs1 = any(dfs_status == 1),
              has_dfs0 = any(dfs_status == 0)) |> 
    filter(!has_dfs1 | !has_dfs0)
  
  if(nrow(test) > 0){
    to_remove_dfs <- c(to_remove_dfs, var)
  }
}

data_train_dfs <- data_train_dfs |> 
  select(-all_of(to_remove_dfs)) |>
  select(-c("os_status", 
            "years_to_os_event"))
```

# RSF with cross-validation

Set up cross-validation folds.

```{r}
#| label: 5_fold_CV
#| echo: true

folds_os <- createFolds(data_train_os$os_status,
                        k = 5,
                        list = TRUE,
                        returnTrain = FALSE)

folds_dfs <- createFolds(data_train_dfs$dfs_status,
                         k = 5,
                         list = TRUE,
                         returnTrain = FALSE)
```

## Predicting overall survival

```{r}
#| label: rsf_os_cv
#| echo: true

rsf_res_os <- map_dfr(tree_grid,
                      function(trees) {

    fold_c <- map_dbl(folds_os,
                      ~ {
        test_idx  <- .x
        train_idx <- setdiff(seq_len(nrow(data_train_os)), 
                             test_idx)

        train_data <- data_train_os[train_idx, ]
        test_data  <- data_train_os[test_idx, ]

        fit <- rfsrc(Surv(years_to_os_event, 
                          os_status) ~ .,
                     data = train_data,
                     ntree = trees,
                     splitrule = rule)

        pred <- predict(fit, 
                        newdata = test_data)

        concordance(Surv(test_data$years_to_os_event, 
                         test_data$os_status) ~ pred$predicted)$concordance
      }
    )

    tibble(trees = trees,
           c_fold_1 = fold_c[1],
           c_fold_2 = fold_c[2],
           c_fold_3 = fold_c[3],
           c_fold_4 = fold_c[4],
           c_fold_5 = fold_c[5])
  }
)

write_tsv(x = rsf_res_os,
          file = paste0("../results/",
                        file_name,
                        "_results_OS.tsv"))

summary_os <- rsf_res_os |> 
  pivot_longer(cols = -trees,
               names_to = "fold",
               values_to = "c_index") |> 
  group_by(trees) |> 
  summarise(mean_c_index = mean(c_index),
            lower_95 = quantile(c_index, 0.025),
            upper_95 = quantile(c_index, 0.975),
            .groups = "drop")

summary_os |> 
  ggplot(mapping = aes(x = trees, 
                       y = mean_c_index)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_95, 
                  ymax = upper_95),
              alpha = 0.2) +
  labs(
    x = "Number of trees",
    y = "C-index (95% CI)",
    title = "5-fold CV RSF model (OS)",
    subtitle = paste("Feature sets included:", caption)
  ) +
  theme_minimal()

ggsave(paste0("../results/",
              file_name,
              "_concordance_OS.png"))

final_os <- summary_os |> 
  filter(trees == max(trees)) |> 
  mutate(model = "RSF",
         features = caption,
         outcome = "OS",
         rule = rule) |> 
  select(c(model, features, outcome, 
           c_index = mean_c_index, lower_95, upper_95, rule))

write_tsv(final_os,
          file = "../results/06_RSF_5CV_model_results.tsv",
          append = TRUE)
```

## Predicting disease-free survival

```{r}
#| label: rsf_dfs_cv
#| echo: true

rsf_res_dfs <- map_dfr(tree_grid,
                       function(trees) {

    fold_c <- map_dbl(folds_dfs,
                      ~ {
        test_idx  <- .x
        train_idx <- setdiff(seq_len(nrow(data_train_dfs)), 
                             test_idx)

        train_data <- data_train_dfs[train_idx, ]
        test_data  <- data_train_dfs[test_idx, ]

        fit <- rfsrc(Surv(years_to_dfs_event, 
                          dfs_status) ~ .,
                     data = train_data,
                     ntree = trees,
                     splitrule = rule)

        pred <- predict(fit, 
                        newdata = test_data)

        concordance(Surv(test_data$years_to_dfs_event, 
                         test_data$dfs_status) ~ pred$predicted)$concordance
      }
    )

    tibble(trees = trees,
           c_fold_1 = fold_c[1],
           c_fold_2 = fold_c[2],
           c_fold_3 = fold_c[3],
           c_fold_4 = fold_c[4],
           c_fold_5 = fold_c[5])
  }
)

write_tsv(x = rsf_res_dfs,
          file = paste0("../results/",
                        file_name,
                        "_results_DFS.tsv"))

summary_dfs <- rsf_res_dfs |> 
  pivot_longer(cols = -trees,
               names_to = "fold",
               values_to = "c_index") |> 
  group_by(trees) |> 
  summarise(mean_c_index = mean(c_index),
            lower_95 = quantile(c_index, 0.025),
            upper_95 = quantile(c_index, 0.975),
            .groups = "drop")

summary_dfs |> 
  ggplot(mapping = aes(x = trees, 
                       y = mean_c_index)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_95, 
                  ymax = upper_95),
              alpha = 0.2) +
  labs(
    x = "Number of trees",
    y = "C-index (95% CI)",
    title = "5-fold CV RSF model (DFS)",
    subtitle = paste("Feature sets included:", caption)
  ) +
  theme_minimal()

ggsave(paste0("../results/",
              file_name,
              "_concordance_DFS.png"))

final_dfs <- summary_dfs |> 
  filter(trees == max(trees)) |> 
  mutate(model = "RSF",
         features = caption,
         outcome = "DFS",
         rule = rule) |> 
  select(c(model, features, outcome, 
           c_index = mean_c_index, lower_95, upper_95, rule))

write_tsv(final_dfs,
          file = "../results/06_RSF_5CV_model_results.tsv",
          append = TRUE)
```
